### 2018-12-22 notes/derivation.tex
started on derivation notes

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Vadim}[1]{{\color{blue}Vadim: #1}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}
\pdfinfo{%
  /Title    ()
  /Author   ()
  /Creator  ()
  /Producer ()
  /Subject  ()
  /Keywords ()
}

\begin{document}

\title{Site-specific substitution model inference with iterative tree likelihood maximization.}
\author{Vadim Puller$^{1,2,3}$ Pavel Sagulenko$^{1}$, Richard Neher$^{1,2,3}$}
\affiliation{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany\\
$^{2}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
$^{3}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland}

\date{\today}

\maketitle

Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
\bq
P(\vec{s}\leftarrow \vec{r},t,\mat{Q}) = \prod_{a=1}^L \left(e^{\mat{Q}^{a} t}\right)_{s_{a},r_{a}}
\label{eq:P_prob}
\eq
where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $i$, and $s_{a}$ and $r_{a}$ are the sequence states at position $i$.
The product runs over all $L$ sites $a$ and amounts to assuming that different sites of the sequence evolve independently.

In absence of recombination, homologous sequences are related by a tree and the likelihood of observing an alignment $\mat{A}=\{\vec{s^k}, k=1..n\}$ conditional on the tree $T$ and the substitution model $\mat{Q}^{a}$ can be written in terms of propagators defined in Eq.~\ref{eq:P_prob}.
It is helpful to express this likelihood as product of sequence propagators defined in Eq.~\ref{eq:P_prob} between sequences at the ends of each branch in the tree (implicitly assuming that on different branches is independent and follows the same time reversible model).
Unknown sequences of internal nodes $\{\vec{x}\}$ need to be summed over and the likelihood can be expressed as
\begin{equation}
	\LH(\mat{A}|T,\mat{Q}) = \sum_{\{\vec{x}\}}\prod_{a=1}^L \eqp^{a}_{x^{0}_a} \prod_{k\in T}P(\vec{s}_{k_c}\leftarrow \vec{s}_{k_p},t,\mat{Q}) \ ,
	\label{eq:LH}
\end{equation}
where $\vec{s}_{k_c}$ and $\vec{s}_{k_p}$ are the child and parent sequences of branch $k$, respectively, and the factor $\prod \eqp^{a}_{x^{0}_a}$ is the product of the probabilities of the root sequence $x^{0}_a$ over all positions $i$.
The probabilities $\vec{\eqp}^{a}$ are the equiblirium probabilities of the substitution model at position $a$.
The latter ensures that the likelihood is insensitive to a particular choice of the tree root.

The likelihood can be computed efficiently using standard dynamic programming techniques.
Nevertheless, it requires $\mathcal{O}(n\times L \times q^2)$ operations (where $q$ is the size of the alphabet $\abet$) and optimizing it with respect to a large number of parameters is costly.
Our goal here is to infer site-specific substitution models using a computationally efficient iterative procedure.

We will parameterize the substitution model as follows:
We describe the substitution process using site-specific GTR
(general time-reversible) model, parametrized as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
\label{eq:Qij}
\end{eqnarray}
where $W_{ij}$ is a symmetric matrix and the second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.

For each branch $k$ in the tree we can compute the likelihoods of the parent and child subtrees obtained by this branch conditional on the sequence at either end of the branch.
This, in turn, allows us to compute the marginal distribution $\rho^{a}_i$ and $\sigma^{a}_i$ of character $i$ at position $a$ of root sequence of the parent and child subtrees, respectively.

Using the product from of the likelihood, we can now write down compact expressions for the derivative of the likelihood and the with respect to the substitution model.
\begin{equation}
	\frac{\partial \log \LH}{\partial \mu^{b}} = \sum_{k\in T} \frac{t_k}{\mu^{b}}\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{equation}
Analogously, we find for the derivaties wrt to $\eqp^{k,b}_{i}$
\begin{equation}
\frac{\partial \log \LH}{\partial \eqp^{b}_i}= \sum_{k\in T} \frac{t_k}{\eqp^{b}_i}\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_i}{\eqp^b_i}
\end{equation}
where $\sigma^{0,b}_i$ corresponds to state of the root sequence at position $b$.
The corresponding derivative for the transition matrix $W_{rs}$ is
\begin{equation}
	\frac{\partial \log \LH}{\partial W_{ij}}= \sum_{k\in T} \frac{t_k}{W_{ij}} \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
\end{equation}
The at an extremum, all these derivatives have to vanish.
In addition we need to impose the normalization constraints which is done in the form of Lagrange multipliers such that the final conditions are
\begin{eqnarray}
	0 &=& \sum_{k\in T} t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \nonumber \\
    0 &=& 	\sum_{k\in T} t_k\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}} + \sigma^{0,b}_i + \lambda^{b}\eqp^b_i + \kappa \sum_j W_{ij}\eqp_i^{b}\eqp_j^{b} \nonumber\\
	0 &=& \sum_{k\in T} t_k \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \kappa W_{ij}\sum_{b=1}^L \eqp^b_i\eqp^b_j
\end{eqnarray}
Summing the second and third equation over $i$ and $i,j$, respectively, and substituting the first equation, we find $\kappa=0$ and $\lambda^a=1$.

To arrive at an iterative scheme to solve these equations, it is helpful to split the individuals terms into \emph{diagonal} contributions (the case in which the sequence does not change along a branch) and \emph{off-diagonal} contributions (substitutions happened on that branch).
We will first consider the limit of short branches were we can approximate $e^{\mat{Q}^at}=\mat{I} + \mat{Q}^at$ and evaluate to leading order
\begin{equation}
\begin{split}
t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \approx &	t_k \frac{-\sum_{i} \mu^b\sum_h p_h^b W_{hi} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i, j\neq i} \mu^b\sigma^{k,b}_i p_i^bW_{ij} \rho^{k,b}_j}{\sum_{i} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i,j\neq i} \mu^b \sigma^{k,b}_i \eqp_i^b W_{ij}t_k \rho^{k,b}_j}\\
\end{split}
\end{equation}
In the limit of short branches (or if ancestral states are known), $\sigma^{k,b}_i$ and $\rho^{k,b}_i$ are either $0$ or $1$ such that only one of the two terms in the numerator and denominator contribute.
In case $\sigma^{k,b}_i=\rho^{k,b}_i$ (no substitution on that branch), the denominator is 1 and the numerator $- \mu^b \sum_{i,h}p_h^b W_{hi}t_k$.
If a substitution happened, numerator and denominator cancel with result 1.
After summing over all branches, the condition for the mutation rate simplifies to
\begin{equation}
	\mu^b \sum_i \eqp_i^b W_{ij}T_j^b = \sum_{i\neq j} n_{ij}
\end{equation}
where $T_j^b$ is the sum of all $t_k$ were the sequence is in state $j$.

\bibliography{GTRbib}

\end{document}

### 2018-12-28 src/plot_toy_data.py
implement iterative scheme

@@ -7,55 +7,88 @@ if __name__ == '__main__':
     mu_dist = defaultdict(lambda: defaultdict(list))
     p_dist = defaultdict(lambda: defaultdict(list))
     W_dist = defaultdict(lambda: defaultdict(list))
+    delta_LH = defaultdict(lambda: defaultdict(list))
+    avg_rate = defaultdict(lambda: defaultdict(list))
 
     mu_vals = set()
     n_vals = set()
 
     for fname in glob.glob('2018-12-17_simulated_data_results/*.pkl'):
         with open(fname, 'rb') as fh:
-            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_mu_dist, tmp_W_dist = pickle.load(fh)
+            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu = pickle.load(fh)
 
         n_vals.update(tmp_n_vals)
         mu_vals.update(tmp_mu_vals)
 
-        for d,tmp_d in [(mu_dist, tmp_mu_dist), (W_dist, tmp_W_dist), (p_dist, tmp_p_dist)]:
+        for d,tmp_d in [(mu_dist, tmp_mu_dist), (W_dist, tmp_W_dist), (p_dist, tmp_p_dist),
+                        (delta_LH, tmp_LH), (avg_rate, tmp_avg_mu)]:
             for x in tmp_d:
                 for y in tmp_d[x]:
                     d[x][y].extend(tmp_d[x][y])
 
 
     L=300
-    n_vals = sorted(n_vals)
-    mu_vals = sorted(mu_vals)
+    n_vals = np.array(sorted(n_vals))
+    mu_vals = np.array(sorted(mu_vals))
 
     from matplotlib import pyplot as plt
+
+    # for each data set size, plot the distance of the inferred equilibrium frequencies
+    # and the substitution rates from the true frequencies and rates
     for n in n_vals:
+        rtt = np.log(n)
         plt.figure()
         for label, data in p_dist.items():
-            plt.errorbar(mu_vals, [np.mean(data[(L,n,mu)]) for mu in mu_vals],
+            plt.errorbar(mu_vals*rtt, [np.mean(data[(L,n,mu)]) for mu in mu_vals],
                         [np.std(data[(L,n,mu)]) for mu in mu_vals], label=label)
 
         plt.yscale('log')
         plt.legend()
[…]

### 2018-12-28 src/reconstruct_toy_data.py

@@ -1,6 +1,5 @@
 import os, gzip, glob, pickle
 import numpy as np
-from Bio import Phylo, AlignIO
 from collections import defaultdict
 
 from treetime.treeanc import TreeAnc
@@ -10,49 +9,7 @@ from treetime.seq_utils import seq2prof, profile_maps, alphabets
 
 from generate_toy_data import *
 from filenames import *
-
-def p_from_aln(in_prefix, params, alphabet='nuc_nogap'):
-    with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
-        aln = AlignIO.read(fh, 'fasta')
-
-    alpha = alphabets[alphabet]
-    aln_array = np.array(aln)
-    af = []
-    for a in alpha:
-        af.append(np.mean(aln_array==a, axis=0))
-    return np.array(af)
-
-
-def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap', marginal=False, reconstructed_tree=False):
-    with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
-        aln = AlignIO.read(fh, 'fasta')
-    tree_fname =  reconstructed_tree_name(in_prefix, params) if reconstructed_tree else tree_name(in_prefix, params)
-    myTree = TreeAnc(gtr=gtr, alphabet=alphabet,
-                     tree=tree_fname, aln=aln,
-                     reduce_alignment=False, verbose = 0)
-
-    if type(gtr)==str:
-        myTree.infer_ancestral_sequences(marginal=True, infer_gtr=True, normalized_rate=False)
-    else:
-        myTree.infer_ancestral_sequences(marginal=True)
-
-    mutation_counts = get_mutation_count(myTree, alphabet=myTree.gtr.alphabet, marginal=marginal)
-    return mutation_counts, myTree.sequence_LH(), myTree
-
-
-def estimate_GTR(mutation_counts, pc=0.1, single_site=False):
-    n_ija, T_ia, root_sequence = mutation_counts
-    root_prof = seq2prof(root_sequence, profile_maps['nuc_nogap']).T
-
-    if single_site:
-        inferred_model = GTR.infer(n_ija.sum(axis=-1), T_ia.sum(axis=-1),
-                          root_state=root_prof.sum(axis=-1), alphabet='nuc_nogap')
-    else:
-        inferred_model = GTR_site_specific.infer(n_ija, T_ia, pc=pc,
[…]

### 2018-12-28 notes/derivation.tex
updating notes

@@
 \end{equation}
-Analogously, we find for the derivaties wrt to $\eqp^{k,b}_{i}$
+The derivative wrt to $\eqp^{b}_{s}$ requires a little more care (and is infact not correct as the commutator needs to be accounted for...):
 \begin{equation}
-\frac{\partial \log \LH}{\partial \eqp^{b}_i}= \sum_{k\in T} \frac{t_k}{\eqp^{b}_i}\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_i}{\eqp^b_i}
+	\frac{\partial Q^b_{ij}}{\partial \eqp^b_s} = -\delta_{ij}\sum_{k\neq i} \frac{\partial \eqp^b_k}{\partial \eqp^b_s} W_{ik} + (1-\delta_{ij})\delta_{is}W_{sj} = -\delta_{ij}(1-\delta_{is}) W_{is} + (1-\delta_{ij})\delta_{is}W_{sj}
 \end{equation}
-where $\sigma^{0,b}_i$ corresponds to state of the root sequence at position $b$.
-The corresponding derivative for the transition matrix $W_{rs}$ is
+Using this, we can now write
 \begin{equation}
-	\frac{\partial \log \LH}{\partial W_{ij}}= \sum_{k\in T} \frac{t_k}{W_{ij}} \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
+\begin{split}
+ \frac{\partial \log \LH}{\partial \eqp^{b}_s} &= \sum_{k\in T} t_k\mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i (-\delta_{ij}(1-\delta_{is}) W_{is} + (1-\delta_{ij})\delta_{is}W_{sj})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_s}{\eqp^b_s} \\
+& =\sum_{k\in T} t_k\mu_b \frac{-\sum_{h,i\neq s}\sigma^{k,b}_i W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,j\neq s}\sigma^{k,b}_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \frac{\sigma^{0,b}_s}{\eqp^b_s}
+\end{split}
 \end{equation}
-The at an extremum, all these derivatives have to vanish.
-In addition we need to impose the normalization constraints which is done in the form of Lagrange multipliers such that the final conditions are
-\begin{eqnarray}
-	0 &=& \sum_{k\in T} t_k\frac{\vec{\sigma}^{k,b} \mat{Q}^{b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \nonumber \\
-    0 &=& 	\sum_{k\in T} t_k\frac{\sum_{h,j}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t} \vec{\rho}^{k,b}} + \sigma^{0,b}_i + \lambda^{b}\eqp^b_i + \kappa \sum_j W_{ij}\eqp_i^{b}\eqp_j^{b} \nonumber\\
-	0 &=& \sum_{k\in T} t_k \sum_{a=1}^L\frac{\sum_{h}\sigma^{k,b}_i Q^{k,b}_{ih} e^{\mat{Q}^{b}t_k}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \kappa W_{ij}\sum_{b=1}^L \eqp^b_i\eqp^b_j
-\end{eqnarray}
-Summing the second and third equation over $i$ and $i,j$, respectively, and substituting the first equation, we find $\kappa=0$ and $\lambda^a=1$.
+where $\sigma^{0,b}_s$ corresponds to the distribution of states of the root sequence at position $b$.
+The corresponding derivative wrt to the transition matrix $W_{rs}$ has to account for the fact that $W_{ij}=W_{ji}$ and hence
+\begin{equation}
+\begin{split}
+\frac{\partial Q^b_{ij}}{\partial W_{rs}} & = \mu^b(1-\delta_{rs})(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \mu_b\delta_{ij}(\sum_{h\neq i} p^b_h (\delta_{ir}\delta_{hs} + \delta_{is}\delta_{hr}) ) \\
+& = \mu^b(1-\delta_{rs})\left[(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})\right]
+\end{split}
+\end{equation}
+We will not consider the case $r=s$ and drop the $1-\delta_{rs}$ condition.
+\begin{equation}
+\begin{split}
+	\frac{\partial \log \LH}{\partial W_{rs}} & = \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i \left[(1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) - \delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})\right] e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} \\
+	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{i,j,h}\sigma^{k,b}_i (1-\delta_{ij})(p^b_r\delta_{ir}\delta_{js} + p^b_s\delta_{is}\delta_{jr}) e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h
+	- \sum_{i,j,h}\sigma^{k,b}_i\delta_{ij}(p^b_s \delta_{ir} + p^b_r\delta_{is})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}\\
+	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{h}\left[\sigma^{k,b}_r p^b_r e^{\mat{Q}^{b}t_k}_{sh} + \sigma^{k,b}_s p^b_s e^{\mat{Q}^{b}t_k}_{rh}\right] \rho^{k,b}_h
+	- \sum_{j,h}\sigma^{k,b}_j(p^b_s \delta_{jr} + p^b_r\delta_{js})e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}\\
+	&= \sum_{k\in T} t_k \sum_{b=1}^L \mu^b \frac{\sum_{h}\left[\sigma^{k,b}_r p^b_r e^{\mat{Q}^{b}t_k}_{sh} + \sigma^{k,b}_s p^b_s e^{\mat{Q}^{b}t_k}_{rh}\right] \rho^{k,b}_h
+	- \sum_{h} \left[\sigma^{k,b}_r p^b_s e^{\mat{Q}^{b}t_k}_{rh} + \sigma^{k,b}_s p^b_r e^{\mat{Q}^{b}t_k}_{sh}\right] \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}}
+\end{split}
+\end{equation}
+At an extremum, all these derivatives have to vanish -- after accounting for additional constraints due to normalization.
+In particular, we would like to enforce $\sum_i \eqp^b_i=0$ which results in the extramal condition
+\begin{equation}
+	0 = \eqp^b_s \sum_{k\in T} t_k \frac{-\sum_{h,i\neq s}\sigma^{k,b}_i W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,j\neq s}\sigma^{k,b}_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \sigma^{0,b}_s + \lambda^b \eqp^b_s
+\end{equation}
+To determine $\lambda^b$, we sum this over $s$.
+\begin{equation}
+\begin{split}
+	0 & = \sum_{k\in T} t_k \frac{-\sum_{h,s,i\neq s}\sigma^{k,b}_i \eqp^b_s W_{is}e^{\mat{Q}^{b}t_k}_{ih} \rho^{k,b}_h + \sum_{h,s, j\neq s}\sigma^{k,b}_s \eqp^b_s W_{sj} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
+	0 & = \sum_{k\in T} t_k \frac{\sum_{h,s,j\neq s} \left[-\sigma^{k,b}_j \eqp^b_s W_{js} + \sigma^{k,b}_s \eqp^b_s W_{sj}\right] e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
+	0 & = \sum_{k\in T} t_k \frac{\sum_{h,s,j\neq s} \left[-\sigma^{k,b}_j + \sigma^{k,b}_s\right]  \eqp^b_s W_{sj}e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b \\
+	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j,s\neq j} \left[-\sigma^{k,b}_j\eqp^b_s W_{sj} + \sigma^{k,b}_s\eqp^b_s W_{sj}\right]  e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b\\
+	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j} \left[\sigma^{k,b}_jQ_{jj} + \sum_{s\neq j}\sigma^{k,b}_sQ_{sj}\right]  e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b\\
+	0 & = \sum_{k\in T} t_k \frac{\sum_{h,j,i} \sigma^{k,b}_i Q_{ij} e^{\mat{Q}^{b}t_k}_{jh} \rho^{k,b}_h}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + 1 + \lambda^b
+\end{split}
+\end{equation}
+The first term vanishes accoring to the extremal condition for $\mu^b$ and hence $\lambda_b=-1$.
+The other Lagrange parameter for $W_{ij}$ vanishes (I think) because there is no source associated with it.
+
+There is a general pattern here: All conditions consist of a positive and negative term that need to balance.
+The former is dominated by `off-diagonal' entries, i.e., when the overlap between $\sigma_i$ and $\rho_i$ is small, the latter by diagonal entries.
+The ratio of the two can be cast into a multiplicative update rule for the rates.
 
@@
 \begin{equation}
-	\mu^b \sum_i \eqp_i^b W_{ij}T_j^b = \sum_{i\neq j} n_{ij}
+	\mu^b \sum_i \eqp_i^b W_{ij}T_j^b = \sum_{i\neq j} n^b_{ij}
 \end{equation}
-where $T_j^b$ is the sum of all $t_k$ were the sequence is in state $j$.
+where $T_j^b$ is the sum of all $t_k$ were the sequence is in state $j$ and $n^b_ij$ are the number of times a mutation from state $j$ to $i$ at position $b$ is observed.
+
+Going through the same manipulations for the conditions for optimal $\eqp^b_i$, we find
+\begin{equation}
+	\begin{split}
+& t_k\frac{\sum_{h,j}\sigma^{k,b}_i Q^{b}_{ih} e^{\mat{Q}^{b}t}_{hj} \rho^{k,b}_j}{\vec{\sigma}^{k,b} e^{\mat{Q}^{b}t_k} \vec{\rho}^{k,b}} + \sigma_i^{0,b}+\eqp_i^b \\
+&\approx t_k \frac{-\sum_{h}\sigma^{k,b}_i\mu^b p_h^b W_{ih} \rho^{k,b}_i + \sum_{j\neq i}\sigma^{k,b}_i \mu^b \eqp_i^b W_{ij} \rho^{k,b}_j}{\sum_{i} \sigma^{k,b}_i \rho^{k,b}_i + \sum_{i,j\neq i} \mu^b \sigma^{k,b}_i \eqp_i^b W_{ij}t_k \rho^{k,b}_j} + \sigma_i^{0,b}+\eqp_i^b
+	\end{split}
+\end{equation}
+This can be solved for $p_i^{b}$ as follows
+% \begin{equation}
+% 	p_i^b(\sum_{h,j\neq i}\eqp_h^b W_{ih}T_i + 1) = \sum_{j\neq i}\sigma^{k,b}_i p_i^h W_{ij} \rho^{k,b}_j}{}
+% \end{equation}
 

### 2018-12-28 src/estimation.py
added iterative iteration with exact mu

@@ -88,31 +88,28 @@ def get_marginal_mutation_count(tree, alphabet):
     L = tree.seq_len
     q = len(alphabet)
     n_ija = np.zeros((q,q,L), dtype=float)
-    n_ija_eff = np.zeros((q,q,L), dtype=float)
     T_ia = np.zeros((q,L),dtype=float)
     for n in tree.tree.get_nonterminals():
         for c in n:
             mut_stack = np.transpose(tree.get_branch_mutation_matrix(c, full_sequence=True), (2,1,0))
-            neff = np.transpose(tree.get_effective_mutation_matrix(c, full_sequence=True), (2,1,0))
             T_ia += 0.5*c.branch_length * mut_stack.sum(axis=0)
             T_ia += 0.5*c.branch_length * mut_stack.sum(axis=1)
 
-            n_ija_eff += neff
             n_ija += mut_stack
 
-    return n_ija, T_ia, tree.tree.root.sequence, n_ija_eff
+    return n_ija, T_ia, tree.tree.root.sequence
 
 
-def get_average_transition_matrix(tree, model):
+def exact_mu_update(tree, alphabet):
+    alphabet_to_index = {a:ai for ai,a in enumerate(alphabet)}
+    L = tree.seq_len
+    q = len(alphabet)
+    mu_diag = np.zeros((L), dtype=float)
+    mu_offdiag = np.zeros((L), dtype=float)
+    for n in tree.tree.get_nonterminals():
+        for c in n:
+            (m1,m2), (p1,p2), (W1,W2) = tree.get_update_matrices(c)
+            mu_diag += m1
+            mu_offdiag += m2
 
-    eQt = np.zeros_like(model.v)
-    tQeQt = np.zeros_like(model.v)
-    for n in tree.find_clades():
-        if n==tree.root:
-            continue
-        else:
-            tmp = model.expQt(n.branch_length)
-            eQt+=tmp
-            tQeQt += n.branch_length*model.Q*tmp
-
-    return eQt, tQeQt
+    return (mu_diag, mu_offdiag)

### 2018-12-28 src/reconstruct_toy_data.py

@@ -25,7 +25,7 @@ if __name__ == '__main__':
     parser.add_argument("-L", type=int, help="length of sequence")
     args=parser.parse_args()
 
-    prefix = '2018-12-17_simulated_data'
+    prefix = '2018-12-26_simulated_data'
     mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
     files = glob.glob(prefix+mask)
 
@@ -38,12 +38,12 @@ if __name__ == '__main__':
     mu_vals = set()
     n_vals = set()
 
-    pc=0.1
-    niter=2
+    pc=0.01
+    niter=5
 
     analysis_types = ['naive', 'single', 'dressed', 'regular', 'phylo', 'marginal', 'true_model', 'iterative']
 
-    for fname in files:
+    for fname in files[:1]:
         print(fname)
 
         params = parse_alignment_name(fname)
@@ -73,7 +73,10 @@ if __name__ == '__main__':
                                                 alphabet='nuc_nogap', marginal=True,
                                                 reconstructed_tree=False)
                         bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
-                        model = estimate_GTR(mc[0], pc=pc, single_site=False, bl=bl)
+                        model = estimate_GTR(mc[0], pc=pc, single_site=False, bl=None)
+                        if i:
+                            mu_num, mu_denom = exact_mu_update(mc[-1], 'nuc_nogap')
+                            model.mu *= mu_num/mu_denom
                 else:
                     rec_model = {'true_model':true_model}
                     mc = reconstruct_counts(prefix, params, gtr=rec_model.get(ana, 'JC69'),

### 2019-01-01 manuscript/ms.tex
added manuscript draft

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Vadim}[1]{{\color{blue}Vadim: #1}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}
\newcommand{\lh}{\ell}
\pdfinfo{%
  /Title    ()
  /Author   ()
  /Creator  ()
  /Producer ()
  /Subject  ()
  /Keywords ()
}

\begin{document}

\title{Site-specific substitution model inference with iterative tree likelihood maximization.}
\author{Vadim Puller$^{1,2,3}$ Pavel Sagulenko$^{1}$, Richard Neher$^{1,2,3}$}
\affiliation{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany\\
$^{2}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
$^{3}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland}

\date{\today}

\maketitle

Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
\bq
P(\vec{s}\leftarrow \vec{r},t,\mat{Q}) = \prod_{a=1}^L \left(e^{\mat{Q}^{a} t}\right)_{s_{a},r_{a}}
\label{eq:P_prob}
\eq
where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $i$, and $s_{a}$ and $r_{a}$ are the sequence states at position $i$.
The product runs over all $L$ sites $a$ and amounts to assuming that different sites of the sequence evolve independently.

In absence of recombination, homologous sequences are related by a tree and the likelihood of observing an alignment $\mat{A}=\{\vec{s^k}, k=1..n\}$ conditional on the tree $T$ and the substitution model $\mat{Q}^{a}$ can be written in terms of propagators defined in Eq.~\ref{eq:P_prob}.
It is helpful to express this likelihood as product of sequence propagators defined in Eq.~\ref{eq:P_prob} between sequences at the ends of each branch in the tree (implicitly assuming that on different branches is independent and follows the same time reversible model).
Unknown sequences of internal nodes $\{\vec{s}'\}$ need to be summed over and the likelihood can be expressed as
\begin{equation}
	\LH(\mat{A}|T,\mat{Q}) = \sum_{\{\vec{s}'\}}\prod_{a=1}^L \eqp^{a}_{x^{0}_a} \prod_{k\in T}P(\vec{s}_{k_c}\leftarrow \vec{s}_{k_p},t,\mat{Q}) = \sum_{\{\vec{s}\}} e^{\lh(\{\vec{s}\} |T, \mat{Q})}  \ ,
	\label{eq:LH}
\end{equation}
where $\vec{s}_{k_c}$ and $\vec{s}_{k_p}$ are the child and parent sequences of branch $k$, respectively, and the factor $\prod \eqp^{a}_{x^{0}_a}$ is the product of the probabilities of the root sequence $s^{0}_a$ over all positions $i$.
The probabilities $\vec{\eqp}^{a}$ are the equiblirium probabilities of the substitution model at position $a$.
The latter ensures that the likelihood is insensitive to a particular choice of the tree root.
This equation defines the log-likelihood $\lh$ of a particular internal node assignment $\{\vec{s}'\}$ which is given by
\begin{equation}
	e^{\lh(\mat{A}, \{\vec{s}'\} |T, \mat{Q})} = \sum_a \left[\log(\eqp^a_{s^{0}_a}) + \sum_{k\in T} \log\left( e^{Q^at_k} \right)_{k^a_c, k^a_p}\right]
\end{equation}
where $k^a_c$ and $k^a_p$ are indices corresponding to the child and parent sequence of branch $k$.

The sum over unknown ancestral sequences can be computed efficiently using standard dynamic programming techniques.
Nevertheless, it requires $\mathcal{O}(n\times L \times q^2)$ operations (where $q$ is the size of the alphabet $\abet$) and optimizing it with respect to a large number of parameters is costly.
Our goal here is to infer site-specific substitution models using a computationally efficient iterative procedure.

Instead of inferring completely independent models for every site in the genome, we only allow for site specific rates and equilibrium frequencies while using the same transition matrix for every site.
Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
\label{eq:Qij}
\end{eqnarray}
where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.

The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
\begin{equation}
	\frac{\partial \LH(\mat{A}|T,\mat{Q})}{\partial X} = \sum_{\{\vec{s}\}} e^{\lh(\{\vec{s}\} |T, \mat{Q})} \frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial X} = 0
\end{equation}
where $X$ is one of the parameters we vary.

These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the appendix for the other (more laborious) update rules.
\begin{equation}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{k\in T} \frac{t_k}{\mu^b}\frac{\sum_i Q^b_{k^b_c,i}\left( e^{Q^bt_k} \right)_{i, k^b_p}}{\left( e^{Q^bt_k} \right)_{k^b_c, k^b_p}}
\end{equation}
The individual terms in this sum behave very differently for cases where $k^b_c=k^b_p$ (no change at site $b$ on branch $k$) and $k^b_c\neq k^b_p$ (at least one mutation).
In the limit of short branches $\mu^b t_k\ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain more insightful expressions and we will separate branches with $k^b_c=k^b_p$  and $k^b_c\neq k^b_p$ right away
\begin{equation}
\begin{split}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} &\approx \sum_{k\in T, k^b_c=k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b,i}(\delta_{i,k^b} + t_k Q^b_{i,k^b}\cdots)}{1 - t_k\mu^b\sum_{j}\eqp_j W_{k^bj}+\ldots} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b_c,i}(\delta_{i,k^b_p} + t_k Q^b_{i,k^b_p}\cdots)}{t_k\mu^b\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{p_{k^b_c}W_{k^b_c, l^b_p} + \cdots}{t_k\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} + 	\sum_{k\in T, k^b_c\neq k^b_p} \frac{1}{\mu^b} \\
		&= -\sum_{ij}\eqp_{i}W_{i,j}\langle T_j^b\rangle + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
	\end{split}
\end{equation}
where $\langle T^b_j\rangle$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
This condition (and the corresponding ones in the supplement suggest solution at fixed $\langle T^b_j\rangle$ and $n^b_{ij}$ using the following iterative update:
\begin{equation}
\label{eq:update}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\langle T_j^b\rangle} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\sum_{j\neq i} n^b_{ij}+\delta_{is^{b}}}{\mu^b\sum_{j}\eqp^b_i W_{ij}\langle T_j^b\rangle + \eqp^b_i} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\langle T_j^b\rangle + \eqp_j^b \langle T_i^b\rangle)} \\
	\end{split}
\end{equation}
To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\langle T_j^b\rangle$ -- each appropriately summed over sites or states.

The validity of this iterative solution will depend on the accuracy of the linear approximation made and the degree to which the ancestral states can be reconstructed or summed over.
To assess these two sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\langle T_j^b\rangle$ for a range of sample sizes, tree shapes, and branch lengths.
Fig.~\ref{fig:dressed} shows the accuracy of the inference as quantified by $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \hat{\eqp}_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ the inferred and true equilibrium probabilities.

\begin{figure}[tb]
	\centering
	% \includegraphics[]{}
	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree. (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns.
	(B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected. }
	\label{fig:dressed}
\end{figure}

Outside simulated sequences, the tree and the ancestral states at internal nodes are unknown and need to be reconstructed and summed over.
In the spirit of iterative updates, we first reconstruct those states using a simple evolutionary model such as Jukes-Cantor model, use the resulting ancestral reconstruction to estimate a site-specific model using Eq.~\ref{eq:update}, which can then be used to estimate an improved reconstruction.
Fig.~\ref{fig:reconstructed} shows model inference accuracy for reconstructed trees and sequences and compares them to the accuracy that would have been obtained if internal states had been known and the naive alignment frequencies.
Model inference from reconstructed data is accurate the root to tip distance is small but rapidly looses accuracy if sequences are very diverged.

\bibliography{GTRbib}

\end{document}

### 2019-01-01 src/estimation.py
add JC simulations

@@ -106,10 +106,17 @@ def exact_mu_update(tree, alphabet):
     q = len(alphabet)
     mu_diag = np.zeros((L), dtype=float)
     mu_offdiag = np.zeros((L), dtype=float)
+
+    p_diag = np.zeros((q,L), dtype=float)
+    p_offdiag = np.zeros((q,L), dtype=float)
     for n in tree.tree.get_nonterminals():
         for c in n:
             (m1,m2), (p1,p2), (W1,W2) = tree.get_update_matrices(c)
+            # import ipdb; ipdb.set_trace()
             mu_diag += m1
             mu_offdiag += m2
 
-    return (mu_diag, mu_offdiag)
+            p_diag += p1
+            p_offdiag += p2
+
+    return (mu_diag, mu_offdiag), (p_diag, p_offdiag)

### 2019-01-01 src/generate_toy_data.py

@@ -32,7 +32,7 @@ def load_model(fname):
     return GTR_site_specific.custom(alphabet=d['alphabet'], mu=d['mu'], pi=d['pi'], W=d['W'])
 
 
-def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5):
+def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False):
     from Bio import AlignIO
     # generate a model
     T = betatree(params['n'], alpha=2.0)
@@ -43,8 +43,13 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5):
 
     for mi in range(n_model):
         params['model']=mi
-        myGTR = GTR_site_specific.random(L=params['L'], alphabet='nuc_nogap')
+        if JC:
+            myGTR = GTR_site_specific.random(L=params['L'], alphabet='nuc_nogap',
+                                             pi_dirichlet_alpha=0, W_dirichlet_alpha=0)
+        else:
+            myGTR = GTR_site_specific.random(L=params['L'], alphabet='nuc_nogap')
         myGTR.mu*=params['m']
+
         if out_prefix:
             save_model(myGTR, model_name(out_prefix, params))
 
@@ -69,15 +74,21 @@ def reconstruct_tree(prefix, params):
 
 
 if __name__ == '__main__':
-    L=300
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument("-m", type=float, help="simulated mutation rate")
+    parser.add_argument("-n", type=int, help="number of taxa")
+    parser.add_argument("-L", type=int, default=300, help="length of sequence")
+    args=parser.parse_args()
 
-    prefix = '2018-12-17_simulated_data'
+    L=args.L
+
+    prefix = '2018-12-31_simulated_data_JC'
     if not os.path.isdir(prefix):
         os.mkdir(prefix)
 
-    for n in [1000]:
-        for mu in [0.25, 0.35, 0.5]: #[0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]:
-            for ti in range(3):
-                params = {'L':L, 'n':n, 'm':mu, 'tree':ti}
-                simplex(params, out_prefix=prefix, n_model=3, n_seqgen=3, yule=True)
+    n = args.n
[…]

### 2019-01-01 src/plot_toy_data.py

@@ -13,7 +13,8 @@ if __name__ == '__main__':
     mu_vals = set()
     n_vals = set()
 
-    for fname in glob.glob('2018-12-17_simulated_data_results/*.pkl'):
+    # for fname in glob.glob('2018-12-17_simulated_data_results/*.pkl'):
+    for fname in glob.glob('2018-12-31_simulated_data_JC_results/*.pkl'):
         with open(fname, 'rb') as fh:
             tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu = pickle.load(fh)
 
@@ -37,13 +38,13 @@ if __name__ == '__main__':
     # and the substitution rates from the true frequencies and rates
     for n in n_vals:
         rtt = np.log(n)
-        plt.figure()
+        plt.figure(1)
         for label, data in p_dist.items():
             plt.errorbar(mu_vals*rtt, [np.mean(data[(L,n,mu)]) for mu in mu_vals],
                         [np.std(data[(L,n,mu)]) for mu in mu_vals], label=label)
 
         plt.yscale('log')
-        plt.legend()
+        #plt.legend()
         plt.xlabel('average root-to-tip distance')
         plt.ylabel('squared deviation')
 

### 2019-01-01 src/reconstruct_toy_data.py

@@ -25,7 +25,7 @@ if __name__ == '__main__':
     parser.add_argument("-L", type=int, help="length of sequence")
     args=parser.parse_args()
 
-    prefix = '2018-12-26_simulated_data'
+    prefix = '2018-12-31_simulated_data_JC'
     mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
     files = glob.glob(prefix+mask)
 
@@ -43,7 +43,7 @@ if __name__ == '__main__':
 
     analysis_types = ['naive', 'single', 'dressed', 'regular', 'phylo', 'marginal', 'true_model', 'iterative']
 
-    for fname in files[:1]:
+    for fname in files:
         print(fname)
 
         params = parse_alignment_name(fname)
@@ -75,8 +75,10 @@ if __name__ == '__main__':
                         bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
                         model = estimate_GTR(mc[0], pc=pc, single_site=False, bl=None)
                         if i:
-                            mu_num, mu_denom = exact_mu_update(mc[-1], 'nuc_nogap')
-                            model.mu *= mu_num/mu_denom
+                            (mu_denom, mu_num), (p_num, p_denom) = exact_mu_update(mc[-1], model.alphabet)
+                            model.mu = model.mu*(mu_num)/(mu_denom)
+                            print(mu_num[:5],mu_denom[:5], mu_num[:5]/mu_denom[:5], model.mu.mean())
+                            # model.Pi *= p_num/p_denom
                 else:
                     rec_model = {'true_model':true_model}
                     mc = reconstruct_counts(prefix, params, gtr=rec_model.get(ana, 'JC69'),
@@ -85,7 +87,8 @@ if __name__ == '__main__':
 
                     bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
 
-                model = estimate_GTR(mc[0], pc=pc, single_site=ana=='single', bl=bl)
+                if ana!='iterative':
+                    model = estimate_GTR(mc[0], pc=pc, single_site=ana=='single', bl=None)
                 avg_rate[ana][dset].append((true_model_average_rate, model.average_rate().mean()))
                 _mc, model_LH, _t = reconstruct_counts(prefix, params, gtr=model, alphabet='nuc_nogap',
                                                  marginal=True, reconstructed_tree=ana=='phylo')

### 2019-01-01 submit_gendata.py

@@ -2,4 +2,4 @@ import os
 
 for n in [300, 1000]:
     for mu in [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]:
-        os.system('sbatch submit.sh src/reconstruct_toy_data.py -L 300 -n {n} -m {mu}'.format(n=n, mu=mu))
+        os.system('sbatch submit.sh src/generate_toy_data.py -L 300 -n {n} -m {mu}'.format(n=n, mu=mu))

### 2019-01-01 src/generate_toy_data.py
switch to iqtree

@@ -35,7 +35,7 @@ def load_model(fname):
 def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False):
     from Bio import AlignIO
     # generate a model
-    T = betatree(params['n'], alpha=2.0)
+    T = betatree.betatree(params['n'], alpha=2.0)
     T.yule=yule
     T.coalesce()
     if out_prefix:
@@ -61,16 +61,23 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=Fa
 
             if out_prefix:
                 save_mutation_count(mySeq, mutation_count_name(out_prefix, params))
-                with gzip.open(alignment_name(out_prefix, params), 'wt') as fh:
+                with open(alignment_name(out_prefix, params), 'wt') as fh:
                     AlignIO.write(mySeq.get_aln(), fh, 'fasta')
                 reconstruct_tree(out_prefix, params)
 
 
 def reconstruct_tree(prefix, params):
-    #call = ['iqtree', '-s', alignment_name(prefix, params), '-st', 'DNA', '-nt', '2', '-m', 'JC']
-    fname =  alignment_name(prefix, params)
-    call = ['gunzip -c' ,fname, '|', 'fasttree', '-nt', '>', reconstructed_tree_name(prefix, params)]
-    os.system(' '.join(call))
+    aln_file = alignment_name(prefix, params)
+    fast_opts = [
+        "-ninit", "2",
+        "-n",     "2",
+        "-me",    "0.05"
+    ]
+    call = ["iqtree"] + fast_opts +["-nt 1", "-s", aln_file, "-m", 'GTR',
+            ">", "iqtree.log"]
+    os.system(" ".join(call))
+    os.system("mv %s.treefile %s"%(aln_file, reconstructed_tree_name(prefix, params)))
+    os.system("rm %s.*"%aln_file)
 
 
 if __name__ == '__main__':
@@ -90,5 +97,5 @@ if __name__ == '__main__':
     mu = args.m
     for ti in range(3):
         params = {'L':L, 'n':n, 'm':mu, 'tree':ti}
-        simplex(params, out_prefix=prefix, n_model=3, n_seqgen=3, yule=True, JC=True)
+        simplex(params, out_prefix=prefix, n_model=3, n_seqgen=3, yule=True, JC=False)
 

### 2019-01-01 src/reconstruct_toy_data.py

@@ -41,7 +41,7 @@ if __name__ == '__main__':
     pc=0.01
     niter=5
 
-    analysis_types = ['naive', 'single', 'dressed', 'regular', 'phylo', 'marginal', 'true_model', 'iterative']
+    analysis_types = ['naive', 'single', 'dressed', 'regular', 'marginal', 'true_model', 'iterative']
 
     for fname in files:
         print(fname)
@@ -71,7 +71,7 @@ if __name__ == '__main__':
                     for i in range(niter):
                         mc = reconstruct_counts(prefix, params, gtr=model,
                                                 alphabet='nuc_nogap', marginal=True,
-                                                reconstructed_tree=False)
+                                                reconstructed_tree=True)
                         bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
                         model = estimate_GTR(mc[0], pc=pc, single_site=False, bl=None)
                         if i:
@@ -83,7 +83,7 @@ if __name__ == '__main__':
                     rec_model = {'true_model':true_model}
                     mc = reconstruct_counts(prefix, params, gtr=rec_model.get(ana, 'JC69'),
                                             alphabet='nuc_nogap', marginal=ana in ['marginal', 'true_model', 'iterative'],
-                                            reconstructed_tree=ana=='phylo')
+                                            reconstructed_tree=ana!='dressed')
 
                     bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
 

### 2019-01-02 manuscript/ms.tex
small edits
reorg plotting script

@@
 	\centering
-	% \includegraphics[]{}
+	\includegraphics[width=0.48\textwidth]{../figures/p_dist_vs_treelength.pdf}
+	\includegraphics[width=0.48\textwidth]{../figures/avg_rate_dressed.pdf}
 	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree.
 (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences.
 	(B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected. }
@@
 Fig.~\ref{fig:reconstructed} shows model inference accuracy for reconstructed trees and sequences and compares them to the accuracy that would have been obtained if internal states had been known and the naive alignment frequencies.
-Model inference from reconstructed data is accurate the root to tip distance is small but rapidly looses accuracy if sequences are very diverged.
+Model inference from reconstructed data is accurate if the root to tip distance is small but rapidly looses accuracy if sequences are very diverged.
+
+\begin{figure}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{../figures/p_dist_vs_rtt.pdf}
+	\caption{At large root-to-tip distances, ancestral reconstruction becomes less and less accurate and estimation of equilibrium state frequencies fails. }
+	\label{fig:reconstructed}
+\end{figure}
+
+\subsection*{Non-linear update rules}
+
+\subsection*{Applications to large HIV alignments}
 

### 2019-01-02 src/estimation.py

@@ -9,7 +9,7 @@ from treetime.seq_utils import seq2prof, profile_maps, alphabets
 from filenames import *
 
 def p_from_aln(in_prefix, params, alphabet='nuc_nogap'):
-    with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
+    with open(alignment_name(in_prefix, params), 'rt') as fh:
         aln = AlignIO.read(fh, 'fasta')
 
     alpha = alphabets[alphabet]
@@ -22,7 +22,7 @@ def p_from_aln(in_prefix, params, alphabet='nuc_nogap'):
 
 def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap',
                        marginal=False, reconstructed_tree=False):
-    with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
+    with open(alignment_name(in_prefix, params), 'rt') as fh:
         aln = AlignIO.read(fh, 'fasta')
     tree_fname =  reconstructed_tree_name(in_prefix, params) if reconstructed_tree else tree_name(in_prefix, params)
     myTree = TreeAnc(gtr=gtr, alphabet=alphabet,

### 2019-01-02 src/generate_toy_data.py

@@ -35,7 +35,7 @@ def load_model(fname):
 def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False):
     from Bio import AlignIO
     # generate a model
-    T = betatree.betatree(params['n'], alpha=2.0)
+    T = betatree(params['n'], alpha=2.0)
     T.yule=yule
     T.coalesce()
     if out_prefix:
@@ -85,17 +85,19 @@ if __name__ == '__main__':
     parser.add_argument("-m", type=float, help="simulated mutation rate")
     parser.add_argument("-n", type=int, help="number of taxa")
     parser.add_argument("-L", type=int, default=300, help="length of sequence")
+    parser.add_argument("--JC", action='store_true', help="simulate JC model")
+    parser.add_argument("--prefix", type=str, help="folder to save data")
     args=parser.parse_args()
 
     L=args.L
 
-    prefix = '2018-12-31_simulated_data_JC'
+    prefix = args.prefix
     if not os.path.isdir(prefix):
         os.mkdir(prefix)
 
     n = args.n
     mu = args.m
-    for ti in range(3):
+    for ti in range(2):
         params = {'L':L, 'n':n, 'm':mu, 'tree':ti}
-        simplex(params, out_prefix=prefix, n_model=3, n_seqgen=3, yule=True, JC=False)
+        simplex(params, out_prefix=prefix, n_model=2, n_seqgen=2, yule=True, JC=args.JC)
 

### 2019-01-02 src/plot_toy_data.py

@@ -2,10 +2,15 @@ import glob, pickle
 import numpy as np
 from collections import defaultdict
 
-if __name__ == '__main__':
+fmts = ['.png', '.pdf']
 
+labels = {'naive':"Alignment frequencies", "dressed":"True counts", "regular":"reconstructed", "marginal":"ancestral sum"}
+colors = {k:'C%d'%(i+1) for i,k in enumerate(sorted(labels.keys()))}
+
+def load_toy_data_results(path):
     mu_dist = defaultdict(lambda: defaultdict(list))
     p_dist = defaultdict(lambda: defaultdict(list))
+    p_entropy = defaultdict(lambda: defaultdict(list))
     W_dist = defaultdict(lambda: defaultdict(list))
     delta_LH = defaultdict(lambda: defaultdict(list))
     avg_rate = defaultdict(lambda: defaultdict(list))
@@ -13,10 +18,9 @@ if __name__ == '__main__':
     mu_vals = set()
     n_vals = set()
 
-    # for fname in glob.glob('2018-12-17_simulated_data_results/*.pkl'):
-    for fname in glob.glob('2018-12-31_simulated_data_JC_results/*.pkl'):
+    for fname in glob.glob(path+'*.pkl'):
         with open(fname, 'rb') as fh:
-            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu = pickle.load(fh)
+            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_p_entropy, tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu = pickle.load(fh)
 
         n_vals.update(tmp_n_vals)
         mu_vals.update(tmp_mu_vals)
@@ -27,82 +31,141 @@ if __name__ == '__main__':
                 for y in tmp_d[x]:
                     d[x][y].extend(tmp_d[x][y])
 
-
-    L=300
     n_vals = np.array(sorted(n_vals))
     mu_vals = np.array(sorted(mu_vals))
 
-    from matplotlib import pyplot as plt
-
-    # for each data set size, plot the distance of the inferred equilibrium frequencies
-    # and the substitution rates from the true frequencies and rates
-    for n in n_vals:
-        rtt = np.log(n)
-        plt.figure(1)
-        for label, data in p_dist.items():
[…]

### 2019-01-02 src/reconstruct_toy_data.py

@@ -23,14 +23,16 @@ if __name__ == '__main__':
     parser.add_argument("-m", type=float, help="simulated mutation rate")
     parser.add_argument("-n", type=int, help="number of taxa")
     parser.add_argument("-L", type=int, help="length of sequence")
+    parser.add_argument("--prefix", type=str, help="folder to save data")
     args=parser.parse_args()
 
-    prefix = '2018-12-31_simulated_data_JC'
+    prefix = args.prefix
     mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
     files = glob.glob(prefix+mask)
 
     mu_dist = defaultdict(lambda: defaultdict(list))
     p_dist = defaultdict(lambda: defaultdict(list))
+    p_entropy = defaultdict(lambda: defaultdict(list))
     W_dist = defaultdict(lambda: defaultdict(list))
     delta_LH = defaultdict(lambda: defaultdict(list))
     avg_rate = defaultdict(lambda: defaultdict(list))
@@ -41,7 +43,7 @@ if __name__ == '__main__':
     pc=0.01
     niter=5
 
-    analysis_types = ['naive', 'single', 'dressed', 'regular', 'marginal', 'true_model', 'iterative']
+    analysis_types = ['naive', 'single', 'dressed', 'regular', 'marginal', 'true_model']
 
     for fname in files:
         print(fname)
@@ -72,31 +74,29 @@ if __name__ == '__main__':
                         mc = reconstruct_counts(prefix, params, gtr=model,
                                                 alphabet='nuc_nogap', marginal=True,
                                                 reconstructed_tree=True)
-                        bl = [n.branch_length for n in mc[-1].tree.find_clades() if n!=mc[-1].tree.root]
-                        model = estimate_GTR(mc[0], pc=pc, single_site=False, bl=None)
+                        model = estimate_GTR(mc[0], pc=pc, single_site=False)
                         if i:
                             (mu_denom, mu_num), (p_num, p_denom) = exact_mu_update(mc[-1], model.alphabet)
                             model.mu = model.mu*(mu_num)/(mu_denom)
-                            print(mu_num[:5],mu_denom[:5], mu_num[:5]/mu_denom[:5], model.mu.mean())
                             # model.Pi *= p_num/p_denom
                 else:
                     rec_model = {'true_model':true_model}
[…]

### 2019-01-11 manuscript/ms.tex
edits to ms, include bib

@@
 
+\section*{Introduction}
+Over time, genome sequences change through mutations and are reshuffeled by recombination.
+Modifications to the genomes are filtered by selection for survival such that beneficial variants spread preferentially and those that impair function are rare.
+As a result, some parts of genomes change rapidly, while other are strongly conserved.
+In addition to variation in rate, different sites in a genome only explore subsets of the available states.
+Some positions in a protein, for example, might only allow for hydrophobic amino acids, while other require acidic side chains.
+
+Phylogenetics aims at reconstructing the relationships and history of homologous sequences from the the substitutions that occurred in the past.
+Modern phylogenetic methods describe this stochastic evolutionary process with probabilistic models of sequence evolution and aim to find phylogenies that either maximize the likelihood of observing the alignment or sample phylogenies from a posterior probability distribution \citep{felsenstein2004inferring}.
+
+Inferring phylogenies is a computationally challenging problem since (i) the number of phylogenies grows super-exponentially with the number of taxa, (ii) calculation of the likelihood is compulationally costly (though linear in $n$), (iii) substitution models and the tree have many unknown continuous parameters that need estimation.
+Here, we will focus on the latter problem and describe ways to estimate substitution models in computationally efficient and economically parameterized form.
+
+The simplest substitution models assume that all sites and sequence states are equivalent and evolve at the same rate, i.e., they assume an unconstrained non-functional sequence that mutates at random between the different sequence states.
+Such simple models are clearly inadequate.
+But increasing the number of parameters, although improving the likelihood, does not necessarily increases the statistical/predictive power of the model \cite{scheffler2014validity}.
+In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{felsenstein2001taking,Rodrigue557,kumar2011statistics,spielman2016extensively}.
+%
+A popular way to overcome this problem is by using mixture models, where every site is assigned to a category with its substitution rate drawn randomly from a statistical distribution assigned to this category \cite{blanquart2008site,mayrose2005gamma,jia2014impact,kosakovsky2004simple,le2012modeling,jayaswal2014mixture}.
+Although this significantly reduces the number of free parameters, choosing the number of the categories and specifying the corresponding statistical models remains rather arbitrary.
+
+Furthermore, rate variation is likely not the most crucial aspect at which positions in a protein differ.
+For the biological function of a protein, the properties of amino acids at different positions are crucial and experimentally measured amino-acid preferences at for each alignment column have been shown to greatly improve the phylogenetic fit \cite{bloom2014experimentally,bloom2014experimentally1}.
+
+\section{Methods}
 Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
@@
 
+% To regularize the parameter estimates when the data are not sufficiently informative to constrain them, we use Dirichlet and Gamma priors for $\eqp_i^a$ and $\mu^a$.
+% \begin{equation}
+% 	\log P(p_i^a) = C + (\alpha-1) \log \eqp_i^a \quad \mathrm{and} \quad \log P(\mu^a) = C -\frac{\beta \mu^a}{\bar{\mu}} + (\beta - 1)\log \mu^a
+% \end{equation}
+
 The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
@@
 	\begin{split}
-		\mu^b & \leftarrow \mu^b\frac{\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\langle T_j^b\rangle} \\
-		\eqp^b_i & \leftarrow \eqp^b_i\frac{\sum_{j\neq i} n^b_{ij}+\delta_{is^{b}}}{\mu^b\sum_{j}\eqp^b_i W_{ij}\langle T_j^b\rangle + \eqp^b_i} \\
+		\mu^b & \leftarrow \mu^b\frac{C+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\langle T_j^b\rangle} \\
+		\eqp^b_i & \leftarrow \eqp^b_i\frac{C+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(qC+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\langle T_j^b\rangle} \\
 		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\langle T_j^b\rangle + \eqp_j^b \langle T_i^b\rangle)} \\
@@
 \end{equation}
+where $C$ is a pseudocount analogous to a Dirichlet prior that  in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $C$ over the total tree length.
 To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
@@
 
-The validity of this iterative solution will depend on the accuracy of the linear approximation made and the degree to which the ancestral states can be reconstructed or summed over.
+The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which the ancestral states can be reconstructed or summed over, and the accuracy of the tree reconstruction.
 To assess these two sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\langle T_j^b\rangle$ for a range of sample sizes, tree shapes, and branch lengths.
-Fig.~\ref{fig:dressed} shows the accuracy of the inference as quantified by $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \hat{\eqp}_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ the inferred and true equilibrium probabilities.
+Fig.~\ref{fig:dressed} shows the accuracy of the inference as quantified by $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ the inferred and true equilibrium probabilities.
 
@@
 In the spirit of iterative updates, we first reconstruct those states using a simple evolutionary model such as Jukes-Cantor model, use the resulting ancestral reconstruction to estimate a site-specific model using Eq.~\ref{eq:update}, which can then be used to estimate an improved reconstruction.
-Fig.~\ref{fig:reconstructed} shows model inference accuracy for reconstructed trees and sequences and compares them to the accuracy that would have been obtained if internal states had been known and the naive alignment frequencies.
-Model inference from reconstructed data is accurate if the root to tip distance is small but rapidly looses accuracy if sequences are very diverged.
+
+Fig.~\ref{fig:reconstructed}A shows model inference accuracy for reconstructed trees (using iq-tree) and ancestral sequence inferred by TreeTime using a Jukes-Cantor model.
+The estimates of $\eqp_i^a$ are about as accurate as those inferred from the true ancestral states und the root-to-tip distance reaches about 0.3 but then start to deteriorate and are eventually even worse than the naive estimates from the alignment.
+This accuracy can be improved moderately by summing over uncertainty of the ancestral reconstruction.
+
+A more substantial improvment in accuracy can be obtained when the inferred model is used to re-infer the ancestral states, followed by repeated model inference.
+With this iterative procedure, model accuracy is comparable to that with known ancestral states up to root-to-tip distance of about 1.0.
+
+In absense of strong regularization, maximum likelihood estimates of site specific equilibrium probabilities tend to `overfit' the data when the total number of observed mutations is low.
+This is reflected in Fig.~\ref{fig:reconstructed}B showing the different in entropy $-\sum_i \eqp_i^a \log \eqp_i^a$ of the inferred and true distributions.
+At low root-to-tip distances, the inferred distributions have too little entropy, meaning they are too concentrates on the few states that are observed.
+This effect, however, is much less pronounced in the tree based inference compared to naive estimates directly from the alignment.
+At very large root-to-tip distances, both the naive and the estimate based on true ancestral states become unbiased, while the model inference based on reconstructions results in broader $\eqp_i^a$ with higher entropy.
 
@@
 	\includegraphics[width=0.48\textwidth]{../figures/p_dist_vs_rtt.pdf}
-	\caption{At large root-to-tip distances, ancestral reconstruction becomes less and less accurate and estimation of equilibrium state frequencies fails. }
+	\includegraphics[width=0.48\textwidth]{../figures/p_entropy_vs_rtt.pdf}
+	\caption{Ancestral reconstruction: (A) At large root-to-tip distances, ancestral reconstruction becomes less and less accurate and estimation of equilibrium state frequencies fails.
+(B) For closely related sequences, the estimated equilibrium frequencies tend to be too concentrated in a few sites (low entropy), while at very large root-to-tip distances the estimates are too flat (high entropy).
+}
 	\label{fig:reconstructed}

### 2019-01-11 src/compare_tree_length.py

@@ -1,0 +1,71 @@
+import glob
+from matplotlib import pyplot as plt
+from Bio import Phylo, AlignIO
+from generate_toy_data import *
+from filenames import *
+from estimation import *
+from treetime import TreeAnc
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument("-m", type=float, help="simulated mutation rate")
+    parser.add_argument("-n", type=int, help="number of taxa")
+    parser.add_argument("-L", type=int, help="length of sequence")
+    parser.add_argument("--prefix", type=str, required=True, help="folder to retrieve data from")
+    args=parser.parse_args()
+
+    prefix = args.prefix
+    mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
+    files = glob.glob(prefix+mask)
+
+    length = []
+    depth = []
+    for fname in files:
+        print(fname)
+        params = parse_alignment_name(fname)
+        true_tree = Phylo.read(tree_name(prefix, params), 'newick')
+        true_model = load_model(model_name(prefix, params))
+        aln = AlignIO.read(alignment_name(prefix, params), 'fasta')
+        iq_tree = Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
+
+        iq_tree_ttl = 0.5*iq_tree.total_branch_length()/params['n']
+        iq_tree_depth = np.mean([x for c,x in iq_tree.depths().items() if c.is_terminal()])
+
+        m = true_model.average_rate().mean()
+        true_model.mu/=m
+        tt = TreeAnc(tree=iq_tree, aln=aln, gtr = true_model, reduce_alignment=False)
+        tt.infer_ancestral_sequences(marginal=True)
+        for i in range(5):
+            x = []
+            for n in tt.tree.find_clades():
+                if n.up is None:
+                    continue
+                x.append([tt.optimal_marginal_branch_length(n), n.branch_length])
+                n.branch_length = 0.5*(x[-1][0] + n.branch_length)
+
+            x = np.array(x)
+            print(x.sum(axis=0))
+            tt.infer_ancestral_sequences(marginal=True)
+            print(tt.sequence_LH())
+
[…]

### 2019-01-11 src/estimation.py

@@ -9,7 +9,7 @@ from treetime.seq_utils import seq2prof, profile_maps, alphabets
 from filenames import *
 
 def p_from_aln(in_prefix, params, alphabet='nuc_nogap'):
-    with open(alignment_name(in_prefix, params), 'rt') as fh:
+    with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
         aln = AlignIO.read(fh, 'fasta')
 
     alpha = alphabets[alphabet]
@@ -21,13 +21,16 @@ def p_from_aln(in_prefix, params, alphabet='nuc_nogap'):
 
 
 def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap',
-                       marginal=False, reconstructed_tree=False):
-    with open(alignment_name(in_prefix, params), 'rt') as fh:
-        aln = AlignIO.read(fh, 'fasta')
-    tree_fname =  reconstructed_tree_name(in_prefix, params) if reconstructed_tree else tree_name(in_prefix, params)
-    myTree = TreeAnc(gtr=gtr, alphabet=alphabet,
-                     tree=tree_fname, aln=aln,
-                     reduce_alignment=False, verbose = 0)
+                       marginal=False, reconstructed_tree=False, tt=None):
+    if tt:
+        myTree=tt
+    else:
+        with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
+            aln = AlignIO.read(fh, 'fasta')
+        tree_fname =  reconstructed_tree_name(in_prefix, params) if reconstructed_tree else tree_name(in_prefix, params)
+        myTree = TreeAnc(gtr=gtr, alphabet=alphabet,
+                         tree=tree_fname, aln=aln,
+                         reduce_alignment=False, verbose = 0)
 
     if type(gtr)==str:
         myTree.infer_ancestral_sequences(marginal=True, infer_gtr=True, normalized_rate=False)
@@ -38,16 +41,22 @@ def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap',
     return mutation_counts, myTree.sequence_LH(), myTree
 
 
-def estimate_GTR(mutation_counts, pc=0.1, single_site=False, bl=None):
+def estimate_GTR(mutation_counts, pc=0.1, single_site=False, tt=None):
[…]

### 2019-01-11 src/filenames.py

@@ -27,6 +27,9 @@ def mutation_count_name(prefix, params):
 def alignment_name(prefix, params):
     return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}.fasta.gz'.format(**params)
 
+def alignment_name_raw(prefix, params):
+    return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}.fasta'.format(**params)
+
 def tree_name(prefix, params):
     return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}.nwk'.format(**params)
 

### 2019-01-11 src/generate_toy_data.py

@@ -61,19 +61,19 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=Fa
 
             if out_prefix:
                 save_mutation_count(mySeq, mutation_count_name(out_prefix, params))
-                with open(alignment_name(out_prefix, params), 'wt') as fh:
+                with open(alignment_name_raw(out_prefix, params), 'wt') as fh:
                     AlignIO.write(mySeq.get_aln(), fh, 'fasta')
                 reconstruct_tree(out_prefix, params)
 
 
 def reconstruct_tree(prefix, params):
-    aln_file = alignment_name(prefix, params)
+    aln_file = alignment_name_raw(prefix, params)
     fast_opts = [
         "-ninit", "2",
         "-n",     "2",
         "-me",    "0.05"
     ]
-    call = ["iqtree"] + fast_opts +["-nt 1", "-s", aln_file, "-m", 'GTR',
+    call = ["iqtree"] + fast_opts +["-nt 1", "-s", aln_file, "-m", 'GTR+G10',
             ">", "iqtree.log"]
     os.system(" ".join(call))
     os.system("mv %s.treefile %s"%(aln_file, reconstructed_tree_name(prefix, params)))

### 2019-01-11 src/global_opt.py

@@ -1,0 +1,99 @@
+import os, gzip, glob, pickle
+import numpy as np
+from collections import defaultdict
+
+from scipy.optimize import minimize
+
+from treetime.treeanc import TreeAnc
+from treetime.gtr_site_specific import GTR_site_specific
+from treetime.gtr import GTR
+from treetime.seq_utils import seq2prof, profile_maps, alphabets
+
+from generate_toy_data import *
+from filenames import *
+from estimation import *
+from reconstruct_toy_data import *
+
+def lh(x,tt):
+    for ni, n in enumerate(tt.tree.find_clades()):
+        n.branch_length = x[ni]**2
+
+    tt.infer_ancestral_sequences(marginal=True)
+    print(tt.sequence_LH(), np.sum(x**2))
+    return -tt.sequence_LH()
+
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument("-m", type=float, help="simulated mutation rate")
+    parser.add_argument("-n", type=int, help="number of taxa")
+    parser.add_argument("-L", type=int, help="length of sequence")
+    parser.add_argument("--prefix", type=str, help="folder to save data")
+    args=parser.parse_args()
+
+    prefix = args.prefix
+    mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
+    files = glob.glob(prefix+mask)
+
+    pc=1
+    niter=5
+
+    for fname in files[:1]:
+        print(fname)
+
+        params = parse_alignment_name(fname)
+        true_model = load_model(model_name(prefix, params))
+        true_model_average_rate = true_model.average_rate().mean()
+        true_mut_counts = load_mutation_count(mutation_count_name(prefix, params))
+        dset = (params['L'], params['n'], params['m'])
+
+        _mc, true_LH, true_tree = reconstruct_counts(prefix, params, gtr=true_model, alphabet='nuc_nogap',
+                                              marginal=True, reconstructed_tree=False)
+        true_tree_length = true_tree.tree.total_branch_length()
+
+        model = 'JC69'
+        for i in range(niter):
+            mc = reconstruct_counts(prefix, params, gtr=model,
+                                    alphabet='nuc_nogap', marginal=True,
+                                    reconstructed_tree=False)
[…]

### 2019-01-11 src/midpoint_root_trees.py

@@ -1,0 +1,21 @@
+import glob
+from Bio import Phylo
+from generate_toy_data import *
+from filenames import *
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="reroot all reconstructed trees")
+    parser.add_argument("--prefix", type=str, required=True, help="folder to retrieve data from")
+    args=parser.parse_args()
+
+    prefix = args.prefix
+    mask = "/L*_n*_m*_*fasta.gz"
+    files = glob.glob(prefix+mask)
+
+    for fname in files:
+        print(fname)
+        params = parse_alignment_name(fname)
+        iq_tree = Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
+        iq_tree.root_at_midpoint()
+        Phylo.write(iq_tree, reconstructed_tree_name(prefix, params), 'newick')

### 2019-01-11 src/plot_toy_data.py

@@ -4,7 +4,9 @@ from collections import defaultdict
 
 fmts = ['.png', '.pdf']
 
-labels = {'naive':"Alignment frequencies", "dressed":"True counts", "regular":"reconstructed", "marginal":"ancestral sum"}
+labels = {'naive':"Alignment frequencies", "dressed":"Inference from true substitution counts", "regular":"Reconstructed",
+          "marginal":"ancestral sum", "iterative":"Iterated", 'branch_length':"non-linear",
+          "iterative_true":"iterative, true tree", 'marginal_true':"ancestral sum, true tree"}
 colors = {k:'C%d'%(i+1) for i,k in enumerate(sorted(labels.keys()))}
 
 def load_toy_data_results(path):
@@ -26,7 +28,7 @@ def load_toy_data_results(path):
         mu_vals.update(tmp_mu_vals)
 
         for d,tmp_d in [(mu_dist, tmp_mu_dist), (W_dist, tmp_W_dist), (p_dist, tmp_p_dist),
-                        (delta_LH, tmp_LH), (avg_rate, tmp_avg_mu)]:
+                        (p_entropy, tmp_p_entropy), (delta_LH, tmp_LH), (avg_rate, tmp_avg_mu)]:
             for x in tmp_d:
                 for y in tmp_d[x]:
                     d[x][y].extend(tmp_d[x][y])
@@ -95,6 +97,30 @@ def plot_pdist_vs_rtt(data, n_vals, mu_vals, methods=None, fname=None):
             plt.savefig(fname+fmt)
 
 
+def plot_pentropy_vs_rtt(data, n_vals, mu_vals, methods=None, fname=None):
+    # for each data set size, plot the difference in entropy between the inferred and true
+    # equilibrium frequencies.
+
+    plt.figure()
+    for n in n_vals:
+        rtt = np.log(n)
+        for label, d in data['S'].items():
+            if methods and label not in methods:
+                continue
+            plt.errorbar(mu_vals*rtt, [-np.mean(np.diff(d[(L,n,mu)], axis=1)) for mu in mu_vals],
+                        [np.std(np.diff(d[(L,n,mu)], axis=1)) for mu in mu_vals],
+                        c=colors[label], label=labels[label] if n==n_vals[0] else '')
+
+    plt.xscale('log')
+    plt.legend()
+    plt.xlabel('root-to-tip distance')
+    plt.ylabel('entropy difference (inferred-true)')
+
+    if fname:
[…]

### 2019-01-11 src/reconstruct_toy_data.py

@@ -40,10 +40,10 @@ if __name__ == '__main__':
     mu_vals = set()
     n_vals = set()
 
-    pc=0.01
+    pc=0.5
     niter=5
 
-    analysis_types = ['naive', 'single', 'dressed', 'regular', 'marginal', 'true_model']
+    analysis_types = ['naive', 'single', 'dressed', 'regular', 'regular_true', 'marginal','marginal_true', 'iterative', 'iterative_true']
 
     for fname in files:
         print(fname)
@@ -57,35 +57,44 @@ if __name__ == '__main__':
         true_mut_counts = load_mutation_count(mutation_count_name(prefix, params))
         dset = (params['L'], params['n'], params['m'])
 
-        _mc, true_LH, _t = reconstruct_counts(prefix, params, gtr=true_model, alphabet='nuc_nogap',
+        _mc, true_LH, true_tree = reconstruct_counts(prefix, params, gtr=true_model, alphabet='nuc_nogap',
                                               marginal=True, reconstructed_tree=False)
-
+        true_tree_length = true_tree.tree.total_branch_length()
         for ana in analysis_types:
             if ana=='naive':
                 naive = p_from_aln(prefix, params)
                 p_dist[ana][dset].append(np.mean([chisq(naive[:,i],true_model.Pi[:,i]) for i in range(params['L'])]))
+                p_entropy[ana][dset].append([-np.mean(np.sum(naive*np.log(naive+1e-10), axis=0)),
+                                            -np.mean(np.sum(true_model.Pi*np.log(true_model.Pi), axis=0))])
             else:
                 if ana=='dressed':
                     mc = (true_mut_counts, None, None)
                     bl = [n.branch_length for n in _t.tree.find_clades() if n!=_t]
-                elif ana=='iterative':
+                elif ana in ['iterative', 'iterative_true']:
                     model = 'JC69'
                     for i in range(niter):
                         mc = reconstruct_counts(prefix, params, gtr=model,
                                                 alphabet='nuc_nogap', marginal=True,
-                                                reconstructed_tree=True)
+                                                reconstructed_tree=ana=='iterative')
                         model = estimate_GTR(mc[0], pc=pc, single_site=False)
-                        if i:
-                            (mu_denom, mu_num), (p_num, p_denom) = exact_mu_update(mc[-1], model.alphabet)
-                            model.mu = model.mu*(mu_num)/(mu_denom)
-                            # model.Pi *= p_num/p_denom
[…]

### 2019-01-28 manuscript/ms.tex
added branch length files, edits to manuscript, etc

@@
 
-\section{Methods}
+\citep{felsenstein1981evolutionary}
+
+More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
+Nevertheless, most phylogenetic inference assumes uniform frequencies across sites and only models rate variation.
+\citet{hilton_modeling_2018} recently showed that experimentally measured preference not only improve the phylogenetic fit, but also results in longer branch length estimates.
+We show here that this effect is dramatic and can result in complete insensitivity of the likelihood to the actual branch length.
+Hence site-specific models are not only interesting to learn about functional constraints of biological sequence, but might also be crucial to properly reconstruct the phylogenies.
+
+\section*{Efficient inference of site specific substitution models}
 Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
@@
 
-Instead of inferring completely independent models for every site in the genome, we only allow for site specific rates and equilibrium frequencies while using the same transition matrix for every site.
+Instead of inferring completely independent models for every site in the genome, we follow \citet{halpern_evolutionary_1998} and only allow for site specific rates and equilibrium frequencies while using the same transition matrix for every site.
 Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
@@
 
+\subsection*{Iterative Likelihood maximation}
 The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
@@
 
-The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which the ancestral states can be reconstructed or summed over, and the accuracy of the tree reconstruction.
-To assess these two sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\langle T_j^b\rangle$ for a range of sample sizes, tree shapes, and branch lengths.
-Fig.~\ref{fig:dressed} shows the accuracy of the inference as quantified by $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ the inferred and true equilibrium probabilities.
-
 \begin{figure}[tb]
@@
 
+\subsection*{Model inference accuracy}
+
+The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which the ancestral states can be reconstructed or summed over, and the accuracy of the tree reconstruction.
+To assess these three sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\langle T_j^b\rangle$ for a range of sample sizes, tree shapes, and branch lengths.
+We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
+
+In the case of simulation data, we can calculate the $n^b_{ij}$ and $\langle T_i^b\rangle$ needed for the iterative optimization directly from the tree and the associated ancestral sequences.
+This allows us to test the effect of the linear approximation directly.
+Fig.~\ref{fig:dressed} shows $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
+The plot is shown on double logarithmic scales, such that the approximatly straight decrease of $\chi^2$ implies that the accuracy is inversely proportional to the expected number of substitutions.
+This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
+The linear approximation does not seem to affect the estimation of the equilibrium frequencies much.
+By contrast, using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry.
+
+The accuracy of substitution rates, however, is affected by linearization and we consistently underestimate $\mu^b$, see Fig.~\ref{fig:dressed}B.
+This is expected as the linearlized update ignores cases where the same site changes twice along a branch.
+We will see later that the linear substitution model inference followed by full non-linear branch length optimization rectifies this problem.
+
 Outside simulated sequences, the tree and the ancestral states at internal nodes are unknown and need to be reconstructed and summed over.
-In the spirit of iterative updates, we first reconstruct those states using a simple evolutionary model such as Jukes-Cantor model, use the resulting ancestral reconstruction to estimate a site-specific model using Eq.~\ref{eq:update}, which can then be used to estimate an improved reconstruction.
-
-Fig.~\ref{fig:reconstructed}A shows model inference accuracy for reconstructed trees (using iq-tree) and ancestral sequence inferred by TreeTime using a Jukes-Cantor model.
-The estimates of $\eqp_i^a$ are about as accurate as those inferred from the true ancestral states und the root-to-tip distance reaches about 0.3 but then start to deteriorate and are eventually even worse than the naive estimates from the alignment.
-This accuracy can be improved moderately by summing over uncertainty of the ancestral reconstruction.
-
-A more substantial improvment in accuracy can be obtained when the inferred model is used to re-infer the ancestral states, followed by repeated model inference.
-With this iterative procedure, model accuracy is comparable to that with known ancestral states up to root-to-tip distance of about 1.0.
-
-In absense of strong regularization, maximum likelihood estimates of site specific equilibrium probabilities tend to `overfit' the data when the total number of observed mutations is low.
-This is reflected in Fig.~\ref{fig:reconstructed}B showing the different in entropy $-\sum_i \eqp_i^a \log \eqp_i^a$ of the inferred and true distributions.
-At low root-to-tip distances, the inferred distributions have too little entropy, meaning they are too concentrates on the few states that are observed.
-This effect, however, is much less pronounced in the tree based inference compared to naive estimates directly from the alignment.
-At very large root-to-tip distances, both the naive and the estimate based on true ancestral states become unbiased, while the model inference based on reconstructions results in broader $\eqp_i^a$ with higher entropy.
+We used the simulated sequence alignments to reconstruct phylogenetic trees with IQ-tree \citep{nguyen_iq-tree:_2015} using a GTR model with ten rate categories.
+In the spirit of iterative updates, we reconstructed the ancestral states of the inferred tree using the simplest evolutionary model (Jukes-Cantor model) and calculated $n^b_{ij}$ and $\langle T_i^b\rangle$ from this reconstruction.
 
@@
 
-\subsection*{Non-linear update rules}
+Fig.~\ref{fig:reconstructed}A shows model inference accuracy for reconstructed trees and ancestral sequence inferred by TreeTime using a Jukes-Cantor model (line labeled 'reconstructed').
+The estimates of $\eqp_i^a$ are about as accurate as those inferred from the true ancestral states until the root-to-tip distance reaches about 0.2 but then start to deteriorate.
+Eventually, the estimates are even worse than the naive estimates from the alignment.
+This accuracy can be improved moderately by summing over uncertainty of the ancestral reconstruction (line labeled `ancestral sum').
+
+A more substantial improvment in accuracy can be obtained when the inferred model is used to re-infer the ancestral states, followed by repeated model inference.
+With this iterative procedure, model accuracy is comparable to that with known ancestral states up to root-to-tip distance of about 1.0.
+At longer still longer root to tip distances, extensize evolution results in inaccurate tree reconstruction.
+If, instead of the reconstructed tree, we supply the true simulated tree to the reconstruction algorithm, accuracy continues to improve even at root-to-tip distances greater than 1.0, see Fig.~\ref{fig:reconstructed}.
+
+In absense of strong regularization, maximum likelihood estimates of site specific equilibrium probabilities tend to `overfit' the data when the total number of observed mutations is low.
+This is reflected in Fig.~\ref{fig:reconstructed}B showing the different in entropy $-\sum_i \eqp_i^a \log \eqp_i^a$ of the inferred and true distributions.
+When the regularizing pseudo-count is low ($C=0.1$), the inferred distributions have too little entropy, meaning they are too concentrated on the few states that are observed.
+This effect, however, is much less pronounced in the tree based inference compared to naive estimates directly from the alignment.
+As the pseudo-count increases, this bias goes away, but in practice the necessary degree of regularization (i.e. the appropriate prior) is unknown.
+At very large root-to-tip distances, the model inference based on reconstructions results in slightly broader $\eqp_i^a$ with higher entropy.
+
+\begin{figure}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{average_branch_length}
+	\includegraphics[width=0.48\textwidth]{root_to_tip_distance}
+	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
+	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
+Panel B shows the results of the same optimization for the average root-to-tip distance.}
+	\label{fig:tree_length}
+\end{figure}
+
+\subsection*{Ignoring site specific frequencies dramatically underestimate branch length}
+As previously observed by various authors \citet{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in branch length underestimation if uniform equilibrium frequencies are assumed.
+This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^b}^2$ is increasing sharpely with the skew.
+Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
+
+Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the root-to-tip distance of trees reconstructed using IQ-tree (panel B) for models where $p_i^b$ are drawn from a uniform distribution on the $(q-1)$ simplex (Dirichlet parameter $\alpha=0$) and a model where all $\eqp_i^b=q^{-1}$.
+While branch length are accurately estimated for the latter, they are systematically underestimated for the former.
+This effect is particularly severe for the root-to-tip distance which is dominated by long branches deep in the phylogeny.
+Here, the inferred root-to-tip distance is essentially uncoupled from the true value for distance greater than 1.
+This effect is entirely due to skewed equilbrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^b=q^{-1}$.
+
+This problem is rectified to a large degree when branch length are optimized using the inferred model.
 
 \subsection*{Applications to large HIV alignments}
+
+\section*{Discussion}
+
+\begin{itemize}
+	\item underestimation of long branches, dubious deep time trees
+	\item relation to frequent reversions
+	\item fitness inference, functional constraints
+	\item caveat about interactions
+\end{itemize}
 

### 2019-01-28 src/calculate_branch_length.py

@@ -1,0 +1,77 @@
+import glob
+from matplotlib import pyplot as plt
+from Bio import Phylo, AlignIO
+from treetime import TreeAnc
+
+from generate_toy_data import *
+from filenames import *
+from estimation import *
+from reconstruct_toy_data import *
+
+
+def optimize_branch_length(tree, aln, model, n_iter=5, damping=0.5):
+    tt = TreeAnc(tree=tree, aln=aln, gtr = model, reduce_alignment=False)
+    for i in range(n_iter):
+        tt.infer_ancestral_sequences(marginal=True)
+        for n in tt.tree.find_clades():
+            if n.up is None:
+                continue
+            new_val = tt.optimal_marginal_branch_length(n, tol=1e-8 + 0.01**(i+1))
+            n.branch_length = new_val*(1-damping**(i+1)) + n.branch_length*damping**(i+1)
+
+    return tt
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument("-m", type=float, help="simulated mutation rate")
+    parser.add_argument("-n", type=int, help="number of taxa")
+    parser.add_argument("-L", type=int, help="length of sequence")
+    parser.add_argument("--aa",action='store_true', help="assume amino acid alphabet")
+    parser.add_argument("--prefix", type=str, required=True, help="folder to retrieve data from")
+    args=parser.parse_args()
+
+    prefix = args.prefix
+    mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
+    files = glob.glob(prefix+mask)
+
+    pc=0.1
+    n_iter = 5
+    alphabet = 'aa_nogap' if args.aa else 'nuc_nogap'
+    out_prefix = prefix+'_results_pc_%1.1f'%pc
+    for fname in files:
+        print(fname)
+        params = parse_alignment_name(fname)
+        true_tree = Phylo.read(tree_name(prefix, params), 'newick')
+        true_model = load_model(model_name(prefix, params))
+        m = true_model.average_rate().mean()
+        true_model.mu/=m
+        for n in true_tree.find_clades():
+            n.branch_length *= m
+        print(true_tree.total_branch_length()*m)
+
[…]

### 2019-01-28 src/compare_tree_length.py

@@ -16,8 +16,9 @@ if __name__ == '__main__':
     args=parser.parse_args()
 
     prefix = args.prefix
-    mask = "/L{L}_n{n}_m{mu}_*fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
-    files = glob.glob(prefix+mask)
+    result_prefix = prefix+'_results_pc_0.1'
+    mask = "/L{L}_n{n}_m{mu}_*reoptimized_true_model.nwk".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
+    files = glob.glob(result_prefix+mask)
 
     length = []
     depth = []
@@ -26,46 +27,39 @@ if __name__ == '__main__':
         params = parse_alignment_name(fname)
         true_tree = Phylo.read(tree_name(prefix, params), 'newick')
         true_model = load_model(model_name(prefix, params))
-        aln = AlignIO.read(alignment_name(prefix, params), 'fasta')
-        iq_tree = Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
-
-        iq_tree_ttl = 0.5*iq_tree.total_branch_length()/params['n']
-        iq_tree_depth = np.mean([x for c,x in iq_tree.depths().items() if c.is_terminal()])
-
         m = true_model.average_rate().mean()
-        true_model.mu/=m
-        tt = TreeAnc(tree=iq_tree, aln=aln, gtr = true_model, reduce_alignment=False)
-        tt.infer_ancestral_sequences(marginal=True)
-        for i in range(5):
-            x = []
-            for n in tt.tree.find_clades():
-                if n.up is None:
-                    continue
-                x.append([tt.optimal_marginal_branch_length(n), n.branch_length])
-                n.branch_length = 0.5*(x[-1][0] + n.branch_length)
+        for n in true_tree.find_clades():
+            n.branch_length *= m
+        
+        iq_tree = Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
+        reoptimized_T = Phylo.read(reoptimized_tree(result_prefix, params), 'newick')
+        reoptimized_T_true_model = Phylo.read(reoptimized_tree_true_model(result_prefix, params), 'newick')
 
-            x = np.array(x)
-            print(x.sum(axis=0))
-            tt.infer_ancestral_sequences(marginal=True)
-            print(tt.sequence_LH())
[…]

### 2019-01-28 src/estimation.py

@@ -28,7 +28,11 @@ def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap',
         with gzip.open(alignment_name(in_prefix, params), 'rt') as fh:
             aln = AlignIO.read(fh, 'fasta')
         tree_fname =  reconstructed_tree_name(in_prefix, params) if reconstructed_tree else tree_name(in_prefix, params)
-        myTree = TreeAnc(gtr=gtr, alphabet=alphabet,
+        if type(gtr)==str:
+            tmp_GTR = GTR.standard(gtr, alphabet=alphabet)
+        else:
+            tmp_GTR=gtr
+        myTree = TreeAnc(gtr=tmp_GTR, alphabet=alphabet,
                          tree=tree_fname, aln=aln,
                          reduce_alignment=False, verbose = 0)
 
@@ -41,17 +45,17 @@ def reconstruct_counts(in_prefix, params, gtr='JC69', alphabet='nuc_nogap',
     return mutation_counts, myTree.sequence_LH(), myTree
 
 
-def estimate_GTR(mutation_counts, pc=0.1, single_site=False, tt=None):
+def estimate_GTR(mutation_counts, pc=0.1, single_site=False, tt=None, alphabet='nuc_nogap'):
     n_ija, T_ia, root_sequence = mutation_counts[:3]
-    root_prof = seq2prof(root_sequence, profile_maps['nuc_nogap']).T
+    root_prof = seq2prof(root_sequence, profile_maps[alphabet]).T
 
 
     if single_site:
         inferred_model = GTR.infer(n_ija.sum(axis=-1), T_ia.sum(axis=-1), pc=pc,
-                          root_state=root_prof.sum(axis=-1), alphabet='nuc_nogap')
+                                   root_state=root_prof.sum(axis=-1), alphabet=alphabet)
     else:
         inferred_model = GTR_site_specific.infer(n_ija, T_ia, pc=pc,
-                          root_state=root_prof, alphabet='nuc_nogap')
+                                                 root_state=root_prof, alphabet=alphabet)
 
     if tt:
         (m_denom, m_num), (t_denom, t_num) = exact_mu_t_update(tt)

### 2019-01-28 src/filenames.py

@@ -35,3 +35,9 @@ def tree_name(prefix, params):
 
 def reconstructed_tree_name(prefix, params):
     return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reconstructed.nwk'.format(**params)
+
+def reoptimized_tree(prefix, params):
+    return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reoptimized.nwk'.format(**params)
+
+def reoptimized_tree_true_model(prefix, params):
+    return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reoptimized_true_model.nwk'.format(**params)

### 2019-01-28 src/generate_toy_data.py

@@ -32,7 +32,7 @@ def load_model(fname):
     return GTR_site_specific.custom(alphabet=d['alphabet'], mu=d['mu'], pi=d['pi'], W=d['W'])
 
 
-def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False):
+def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False, alphabet='nuc_nogap'):
     from Bio import AlignIO
     # generate a model
     T = betatree(params['n'], alpha=2.0)
@@ -44,10 +44,10 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=Fa
     for mi in range(n_model):
         params['model']=mi
         if JC:
-            myGTR = GTR_site_specific.random(L=params['L'], alphabet='nuc_nogap',
+            myGTR = GTR_site_specific.random(L=params['L'], alphabet=alphabet,
                                              pi_dirichlet_alpha=0, W_dirichlet_alpha=0)
         else:
-            myGTR = GTR_site_specific.random(L=params['L'], alphabet='nuc_nogap')
+            myGTR = GTR_site_specific.random(L=params['L'], alphabet=alphabet)
         myGTR.mu*=params['m']
 
         if out_prefix:
@@ -63,17 +63,17 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=Fa
                 save_mutation_count(mySeq, mutation_count_name(out_prefix, params))
                 with open(alignment_name_raw(out_prefix, params), 'wt') as fh:
                     AlignIO.write(mySeq.get_aln(), fh, 'fasta')
-                reconstruct_tree(out_prefix, params)
+                reconstruct_tree(out_prefix, params, aa='aa' in alphabet)
 
 
-def reconstruct_tree(prefix, params):
+def reconstruct_tree(prefix, params, aa=False):
     aln_file = alignment_name_raw(prefix, params)
     fast_opts = [
         "-ninit", "2",
         "-n",     "2",
         "-me",    "0.05"
     ]
-    call = ["iqtree"] + fast_opts +["-nt 1", "-s", aln_file, "-m", 'GTR+G10',
+    call = ["iqtree"] + fast_opts +["-nt 1", "-s", aln_file, "-m", 'LG' if aa else 'GTR+G10',
             ">", "iqtree.log"]
     os.system(" ".join(call))
[…]

### 2019-01-28 src/plot_toy_data.py

@@ -3,9 +3,10 @@ import numpy as np
 from collections import defaultdict
 
 fmts = ['.png', '.pdf']
+fs = 12
 
-labels = {'naive':"Alignment frequencies", "dressed":"Inference from true substitution counts", "regular":"Reconstructed",
-          "marginal":"ancestral sum", "iterative":"Iterated", 'branch_length':"non-linear",
+labels = {'naive':"Alignment frequencies", "dressed":"True substitution counts", "regular":"Reconstructed",
+          "marginal":"ancestral sum", "iterative":"Iterative, reconstructed tree", 'branch_length':"non-linear",
           "iterative_true":"iterative, true tree", 'marginal_true':"ancestral sum, true tree"}
 colors = {k:'C%d'%(i+1) for i,k in enumerate(sorted(labels.keys()))}
 
@@ -61,9 +62,11 @@ def plot_pdist_vs_tree_length(data, n_vals, mu_vals, methods=None, fname=None):
     plt.plot([10,1000], [0.1,0.001], label=r'$\sim x^{-1}$', c='k')
     plt.yscale('log')
     plt.xscale('log')
-    plt.legend()
-    plt.xlabel('average number of substitutions per site')
-    plt.ylabel('squared deviation of $p_i^a$')
+    plt.legend(fontsize=fs)
+    plt.tick_params(labelsize=0.8*fs)
+    plt.xlabel('average number of substitutions per site', fontsize=fs)
+    plt.ylabel('squared deviation of $p_i^a$', fontsize=fs)
+    plt.tight_layout()
 
     if fname:
         for fmt in fmts:
@@ -88,33 +91,39 @@ def plot_pdist_vs_rtt(data, n_vals, mu_vals, methods=None, fname=None):
 
     plt.yscale('log')
     plt.xscale('log')
-    plt.legend()
-    plt.xlabel('root-to-tip distance')
-    plt.ylabel('squared deviation')
+    plt.legend(fontsize=fs)
+    plt.tick_params(labelsize=0.8*fs)
+    plt.xlabel('root-to-tip distance', fontsize=fs)
+    plt.ylabel('squared deviation', fontsize=fs)
+    plt.tight_layout()
 
     if fname:
         for fmt in fmts:
             plt.savefig(fname+fmt)
 
 
-def plot_pentropy_vs_rtt(data, n_vals, mu_vals, methods=None, fname=None):
+def plot_pentropy_vs_rtt(data, n_vals, mu_vals, pc_vals, methods=None, fname=None):
[…]

### 2019-01-28 src/reconstruct_toy_data_profile.py

@@ -10,6 +10,7 @@ from treetime.seq_utils import seq2prof, profile_maps, alphabets
 from generate_toy_data import *
 from filenames import *
 from estimation import *
+import cProfile, pstats, io
 
 def KL(p,q):
     return np.sum(p*log(p/q))
@@ -47,6 +48,8 @@ if __name__ == '__main__':
     analysis_types = ['naive', 'single', 'dressed', 'regular', 'regular_true',
                       'marginal','marginal_true', 'iterative', 'iterative_true']
 
+    analysis_types = ['iterative']
+
     alphabet='aa_nogap' if args.aa else 'nuc_nogap'
     for fname in files:
         print(fname)
@@ -75,12 +78,21 @@ if __name__ == '__main__':
                     bl = [n.branch_length for n in _t.tree.find_clades() if n!=_t]
                 elif ana in ['iterative', 'iterative_true']:
                     model = 'JC69'
+                    pr = cProfile.Profile()
+                    pr.enable()
                     for i in range(niter):
                         mc = reconstruct_counts(prefix, params, gtr=model,
                                                 alphabet=alphabet, marginal=True,
                                                 reconstructed_tree=ana=='iterative')
                         model = estimate_GTR(mc[0], pc=pc, single_site=False, alphabet=alphabet)
                         print(i, np.mean([chisq(model.Pi[:,i],true_model.Pi[:,i]) for i in range(params['L'])]))
+                    pr.disable()
+
+                    s = io.StringIO()
+
+                    ps = pstats.Stats(pr, stream=s).sort_stats('cumtime')
+                    ps.print_stats()
+                    print(s.getvalue())
                 elif ana=='branch_length':
                     model = 'JC69'
                     kappa=1.0

### 2019-07-27 manuscript/ms.tex
added methods, fleshed out HIV fitness description

--- 2019-01-02 manuscript/ms.tex
@@
-\documentclass[aps,rmp, onecolumn]{revtex4}
+\documentclass[aps,rmp,twocolumn]{revtex4}
 %\documentclass[a4paper,10pt]{scrartcl}
@@
 
+\section*{Introduction}
+Over time, genome sequences change through mutations and are reshuffeled by recombination.
+Modifications to the genomes are filtered by selection for survival such that beneficial variants spread preferentially and those that impair function are purged.
+As a result, some parts of genomes change rapidly, while other are strongly conserved.
+In addition to variation of the evolutionary rate, different sites in a genome only explore different subsets of the available states.
+Some positions in a protein, for example, might only allow for hydrophobic amino acids, while others require acidic side chains.
+Patterns of conservation and variation, possibly involving more than one site, are therefore shaped by functional contraints which in turn allows inference of biological function from genetic variation.
+
+Similarly, phylogenetics aims at reconstructing the relationships and history of homologous sequences from the substitutions that occurred in the past.
+Modern phylogenetic methods describe this stochastic evolutionary process with probabilistic models of sequence evolution and aim to find phylogenies that either maximize the likelihood of observing the alignment or sample phylogenies from a posterior probability distribution \citep{felsenstein2004inferring}.
+
+Inferring phylogenies is a computationally challenging problem since the number of phylogenies grows super-exponentially with the number of taxa and because the calculation of the likelihood is compulationally costly (though linear in the number of taxa).
+Due to this computational complexity, the most commonly used substitution models are simple caricatures of biological complexity.
+The simplest substitution models assume that all sites and sequence states are equivalent and evolve at the same rate, i.e., they assume an unconstrained non-functional sequence that mutates at random between the different sequence states.
+Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result, among other issues, in biased estimates of divergence times \citep{yang1996among}.
+Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
+To avoid over-fitting, these methods typically don't estimate rates for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple uni-model distributions \citep{mayrose2005gamma}, or a small number of fixed rates).
+
+In addition to rate variation, different sites in a protein differ in amino acid they allow.
+Recent deep mutational scanning experimentes have shown that site specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
+Using such experimentally inferred site specific models in phylogenetic inference greatly improves the likelihood of the data \cite{bloom2014experimentally}.
+More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
+\citet{hilton_modeling_2018} recently showed that experimentally measured preference not only improve the phylogenetic fit, but also results in longer branch length estimates.
+
+On the otherhand, estimating such models, also known as mutation-selection balance models \citep{bruno1996modeling,yang2008mutation}, from data exacerbates the over-fitting problem.
+In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
+With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
+Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on model deviations.
+
+\begin{figure*}[htb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha3.0}.pdf}
+	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha3.0}.pdf}
+	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree.
+(A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences.
+	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
+	(C) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+Regularization via pseudo-counts reduces over-fitting at low divergence.
+}
+	\label{fig:dressed}
+\end{figure*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Model
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Efficient inference of site specific substitution models}
+Following previous authors \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
+the overall rates and the equilibrium frequencies depend on the site.
+Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
+\begin{eqnarray}
+Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
+Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
+\label{eq:Qij}
+\end{eqnarray}
+where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
+In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.
+
+Extending the iterative expectation maximization approach by \citet{bruno1996modeling}, we show in the appendix that the iterative update rules
+\begin{equation}
+\label{eq:update}
+	\begin{split}
+		\mu^b & \leftarrow \mu^b\frac{C+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
+		\eqp^b_i & \leftarrow \eqp^b_i\frac{C+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(qC+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
+		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
+	\end{split}
+\end{equation}
+approximately maximize the likelihood of the data.
+Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
+$C$ is a pseudocount analogous to a Dirichlet prior that  in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $C$ over the total tree length.
+To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
+This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% ACCURACY
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Accuracy of inferences}
+The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and the accuracy of the tree reconstruction.
+In addition, the most likely model parameters will deviate from the true parameters due to over-fitting.
+
+To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites.
+We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha3.0}.pdf}
+	\caption{Errors stemming from tree and ancestral reconstruction: At large root-to-tip distances, ancestral reconstruction becomes more and more uncertain and estimation of equilibrium state frequencies fails.
+	These errors are gradually eliminated by first summing over ancestral uncertainty, iterative redoing ancestral reconstruction using the inferred model, and re-optimizating branch length using the updated models (or using the true tree).
+	The right panel shows analysis of simulated with four states (nucleotides), while the right panels shows the corresponding graph of 20 states (amino acids).
+As expected, saturation and non-linear effects set in later with larger alphabet.}
+	\label{fig:reconstructed}
+\end{figure*}
+
+Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site for estimates of $\eqp_i^a$ from the alignment and using the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
+The plot is shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope 1 implies that the accuracy is inversely proportional to the expected number of substitutions.
+This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
+At the largest substitution rates, branch length are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
+Nevertheless, the acccuracy of the $\hat{\eqp}_i^a$ continues to increase.
+By contrast, using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry.
+
+The accuracy of substitution rates, however, is affected by linearization and we consistently underestimate $\mu^b$.
+This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
+The relative subsitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.
+
+%%% effect of reconstruction
+In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized)
+To test the influence of tree and ancestral state reconstruction, we built phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{Price} using the default 20 category model for simulated nucleotide and amino-acid sequences, respectively.
+Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
+The simplest approach is to take the inferred tree as given, reconstruct the ancestral states a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
+This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate.
+Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvemet in accurary.
+More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site specific model.
+The accuracy now continues to improve up to root-to-tip distances of about one.
+At this level of divergence, tree reconstruction starts becoming problematic and branch length deviate substantially from their true values.
+Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence.
+Similar improvements are achieved by optimizing tree branch length along with the model.
+
+\subsection*{Ignoring site specific frequencies dramatically underestimates branch lengths}
+As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in branch length underestimation if uniform equilibrium frequencies are assumed.
+This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^b}^2$ is increasing sharply with the skew.
+Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_length}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_depth}.pdf}
+	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
+	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
+Panel B shows the results of the same optimization for the average root-to-tip distance.}
+	\label{fig:tree_length}
+\end{figure*}
+
+%% dependence on pseudo count
+Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted tree (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
+While branch length are accurately estimated for when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
+Underestimation is particularly severe for the root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
+The inferred root-to-tip distance is essentially independent from the true value for distance greater than 1.
+This effect is entirely due to skewed equilbrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^b=q^{-1}$.
+
+%% dependence on nucleotide and amino acid skew
+The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed.
+
+\begin{figure}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_2019-06-01_simulatedData_L1000_ratealpha3.0_alpha1.0_nuc_n100_m0.35}.pdf}
+	\caption{{\bf Moderate model inaccuracies result in large substantial branch length underestimation}.
+	}
+	\label{fig:model_deviation}
+\end{figure}
+
+\subsection*{Minor model deviations can have substantial effects on divergence estimates}
+In the previous section, we have shown that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
+To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$.
+
+Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
+Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order 0.3.
+If the same preference are used for every site, however, deviations are substantial.
+Deviations in rate and preferences are approximatly additive.
+Fig.~\ref{fig:model_deviation}B shows the degree of mis-estimation when using flat preferences, flat rates, or both as a function of the degree of divergence.
+Initially, errors accumulate linearly with increasing.
+
+\subsection*{Applications to large HIV alignments}
+Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
+For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
+This large sample of moderately diverged sequences is a unique opportunity to apply our iterative inference frame work.
+We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level.
+
+One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
+Simple population genetic models predict that the fixation probability of state $i$ if starting from state $j$ depends on the fitness different $s_{ij}$ and the effective population size $N$ as
+\begin{equation}
+	f_{ij} = \frac{1-e^{-2s_{ij}}}{1-e^{-2Ns_{ij}}} \approx
+	\begin{cases}
+		2s_{ij} & s_{ij}>0 \\
+		2s_{ij} e^{2Ns_{ij}} & s_{ij} < 0
+	\end{cases}
+\end{equation}
+On longer time scales, the fixation rates $f_{ij}$ should correspond to transitions rates of the site specific GTR model $\muQ_{ij}$.
+The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations.
+Nevertheless, the logarithm of the ratio $\log f_{ij}/f_{jj} \approx 2Ns_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
+We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $a$
+\begin{equation}
+	e^{2N s_i^a} = \frac{\Gamma_{in}}{\Gamma_{out}} = \frac{\sum_j \pi_i^a W_{ij}}{\sum_j W_{ij}\pi_j^a}
+\end{equation}
+The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
+We compare this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
+Fig.~\ref{fig:HIV_fitness} shows a linear relation ship between the logarithmic fitness effects measured with-in host and from the phylogeny that explains about half the variance.
+The linear relationship holds both for nucleotide states and amino acid states.
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.001}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.001_aa}.pdf}
+	\caption{{\bf Intra-host vs cross-sectional mutation selection balance are consistent. } }
+	\label{fig:HIV_fitness}
+\end{figure*}
+
+\section*{Discussion}
+Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize and a common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
+Different positions in a DNA sequence or protein, however, also differ in the preferences for different states.
+While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern1998evolutionary}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
+However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data sets are increasing rapidly such that more complex models can be estimated.
+Large data sets, however, require efficient and scalable methods.
+
+Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-stype approach inspired by the original paper by \citet{bruno1996modeling}.
+We show that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
+For short branch length a parsimonious ancestral reconstruction is sufficient, while for more diverged samples and iterative inference of the model, the ancestral states, and the branch lengths is necessary.
+The computational complexity of the inference is dominated by ancestral reconstruction and branch length optimization, which is quadratic in the alphabet size and linear in sample size and sequence length.
+
+We further explore how branch length estimation depends on model choice and inference accuracy.
+As has been reported before, branch length are underestimated with variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
+Even moderate model deviation can result in substantial underestimates.
+This mirrors observations by \citet{hilton_modeling_2018}, who found that preferences measured for influenza HA proteins of type H1 or H3 affect branch length of in the vicinity of the focal sequence, but not globally on the tree.
+
+The potentially large error in branch length estimating can seriously affect deep divergence time estimates.
+Typically, short branches close to the tips of the tree are used to calibrate molecular clock models.
+The deep branches, however, tend to be much longer and rate estimation and variation of site specific preferences will result in saturation effects not accoutned for by the model \citep{hilton_modeling_2018}.
+This effective variation in rate is related to the effect of transient deletrious mutations that inflate the rate of short time scales.
+
+\begin{itemize}
+	\item underestimation of long branches, dubious deep time trees
+	\item relation to frequent reversions
+	\item fitness inference, functional constraints
+	\item caveat about interactions
+\end{itemize}
+
+\bibliography{GTRbib}
+
+\appendix
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\onecolumngrid
+
+\section*{Model and Notation}
 Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
@@
 
+% To regularize the parameter estimates when the data are not sufficiently informative to constrain them, we use Dirichlet and Gamma priors for $\eqp_i^a$ and $\mu^a$.
+% \begin{equation}
+% 	\log P(p_i^a) = C + (\alpha-1) \log \eqp_i^a \quad \mathrm{and} \quad \log P(\mu^a) = C -\frac{\beta \mu^a}{\bar{\mu}} + (\beta - 1)\log \mu^a
+% \end{equation}
+
+\section*{Iterative optimization algorithm}
 The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
@@
 		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} + 	\sum_{k\in T, k^b_c\neq k^b_p} \frac{1}{\mu^b} \\
-		&= -\sum_{ij}\eqp_{i}W_{i,j}\langle T_j^b\rangle + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
+		&= -\sum_{ij}\eqp_{i}W_{i,j}\tau_j^b + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
 	\end{split}
 \end{equation}
-where $\langle T^b_j\rangle$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
-This condition (and the corresponding ones in the supplement suggest solution at fixed $\langle T^b_j\rangle$ and $n^b_{ij}$ using the following iterative update:
+where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
+This condition (and the corresponding ones in the supplement suggest solution at fixed $\tau^b_j$ and $n^b_{ij}$ using the following iterative update:
 \begin{equation}
@@
 	\begin{split}
-		\mu^b & \leftarrow \mu^b\frac{\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\langle T_j^b\rangle} \\
-		\eqp^b_i & \leftarrow \eqp^b_i\frac{\sum_{j\neq i} n^b_{ij}+\delta_{is^{b}}}{\mu^b\sum_{j}\eqp^b_i W_{ij}\langle T_j^b\rangle + \eqp^b_i} \\
-		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\langle T_j^b\rangle + \eqp_j^b \langle T_i^b\rangle)} \\
+		\mu^b & \leftarrow \mu^b\frac{C+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
+		\eqp^b_i & \leftarrow \eqp^b_i\frac{C+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(qC+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
+		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
 	\end{split}
 \end{equation}
+where $C$ is a pseudocount analogous to a Dirichlet prior that in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $C$ over the total tree length.
 To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
-This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\langle T_j^b\rangle$ -- each appropriately summed over sites or states.
+This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
 
-The validity of this iterative solution will depend on the accuracy of the linear approximation made and the degree to which the ancestral states can be reconstructed or summed over.
-To assess these two sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\langle T_j^b\rangle$ for a range of sample sizes, tree shapes, and branch lengths.
-Fig.~\ref{fig:dressed} shows the accuracy of the inference as quantified by $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \hat{\eqp}_i^a)^2$ where $\hat{\eqp}_i^a$ and $\hat{\eqp}_i^a$ the inferred and true equilibrium probabilities.
+\section*{HIV sequence analysis}
+HIV-1 sequences were downloaded from LANL HIV database \citep{LANL} setting filters to ``one sequence per patient'', ``non-ACGT\textless 0.3''.
+Separate downloads where made for sequences the following sets: \gene{pol}, subtype B (2019-05-05); \gene{pol}, subtype C (2019-05-13); \gene{gag}, \gene{nef}, subtype B (2019-06-07).
 
-\begin{figure}[tb]
+Sequences were aligned to the HXB2 reference sequences using mafft \citep{katoh2013mafft}, phylogenies were inferred using IQ-tree \citep{nguyen_iq-tree:_2015}, and ancestral sequences were inferred using TreeTime \citep{sagulenko2017treetime} via the nextstrain's augur pipeline \citep{hadfield_nextstrain:_2018}.
+The different steps were assembled into a pipeline using the workflow manager Snakemake \citep{koster_snakemakescalable_2012}.
+
+The sequences and trees were then used for site specific GTR inference as implemented in TreeTime \citep{sagulenko2017treetime}.
+The scripts detailing this analysis and producing the figures are available on GitHub in repository \href{https://github.com/neherlab/SiteSpecificGTR}{neherlab/SiteSpecificGTR}.
+
+\clearpage
+
+\section*{Supplementary figures}
+\begin{figure*}[tb]
 	\centering
-	\includegraphics[width=0.48\textwidth]{../figures/p_dist_vs_treelength.pdf}
-	\includegraphics[width=0.48\textwidth]{../figures/avg_rate_dressed.pdf}
-	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree.
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha3.0}.pdf}
+	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha3.0}.pdf}
+	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree, for amino acid alphabets.
 (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
 	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences.
-	(B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected. }
+	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
+	(C) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+Regularization via pseudo-counts reduces over-fitting at low divergence.
+}
 	\label{fig:dressed}
-\end{figure}
-
-Outside simulated sequences, the tree and the ancestral states at internal nodes are unknown and need to be reconstructed and summed over.
-In the spirit of iterative updates, we first reconstruct those states using a simple evolutionary model such as Jukes-Cantor model, use the resulting ancestral reconstruction to estimate a site-specific model using Eq.~\ref{eq:update}, which can then be used to estimate an improved reconstruction.
-Fig.~\ref{fig:reconstructed} shows model inference accuracy for reconstructed trees and sequences and compares them to the accuracy that would have been obtained if internal states had been known and the naive alignment frequencies.
-Model inference from reconstructed data is accurate if the root to tip distance is small but rapidly looses accuracy if sequences are very diverged.
-
-\begin{figure}[tb]
-	\centering
-	\includegraphics[width=0.48\textwidth]{../figures/p_dist_vs_rtt.pdf}
-	\caption{At large root-to-tip distances, ancestral reconstruction becomes less and less accurate and estimation of equilibrium state frequencies fails. }
-	\label{fig:reconstructed}
-\end{figure}
-
-\subsection*{Non-linear update rules}
-
-\subsection*{Applications to large HIV alignments}
-
-\bibliography{GTRbib}
-
+\end{figure*}
 \end{document}

### 2019-07-27 src/analyze_HIV_tree.py

@@ -24,7 +24,11 @@ def convert_vals(x):
 
 def load_fitness_landscape(gene, aa=False, subtype='B'):
     import pandas as pd
-    d = pd.read_csv('HIV_fitness_landscape/data/fitness_pooled/nuc_%s_selection_coeffcients_%s.tsv'%(gene, subtype),
+    if aa:
+        d = pd.read_csv('HIV_fitness_landscape/data/fitness_pooled_aa/aa_%s_fitness_costs_uncensored_st_%s.tsv'%(gene, subtype),
+                    sep='\t', skiprows=1)
+    else:
+        d = pd.read_csv('HIV_fitness_landscape/data/fitness_pooled/nuc_%s_selection_coeffcients_uncensored_%s.tsv'%(gene, subtype),
                     sep='\t', skiprows=1)
     for col in d.columns[3:6]:
         d.loc[:,col] = d.loc[:,col].apply(convert_vals)
@@ -39,11 +43,12 @@ def calculate_in_out_ratio(gtr, major_allele):
     '''
     major_index = np.zeros_like(major_allele, dtype=int)
     for ni, n in enumerate(gtr.alphabet):
-        major_index[major_allele==n]=ni
+        major_index[major_allele==n.decode()]=ni
+    #major_index = np.argmax(gtr.Pi, axis=0)
 
     r_ind = np.arange(gtr.Pi.shape[-1])
-    return np.einsum('a,aj->a', gtr.Pi[major_index, r_ind], gtr.W[major_index])/\
-           np.einsum('aj,ja->a', gtr.W[major_index], gtr.Pi)
+    return np.einsum('a,a,aj->a', gtr.mu, gtr.Pi[major_index, r_ind], gtr.W[major_index])/\
+           np.einsum('a,aj,ja->a', gtr.mu, gtr.W[major_index], gtr.Pi)
 
 
 def diversity(p):
@@ -61,13 +66,14 @@ if __name__ == '__main__':
     parser.add_argument("--redo", action='store_true', default=False, help="re-estimate")
     args=parser.parse_args()
 
-    alphabet='aa' if args.aa else 'nuc'
+    alphabet='aa' if args.aa else 'nuc_nogap'
     pc = args.pc
+    fs = 16
 
-    aln = args.prefix +  '_aligned.fasta'
[…]

### 2019-07-27 src/calculate_branch_length.py

@@ -10,10 +10,10 @@ from reconstruct_toy_data import *
 
 
 
-def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False):
+def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False, pc=0.1):
     params = parse_alignment_name(fname)
     m = params['m']
-    tree = Phylo.read(tree_name(prefix, params), 'newick') if args.true_tree else  Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
+    tree = Phylo.read(tree_name(prefix, params), 'newick') if true_tree else  Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
     tree.root.branch_length = 0.001
     tree.ladderize()
     old_bl = []
@@ -21,7 +21,7 @@ def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False):
     print(tree.root.clades[0].branch_length/tree.root.clades[1].branch_length)
     for n in tree.find_clades():
         old_bl.append(n.branch_length)
-        if args.true_tree:
+        if true_tree:
             n.branch_length *= m*(0.6+0.4*np.random.random())
     print(np.sum(old_bl)*(m if true_tree else 1.0),m)
     if true_model:
@@ -37,16 +37,17 @@ def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False):
                  alphabet=alphabet, verbose=3)
 
     tt.optimize_tree(branch_length_mode='marginal', max_iter=n_iter,
-                     infer_gtr=not args.true_model, site_specific_gtr=True, pc=args.pc)
+                     infer_gtr=not true_model, site_specific_gtr=True, pc=pc, damping=0.75)
+
+    new_bl = []
+    for n in tt.tree.find_clades():
+        new_bl.append(n.branch_length)
+
 
     print(tt.tree.total_branch_length(),tt.gtr.average_rate().mean())
     # tt.tree.root_at_midpoint()
     tfname = reoptimized_tree_true_model(out_prefix, params) if args.true_model else reoptimized_tree(out_prefix, params)
     Phylo.write(tt.tree, tfname, 'newick')
-
-    new_bl = []
-    for n in tt.tree.find_clades():
-        new_bl.append(n.branch_length)
 
     print(np.mean([x for c,x in tt.tree.depths().items() if c.is_terminal()]), tt.tree.total_branch_length())
[…]

### 2019-07-27 src/compare_BandC.py

@@ -1,0 +1,47 @@
+import os, gzip, glob, pickle
+from Bio import AlignIO
+import numpy as np
+from scipy.stats import pearsonr, spearmanr
+from collections import defaultdict
+import matplotlib.pyplot as plt
+
+from treetime.treeanc import TreeAnc
+from treetime.gtr_site_specific import GTR_site_specific
+from treetime.gtr import GTR
+from treetime.seq_utils import seq2prof, profile_maps, alphabets
+
+from estimation import *
+from generate_toy_data import save_model, load_model
+from analyze_HIV_tree import *
+
+
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    hiv_out = 'HIV_inferences/'
+    parser.add_argument("--prefix", type=str, help="data set")
+    parser.add_argument("--gene", type=str, default='pol')
+    parser.add_argument("--aa", action='store_true', help="use amino acid alphabet")
+    parser.add_argument("--pc", type=float, default=1.0, help="pseudocount for reconstruction")
+    args=parser.parse_args()
+
+    alphabet='aa' if args.aa else 'nuc_nogap'
+    pc = args.pc
+
+    aln = args.prefix +  '_aligned.fasta'
+    tree = args.prefix + '_tree.nwk'
+    model_name = hiv_out + os.path.basename(aln)[:-5]+'%1.2f_inferred_model.npz'%args.pc
+    aln_freq_name = hiv_out + os.path.basename(aln[:-5])+'alignment_frequencies'
+
+    aln_freq = {}
+    gtr = {}
+    fitness = {}
+    for st in ['B', 'C']:
+        aln_freq[st] = np.loadtxt(aln_freq_name.replace('#st#', st))
+        gtr[st] = load_model(model_name.replace('#st#', st))
+        fitness[st] = calculate_in_out_ratio(gtr[st], fabio_fitness.loc[:,"consensus"])
+
+    overlap_gtr = np.sum(gtr['B'].Pi*gtr['C'].Pi, axis=0).mean()
+    overlap_afreq = np.sum(aln_freq['B']*aln_freq['C'], axis=0).mean()
+

### 2019-07-27 src/compare_tree_length.py

@@ -11,63 +11,96 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
     parser.add_argument("-m", type=float, help="simulated mutation rate")
     parser.add_argument("-n", type=int, help="number of taxa")
-    parser.add_argument("--pc", type=float, help="pseudo count")
+    parser.add_argument("--pc", nargs='+', default=[0.1],type=float, help="pseudo count")
     parser.add_argument("-L", type=int, help="length of sequence")
     parser.add_argument("--prefix", type=str, required=True, help="folder to retrieve data from")
     args=parser.parse_args()
 
     prefix = args.prefix
-    result_prefix = prefix+'_results_pc_%1.2f'%args.pc
-    mask = "/L{L}_n{n}_m{mu}_*reoptimized.nwk".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
-    files = glob.glob(result_prefix+mask)
+    aa ='_aa' in args.prefix
 
-    length = []
-    depth = []
+    length = {}
+    depth = {}
+    terminal_bl = {}
+
+    for dset in ['true', 'ML']+args.pc:
+        length[dset]=[]
+        depth[dset]=[]
+        terminal_bl[dset]=[]
+
+    mask = "/L{L}_n{n}_m{mu}_*.fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
+    files = glob.glob(prefix+mask)
     for fname in files:
+        print(fname)
         params = parse_alignment_name(fname)
         true_tree = Phylo.read(tree_name(prefix, params), 'newick')
-        true_model = load_model(model_name(prefix, params))
-        cf = true_model.average_rate().mean()/params['m']
+        #true_model = load_model(model_name(prefix, params))
         for n in true_tree.find_clades():
             n.branch_length *= params['m']
+        true_ttl = 0.5*true_tree.total_branch_length()/params['n']
+        true_depth = np.mean([x for c,x in true_tree.depths().items() if c.is_terminal()])
+        true_terminal = np.mean([x.branch_length for x in true_tree.get_terminals()])
+        for dset in ['true', 'ML']+args.pc:
+            result_prefix = args.prefix+'_results'
+            if type(dset)==float:
[…]

### 2019-07-27 src/generate_toy_data.py

@@ -34,7 +34,7 @@ def load_model(fname, flatten_p=0.0, flatten_mu=0.0, flatten_W=0.0):
                                     pi=d['pi']*(1-flatten_p)+flatten_p/d['pi'].shape[0],
                                     W=d['W']*(1-flatten_W) + 2*flatten_W/d['W'].shape[0]/(d['W'].shape[0]-1))
 
-def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False, alphabet='nuc_nogap', alpha=1.0):
+def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=False, alphabet='nuc_nogap', alpha=1.0, rate_alpha=1.5):
     from Bio import AlignIO
     # generate a model
     T = betatree(params['n'], alpha=2.0)
@@ -50,9 +50,9 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5, JC=Fa
         params['model']=mi
         if JC:
             myGTR = GTR_site_specific.random(L=params['L'], alphabet=alphabet,
-                                             pi_dirichlet_alpha=0, W_dirichlet_alpha=0)
+                                             pi_dirichlet_alpha=0, W_dirichlet_alpha=0, mu_gamma_alpha=rate_alpha)
         else:
-            myGTR = GTR_site_specific.random(L=params['L'], alphabet=alphabet, pi_dirichlet_alpha = alpha, mu_gamma_alpha=1.5)
+            myGTR = GTR_site_specific.random(L=params['L'], alphabet=alphabet, pi_dirichlet_alpha = alpha, mu_gamma_alpha=rate_alpha)
 
         myGTR.mu*=params['m']
 
@@ -131,6 +131,7 @@ if __name__ == '__main__':
     parser.add_argument("--JC", action='store_true', help="simulate JC model")
     parser.add_argument("--aa", action='store_true', help="use amino acid alphabet")
     parser.add_argument("--alpha", default=1.0, type=float,  help="parameter of the dirichlet distribution for preferences")
+    parser.add_argument("--rate-alpha", default=1.5, type=float,  help="parameter of the gamma distribution for rates")
     parser.add_argument("--prefix", type=str, help="folder to save data")
     args=parser.parse_args()
 
@@ -144,5 +145,5 @@ if __name__ == '__main__':
     mu = args.m
[…]

### 2019-07-27 src/model_deviation.py

@@ -24,7 +24,7 @@ def run_tree(fname, out_prefix, alphabet, model):
         aln = AlignIO.read(fh, 'fasta')
 
     tt = TreeAnc(tree=tree, aln=aln, gtr = model, compress=False,
-                 alphabet=alphabet, verbose=1)
+                 alphabet=alphabet, verbose=3)
 
     tt.optimize_tree(branch_length_mode='marginal', max_iter=n_iter,
                      infer_gtr=False)
@@ -47,59 +47,124 @@ if __name__ == '__main__':
 
     n_iter = 20
     alphabet = 'aa_nogap' if args.aa else 'nuc_nogap'
-    out_prefix = prefix+'_results_model_dev'
+    out_prefix = prefix+'_results'
+    muvals =  [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]
+    out_file = out_prefix+'/model_deviation_n{n}_m{m}.pkl'.format(n=args.n, m=args.m)
+
+    try:
+        with open(out_file, 'rb') as fh:
+            import pickle
+            ttl, ttl_flat = pickle.load(fh)
+    except:
+        ttl_flat = {'pi':[], 'mu':[], 'all':[]}
+        ttl = {'mu':[], 'pi':[], 'all':[]}
+        eps_vals = list(np.linspace(0.0,1,11))
+
+        for fname in files:
+            print(fname)
+            params = parse_alignment_name(fname)
+
+            true_model = load_model(model_name(prefix, params))
+            n = true_model.n_states
+
+            tree = Phylo.read(tree_name(prefix, params), 'newick')
+            true_ttl = tree.total_branch_length()*params['m']
+
+            total_branch_length_pi = []
+            total_branch_length_mu = []
+            total_branch_length_all = []
+
+            for eps in eps_vals:
+                model = load_model(model_name(prefix, params), flatten_p=eps)
+                model.mu /= model.average_rate().mean()
+                tt = run_tree(fname, out_prefix, alphabet, model)
+                total_branch_length_pi.append((eps, tt.tree.total_branch_length(), true_ttl, model.mu.mean()))
+                print(total_branch_length_pi[-1])
+
+            for eps in eps_vals:
+                model = load_model(model_name(prefix, params), flatten_mu=eps)
+                model.mu /= model.average_rate().mean()
+                tt = run_tree(fname, out_prefix, alphabet, model)
[…]

### 2019-07-27 src/plot_toy_data.py

@@ -5,11 +5,18 @@ from collections import defaultdict
 fmts = ['.png', '.pdf']
 fs = 12
 
-labels = {'naive':"Alignment frequencies", "dressed":"Substitution counts", "regular":"Reconstructed",
-          "marginal":"ancestral sum", "iterative":"Iterative, reconstructed tree", 'branch_length':"non-linear",
-          "iterative_true":"iterative, true tree", 'marginal_true':"ancestral sum, true tree",
-          "optimize_tree":"Optimize tree and model", "optimize_tree_true":"Optimize tree(true) and model"}
-colors = {k:'C%d'%((i+1)%10) for i,k in enumerate(sorted(labels.keys()))}
+labels = {'naive':"Alignment frequencies",
+          "dressed":"Known ancestral sequences",
+          "regular":"Reconstructed ancestral sequences",
+          "marginal":"Marginalized ancestral sequences",
+          "iterative":"Iterative model estimation",
+          "optimize_tree":"Iterative tree and model optimization",
+          "optimize_tree_true":"Optimize tree(true) and model",
+          "iterative_true":"Iterative model estimation (true tree)",
+          "regular_true":"Reconstructed ancestral sequences",
+          "marginal_true":"Marginalized ancestral sequences",
+    }
+colors = {k:'C%d'%((i+1)%10) for i,k in enumerate(labels.keys())}
 
 def load_toy_data_results(path):
     mu_dist = defaultdict(lambda: defaultdict(list))
@@ -18,19 +25,22 @@ def load_toy_data_results(path):
     W_dist = defaultdict(lambda: defaultdict(list))
     delta_LH = defaultdict(lambda: defaultdict(list))
     avg_rate = defaultdict(lambda: defaultdict(list))
+    mucorr_dist = defaultdict(lambda: defaultdict(list))
 
     mu_vals = set()
     n_vals = set()
 
     for fname in glob.glob(path+'*.pkl'):
         with open(fname, 'rb') as fh:
-            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_p_entropy, tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu = pickle.load(fh)
+            tmp_mu_vals, tmp_n_vals, tmp_p_dist, tmp_p_entropy, \
+                tmp_mu_dist, tmp_W_dist, tmp_LH, tmp_avg_mu, tmp_mucorr = pickle.load(fh)
 
[…]

### 2019-07-27 src/reconstruct_toy_data.py

@@ -24,6 +24,7 @@ def assess_reconstruction(model, true_model):
             "model_entropy": -np.mean(np.sum(model.Pi*np.log(model.Pi), axis=0)),
             "true_entropy": -np.mean(np.sum(true_model.Pi*np.log(true_model.Pi), axis=0)),
             "chisq_mu": chisq(model.mu/model.average_rate().mean(), true_model.mu/true_model_average_rate),
+            "r_mu": np.corrcoef(model.mu, true_model.mu)[0,1],
             "chisq_W":chisq(model.W.flatten(), true_model.W.flatten())}
 
 
@@ -93,6 +94,7 @@ if __name__ == '__main__':
     W_dist = defaultdict(lambda: defaultdict(list))
     delta_LH = defaultdict(lambda: defaultdict(list))
     avg_rate = defaultdict(lambda: defaultdict(list))
+    mucorr_dist = defaultdict(lambda: defaultdict(list))
 
     mu_vals = set()
     n_vals = set()
@@ -140,6 +142,7 @@ if __name__ == '__main__':
         p_dist["dressed"][dset].append(accuracy["chisq_p"])
         p_entropy["dressed"][dset].append([accuracy["model_entropy"], accuracy["true_entropy"]])
         mu_dist["dressed"][dset].append(accuracy["chisq_mu"])
+        mucorr_dist["dressed"][dset].append(accuracy["r_mu"])
 
         for tree in [tree_name(prefix, params), reconstructed_tree_name(prefix, params)]:
             for ana in analysis_types_tree:
@@ -150,6 +153,7 @@ if __name__ == '__main__':
                     p_dist[ana+s][dset].append(accuracy["chisq_p"])
                     p_entropy[ana+s][dset].append([accuracy["model_entropy"], accuracy["true_entropy"]])
                     mu_dist[ana+s][dset].append(accuracy["chisq_mu"])
+                    mucorr_dist[ana+s][dset].append(accuracy["r_mu"])
 
                 delta_LH[ana+s][dset].append( (true_LH, accuracy['LH'] ))
                 avg_rate[ana+s][dset].append(( true_model_average_rate, accuracy['avg_rate']))
@@ -165,4 +169,4 @@ if __name__ == '__main__':
                                         for n in ['L', 'n', 'm'] if args.__getattribute__(n)]) + '.pkl'
         with open(out_fname, 'wb') as fh:
[…]

### 2019-07-27 manuscript/ms.tex
edits, added 'true_rate' option for branch length comparison

@@
 	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha3.0}.pdf}
-	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree.
-(A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences.
+	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
 	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
-	(C) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+	(B) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
 Regularization via pseudo-counts reduces over-fitting at low divergence.
@@
 	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha3.0}.pdf}
-	\caption{Errors stemming from tree and ancestral reconstruction: At large root-to-tip distances, ancestral reconstruction becomes more and more uncertain and estimation of equilibrium state frequencies fails.
-	These errors are gradually eliminated by first summing over ancestral uncertainty, iterative redoing ancestral reconstruction using the inferred model, and re-optimizating branch length using the updated models (or using the true tree).
-	The right panel shows analysis of simulated with four states (nucleotides), while the right panels shows the corresponding graph of 20 states (amino acids).
-As expected, saturation and non-linear effects set in later with larger alphabet.}
+	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
+	The left and right panels show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
+	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
+	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
+	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=3$.}
 	\label{fig:reconstructed}
@@
 Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence.
-Similar improvements are achieved by optimizing tree branch length along with the model.
+Similar improvements are achieved by optimizing tree branch lengths along with the model.
 
-\subsection*{Ignoring site specific frequencies dramatically underestimates branch lengths}
-As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in branch length underestimation if uniform equilibrium frequencies are assumed.
-This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^b}^2$ is increasing sharply with the skew.
+\subsection*{Ignoring site specific frequencies results in dramatic branch length underestimates}
+As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
+This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
 Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
+In other words, this underestimation is the result of incorrect models of saturation.
 
@@
 	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
-Panel B shows the results of the same optimization for the average root-to-tip distance.}
+Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
+Not that inferred site specific models only partially ameliorate underestimation, see main text.
+}
 	\label{fig:tree_length}
@@
 Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted tree (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
-While branch length are accurately estimated for when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
+While branch length are accurately estimated when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
 Underestimation is particularly severe for the root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
-The inferred root-to-tip distance is essentially independent from the true value for distance greater than 1.
-This effect is entirely due to skewed equilbrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^b=q^{-1}$.
-
+When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distance greater than 1 (Fig.~\ref{fig:tree_length}B).
+This effect is entirely due to skewed equilbrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$.
 %% dependence on nucleotide and amino acid skew
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_2019-06-01_simulatedData_L1000_ratealpha3.0_alpha1.0_nuc_n100_m0.35}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_n100_m0.35}.pdf}
 	\caption{{\bf Moderate model inaccuracies result in large substantial branch length underestimation}.
@@
 
+Surprisingly, the inferred site specific model only partially rectify the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
+The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation.
+If, however, preferences are estimated poorly for example through excessive regularization ($\pc=10$ in Fig.~\ref{fig:tree_length}), using the correct rates hardly improves the estimate.
+The latter is consistent with the observation that standard ML tree inference using IQ-tree with a GTR+R10 model severely underestimates branch lengths.
+
 \subsection*{Minor model deviations can have substantial effects on divergence estimates}
-In the previous section, we have shown that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
+In the previous section, we have observed that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
 To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$.
@@
 If the same preference are used for every site, however, deviations are substantial.
-Deviations in rate and preferences are approximatly additive.
+Deviations in rate and preferences are approximately additive.
+These observations are consistent with the finding above using inferred preferences and true substitution rates result in much more accurate branch length estimates than inferring both rates and preferences.
+
 Fig.~\ref{fig:model_deviation}B shows the degree of mis-estimation when using flat preferences, flat rates, or both as a function of the degree of divergence.
-Initially, errors accumulate linearly with increasing.
+The relative error increases approximately linearly with the average evolutionary rate.
 
@@
 One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
-Simple population genetic models predict that the fixation probability of state $i$ if starting from state $j$ depends on the fitness different $s_{ij}$ and the effective population size $N$ as
+Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as
 \begin{equation}
-	f_{ij} = \frac{1-e^{-2s_{ij}}}{1-e^{-2Ns_{ij}}} \approx
+	f^a_{ij} = \frac{1-e^{-2s^a_{ij}}}{1-e^{-2Ns^a_{ij}}} \approx
 	\begin{cases}
-		2s_{ij} & s_{ij}>0 \\
-		2s_{ij} e^{2Ns_{ij}} & s_{ij} < 0
+		2s^a_{ij} & s^a_{ij}>0 \\
+		2s^a_{ij} e^{2Ns^a_{ij}} & s^a_{ij} < 0
 	\end{cases}
 \end{equation}
-On longer time scales, the fixation rates $f_{ij}$ should correspond to transitions rates of the site specific GTR model $\muQ_{ij}$.
+On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site specific GTR model $\mu^a Q^a_{ij}$.
 The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations.
-Nevertheless, the logarithm of the ratio $\log f_{ij}/f_{jj} \approx 2Ns_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
-We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $a$
+Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
+We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $i$
 \begin{equation}
@@
 We compare this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
-Fig.~\ref{fig:HIV_fitness} shows a linear relation ship between the logarithmic fitness effects measured with-in host and from the phylogeny that explains about half the variance.
-The linear relationship holds both for nucleotide states and amino acid states.
+However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observe a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
+see Fig.~\ref{fig:HIV_fitness})
+This relationship explains about half the variance of nucleotide effects and slightly less of amino acid effects.
+
+The fact that this relationship deviates from the `cross-sectional$\sim$ with-in host' expectation and instead is approximately `cross-sectional $\sim \log$ with-in host' points towards different process that drive cross-sectional and with-in host diversification.
+Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
+The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_fitness}).
 
@@
 	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.001_aa}.pdf}
-	\caption{{\bf Intra-host vs cross-sectional mutation selection balance are consistent. } }
+	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
+	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
+	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
 	\label{fig:HIV_fitness}
@@
 \section*{Discussion}
-Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize and a common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
-Different positions in a DNA sequence or protein, however, also differ in the preferences for different states.
+Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize.
+A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
+Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site specific preference is the basis of all common homology search tools.
 While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern1998evolutionary}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
@@
 
-Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-stype approach inspired by the original paper by \citet{bruno1996modeling}.
+Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-stype approach inspired by \citet{bruno1996modeling}.
 We show that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
 For short branch length a parsimonious ancestral reconstruction is sufficient, while for more diverged samples and iterative inference of the model, the ancestral states, and the branch lengths is necessary.
-The computational complexity of the inference is dominated by ancestral reconstruction and branch length optimization, which is quadratic in the alphabet size and linear in sample size and sequence length.
+The computational complexity of the inference is dominated by ancestral reconstruction and branch length optimization, which is quadratic in the alphabet size and linear in sample size and sequence length \citep{felsenstein2004inferring}.
+We have implemented the algorithm for site-specific model inference and branch lengths optimization in TreeTime \citep{sagulenko2017treetime}.
 
-We further explore how branch length estimation depends on model choice and inference accuracy.
-As has been reported before, branch length are underestimated with variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
-Even moderate model deviation can result in substantial underestimates.
+We further explored how model choice affects the maximum likelihood estimates of branch length.
+As has been reported before, branch length are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
+While such underestimation often has a moderate effect on the total length of a tree, it can result in substantial underestimation of root-to-tip distance that are dominated by a few long branches close to the root.
+While using the correct model results in correct branch length estimates, joint inference of site-specific models and branch lengths is typically unable to recover the true branch lengths and substantial underestimation remains.
 This mirrors observations by \citet{hilton_modeling_2018}, who found that preferences measured for influenza HA proteins of type H1 or H3 affect branch length of in the vicinity of the focal sequence, but not globally on the tree.
@@
 Typically, short branches close to the tips of the tree are used to calibrate molecular clock models.
-The deep branches, however, tend to be much longer and rate estimation and variation of site specific preferences will result in saturation effects not accoutned for by the model \citep{hilton_modeling_2018}.
-This effective variation in rate is related to the effect of transient deletrious mutations that inflate the rate of short time scales.
+The deep branches, however, tend to be much longer and rate estimation and variation of site specific preferences will result in saturation effects not accounted for by the model \citep{hilton_modeling_2018}.
+This effective variation in rate is related to the effect of transient deleterious mutations that inflate the rate on short time scales \citep{ho_time_2005}: the time it takes purifying selection to prune deleterious mutations is related to the relaxation time scales of GTR models with site specific preferences.
+Instead of the $q-1$ degenerate eigenvalues of a Jukes-Cantor model, each site has a spectrum of eigenvalues and different eigenmodes relax at different speeds, generating apparently time dependent rates.
+
+While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint infererence of model and branch lengths remains difficult.
+This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one primary reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
 
@@
 
-\appendix
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
 \onecolumngrid
 
-\section*{Model and Notation}
+\section*{Materials and Methods}
+\subsection*{Model and notation}
+\label{sec:MM_model}
 Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
@@
 
-\section*{Iterative optimization algorithm}
+\subsection*{Iterative optimization algorithm}
 The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
@@
 
-\section*{HIV sequence analysis}
+\subsection*{HIV sequence analysis}
 HIV-1 sequences were downloaded from LANL HIV database \citep{LANL} setting filters to ``one sequence per patient'', ``non-ACGT\textless 0.3''.
@@
 \clearpage
+\appendix
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
 
 \section*{Supplementary figures}
-\begin{figure*}[tb]
+\begin{figure*}[htb]
 	\centering
@@
 (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
 	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
@@
 \end{figure*}
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.001}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.001}.pdf}
+	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
+	In analogy to Fig.~\ref{fig:HIV_fitness}, the panels A\&B show the ratio of in/out rates for consensus nucleotides acids along the \gene{gag} and \gene{nef} gene of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
+	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
+	\label{fig:HIV_fitness}
+\end{figure*}
+
+\begin{figure*}[htb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.001}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.001_aa}.pdf}
+	\caption{{\bf Diversity and GTR rate estimates.}
+	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs the diversity in columns of the corresponding alignment.}
+	\label{fig:HIV_diversity}
+\end{figure*}
+
 \end{document}

### 2019-07-27 src/analyze_HIV_tree.py

@@ -131,8 +131,9 @@ if __name__ == '__main__':
     plt.legend(fontsize=fs)
     plt.xlim([3e-5, 3e0])
     plt.ylim([3e-2, 3e3])
-    plt.xlabel('intra-host fitness estimates', fontsize=fs)
-    plt.ylabel('GTR based fitness estimates', fontsize=fs)
+    plt.title('amino acids' if args.aa else 'nucleotides',fontsize=fs)
+    plt.xlabel('intra-host fitness cost estimates', fontsize=fs)
+    plt.ylabel(r'GTR based fitness estimates $\Gamma_{in}/\Gamma_{out}$', fontsize=fs)
     plt.xscale('log')
     plt.yscale('log')
     plt.tick_params(labelsize=0.8*fs)
@@ -144,3 +145,18 @@ if __name__ == '__main__':
     plt.legend()
     plt.xscale('log')
     plt.yscale('log')
+
+    plt.figure()
+    cc_div = ccoef(np.log(1e-10+div_aln), np.log(1e-10+fitness_estimates_gtr))
+    plt.scatter(div_aln, fitness_estimates_gtr, label=r'$r^2=$'+'%1.2f'%cc_div[0]**2)
+    plt.legend(fontsize=fs)
+    plt.title('amino acids' if args.aa else 'nucleotides',fontsize=fs)
+    plt.xlim([2e-4, 3e0])
+    plt.ylim([3e-2, 9e3])
+    plt.xlabel('diversity in alignment', fontsize=fs)
+    plt.ylabel(r'GTR based fitness estimates $\Gamma_{in}/\Gamma_{out}$', fontsize=fs)
+    plt.xscale('log')
+    plt.yscale('log')
+    plt.tick_params(labelsize=0.8*fs)
+    plt.tight_layout()
+    plt.savefig('figures/'+args.prefix.split('/')[-1]+'_alignment_div_pc_%1.3f'%args.pc+ ('_aa' if args.aa else '')+'.pdf')

### 2019-07-27 src/calculate_branch_length.py

@@ -10,7 +10,7 @@ from reconstruct_toy_data import *
 
 
 
-def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False, pc=0.1):
+def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False, pc=0.1, true_rates=False):
     params = parse_alignment_name(fname)
     m = params['m']
     tree = Phylo.read(tree_name(prefix, params), 'newick') if true_tree else  Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
@@ -24,8 +24,9 @@ def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False, pc=
         if true_tree:
             n.branch_length *= m*(0.6+0.4*np.random.random())
     print(np.sum(old_bl)*(m if true_tree else 1.0),m)
+    true_GTR = load_model(model_name(prefix, params))
     if true_model:
-        model = load_model(model_name(prefix, params))
+        model = true_GTR
         model.mu/=m
     else:
         model='JC69'
@@ -39,14 +40,20 @@ def run_tree(fname, out_prefix, alphabet, true_tree=False, true_model=False, pc=
     tt.optimize_tree(branch_length_mode='marginal', max_iter=n_iter,
                      infer_gtr=not true_model, site_specific_gtr=True, pc=pc, damping=0.75)
 
+    if true_rates:
+        tt.gtr.mu = true_GTR.mu/m
+        tt.optimize_tree(branch_length_mode='marginal', max_iter=n_iter,
+                         infer_gtr=False, site_specific_gtr=True, pc=pc, damping=0.75)
+
+
     new_bl = []
     for n in tt.tree.find_clades():
         new_bl.append(n.branch_length)
 
 
     print(tt.tree.total_branch_length(),tt.gtr.average_rate().mean())
-    # tt.tree.root_at_midpoint()
-    tfname = reoptimized_tree_true_model(out_prefix, params) if args.true_model else reoptimized_tree(out_prefix, params)
+    tt.tree.root_at_midpoint()
+    tfname = reoptimized_tree_true_model(out_prefix, params) if args.true_model else reoptimized_tree(out_prefix, params, true_rates=true_rates)
     Phylo.write(tt.tree, tfname, 'newick')
 
     print(np.mean([x for c,x in tt.tree.depths().items() if c.is_terminal()]), tt.tree.total_branch_length())
@@ -60,15 +67,22 @@ if __name__ == '__main__':
[…]

### 2019-07-27 src/compare_tree_length.py

@@ -11,6 +11,7 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
     parser.add_argument("-m", type=float, help="simulated mutation rate")
     parser.add_argument("-n", type=int, help="number of taxa")
+    parser.add_argument("--true-rates", action='store_true', help="include estimates with true rates")
     parser.add_argument("--pc", nargs='+', default=[0.1],type=float, help="pseudo count")
     parser.add_argument("-L", type=int, help="length of sequence")
     parser.add_argument("--prefix", type=str, required=True, help="folder to retrieve data from")
@@ -23,10 +24,17 @@ if __name__ == '__main__':
     depth = {}
     terminal_bl = {}
 
-    for dset in ['true', 'ML']+args.pc:
+    for dset in ['true', 'ML']+[(pc, 'inferred-rates') for pc in args.pc]:
         length[dset]=[]
         depth[dset]=[]
         terminal_bl[dset]=[]
+
+    if args.true_rates:
+        for dset in [(pc, 'true-rates') for pc in args.pc]:
+            length[dset]=[]
+            depth[dset]=[]
+            terminal_bl[dset]=[]
+
 
     mask = "/L{L}_n{n}_m{mu}_*.fasta.gz".format(L=args.L or '*', n=args.n or '*', mu=args.m or "*")
     files = glob.glob(prefix+mask)
@@ -40,16 +48,19 @@ if __name__ == '__main__':
         true_ttl = 0.5*true_tree.total_branch_length()/params['n']
         true_depth = np.mean([x for c,x in true_tree.depths().items() if c.is_terminal()])
         true_terminal = np.mean([x.branch_length for x in true_tree.get_terminals()])
-        for dset in ['true', 'ML']+args.pc:
+        true_rates = False
+        for dset in depth.keys():
             result_prefix = args.prefix+'_results'
-            if type(dset)==float:
-                result_prefix += '_pc_%1.2f'%dset
+            if type(dset)==tuple:
+                result_prefix += '_pc_%1.2f'%dset[0]
+                true_rates = dset[1]=='true-rates'
+
             if dset=='ML':
                 tree = Phylo.read(reconstructed_tree_name(prefix, params), 'newick')
             elif dset=='true':
                 tree = Phylo.read(reoptimized_tree_true_model(result_prefix, params), 'newick')
             else:
[…]

### 2019-07-27 src/filenames.py

@@ -36,8 +36,8 @@ def tree_name(prefix, params):
 def reconstructed_tree_name(prefix, params):
     return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reconstructed.nwk'.format(**params)
 
-def reoptimized_tree(prefix, params):
-    return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reoptimized.nwk'.format(**params)
+def reoptimized_tree(prefix, params, true_rates=False):
+    return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_{tr}_reoptimized.nwk'.format(**params, tr="trueRates" if true_rates else "")
 
 def reoptimized_tree_true_model(prefix, params):
     return prefix+'/L{L}_n{n}_m{m}_tree{tree:02d}_model{model:02d}_seqgen{seqgen:02}_reoptimized_true_model.nwk'.format(**params)

### 2019-07-27 src/plot_toy_data.py

@@ -256,7 +256,7 @@ if __name__ == '__main__':
     plot_rate_correlation(data, n_vals_to_plot, mu_vals, [0.1, 0.5, 1.0], methods=['dressed'],
                         fname='figures/rate_correlation_dressed'+suffix)
 
-    #### Fig 3: comparison of different models as a function of tree length for one pc
+    #### comparison of different inference schemes as a function of tree length for one pc
     plot_pdist_vs_rtt(data[rate_alpha][pc_general], n_vals_to_plot, mu_vals,
                       methods=['naive', 'regular', 'marginal',
                                'iterative','iterative_true', 'optimize_tree','dressed'],

### 2019-07-27 submit_branch_length.py

@@ -1,5 +1,6 @@
 import os
 from datetime import date
+from itertools import product
 
 L=1000
 muvals =  [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]
@@ -10,20 +11,24 @@ parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
 parser.add_argument("--muvals", nargs='+', default=muvals, type=float,  help="number of taxa")
 parser.add_argument("--submit", action='store_true', help="submit")
 parser.add_argument("--date", type=str,  help="date prefix")
+parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
 args=parser.parse_args()
+dsets = [(1.0, False, False)] #, (0.0, True, False)]
+#dsets = [(0.2, False, True), (0.5, False, True)]
 
 date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
-for pc, model in [(10.0, False),(1.0, False), (0.1, False), (0,True)]:
-    for alpha, JC, aa in [(1.0, False, False), (0.0, True, False)]: #, (0.2, False, True), (0.5, False, True)]:
-        prefix = '%s_simulatedData_L%d_alpha%1.1f_%s'%(date_prefix, L, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
+for pc, model, rates in [(10.0, False, False),(1.0, False, False), (0.1, False, False),
+                         (10.0, False, True),(1.0, False, True), (0.1, False, True), (0,True, False)]:
+    for alpha, JC, aa in dsets:
+        prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(date_prefix, L, args.rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
+
         for n in args.nvals:
             submit_script = "submit.sh" if n>500 else "submit_small.sh"
             for mu in args.muvals:
-                cmd = 'sbatch {script} src/calculate_branch_length.py -L {L} -n {n} -m {mu} --prefix {prefix} {aa} --pc {pc} {tm} --true-tree'.format(
-                            script=submit_script, L=L, n=n, mu=mu, prefix=prefix,
-                            aa="--aa" if aa else "", pc=pc, tm='--true-model' if model else '')
-                print(cmd)
-                if args.submit:
-                    os.system(cmd)
[…]

### 2019-07-27 submit_gendata.py

@@ -8,19 +8,26 @@ import argparse
 parser = argparse.ArgumentParser(description = "", usage="generate_toy_data")
 parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
 parser.add_argument("--muvals", nargs='+', default=muvals, type=float,  help="number of taxa")
+parser.add_argument("--date", type=str,  help="date prefix")
 parser.add_argument("--submit", action='store_true', help="submit")
 args=parser.parse_args()
-today = date.today().strftime('%Y-%m-%d')
+today = args.date or date.today().strftime('%Y-%m-%d')
+rate_alpha=0
+datasets = [#(1.0, False, False),
+    (0.0, True, False)] #, (0.2, False, True), (0.5, False, True)]
 
-for alpha, JC, aa in [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]:
-    prefix = '%s_simulatedData_L%d_alpha%1.1f_%s'%(today, L, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
+for alpha, JC, aa in datasets:
+    prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(today, L, rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
 
     for n in args.nvals:
         submit_script = "submit.sh" if n>500 else "submit_small.sh"
         for mu in args.muvals:
-            cmd = 'sbatch {script} src/generate_toy_data.py -L {L} -n {n} -m {mu} --prefix {prefix} {JC} {aa} --alpha {alpha}'.format(
-                   script=submit_script, L=L, n=n, mu=mu, prefix=prefix, JC="--JC" if JC else "", aa="--aa" if aa else "", alpha=alpha)
-            print(cmd)
-            if args.submit:
-                os.system(cmd)
+            cmd = 'sbatch {script} src/generate_toy_data.py -L {L} -n {n} -m {mu} --prefix {prefix} {JC} {aa} --alpha {alpha} --rate-alpha {rate_alpha}'.format(
+                   script=submit_script, L=L, n=n, mu=mu, prefix=prefix, JC="--JC" if JC else "", aa="--aa" if aa else "", alpha=alpha, rate_alpha=rate_alpha)
+            target = prefix+'/L{L}_n{n}_m{m}_tree00.nwk'.format(n=n, m=mu, L=L)
+            print(target)
[…]

### 2019-07-27 submit_model_deviation.py

@@ -1,0 +1,30 @@
+import os
+from datetime import date
+from itertools import product
+
+L=1000
+
+import argparse
+parser = argparse.ArgumentParser(description = "", usage="generate_toy_data")
+parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
+parser.add_argument("--muvals", nargs='+', type=float,  help="mutation rate for panelB")
+parser.add_argument("--submit", action='store_true', help="submit")
+parser.add_argument("--date", type=str,  help="date prefix")
+parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
+
+args=parser.parse_args()
+aa = False
+JC = False
+alpha=1.0
+
+date_prefix = args.date or date.today().strftime('%Y-%m-%d')
+
+prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(date_prefix, L, args.rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
+for n in args.nvals:
+    submit_script = "submit.sh" if n>500 else "submit_small.sh"
+    for mu in args.muvals:
+        cmd = f"sbatch {submit_script} src/model_deviation.py -L {L} -n {n} -m {mu} --prefix {prefix}"
+        print(cmd)
+        if args.submit:
+            os.system(cmd)
+

### 2019-07-27 submit_reconstruction.py

@@ -10,16 +10,17 @@ parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
 parser.add_argument("--muvals", nargs='+', default=muvals, type=float,  help="number of taxa")
 parser.add_argument("--submit", action='store_true', help="submit")
 parser.add_argument("--date", type=str,  help="date prefix")
+parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
 args=parser.parse_args()
 
 date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
 for pc in [0.1, 0.5, 1.0]:
     for alpha, JC, aa in [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]:
-        prefix = '%s_simulatedData_L%d_alpha%1.1f_%s'%(date_prefix, L, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
+        prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(date_prefix, L, args.rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
 
         for n in args.nvals:
-            submit_script = "submit.sh" if n>500 else "submit_small.sh"
+            submit_script = "submit.sh" if n>200 else "submit_small.sh"
             for mu in args.muvals:
                 target = prefix+'_results_pc_%1.2f/L%d_n%d_m%s.pkl'%(pc, L,n, str(mu))
                 print(target)

### 2019-07-28 manuscript/ms.tex
ms edits, more HIV analysis

@@
 \newcommand{\eqp}{p}
+\newcommand{\pc}{c}
 \newcommand{\LH}{\mathcal{L}}
@@
 	\begin{split}
-		\mu^b & \leftarrow \mu^b\frac{C+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
-		\eqp^b_i & \leftarrow \eqp^b_i\frac{C+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(qC+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
+		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
+		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
 		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
@@
 Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
-$C$ is a pseudocount analogous to a Dirichlet prior that  in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $C$ over the total tree length.
+$\pc$ is a pseudocount analogous to a Dirichlet prior that  in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
 To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
@@
 Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
-In other words, this underestimation is the result of incorrect models of saturation.
+In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate estimates \citet{wertheim_purifying_2011}.
 
@@
 
-Surprisingly, the inferred site specific model only partially rectify the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
+Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
 The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation.
@@
 
+Next, we quantified the effect that ignoring site-specific preferences has on branch length estimates.
+We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with the average preferences and the same per site rate variation.
+As soon as the length of the simulated branch approaches 0.2, the length inferred by a model without site-specific preferences deviates substantially from the true value and saturates around 0.5 for larger and larger $t$.
+As expected, these deviations are even more severe when estimating branch length as simple sequence divergence (p-distance), while using the generating model reproduces the correct branch length.
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.8\textwidth]{{../figures/HIV_B_pol_branch_length_error_pc_0.010}.pdf}
+	\caption{{\bf Underestimation of divergence in HIV.}
+	  Panel A shows the estimated ML branch length for the model used to generate the sequences and a model with constant equilibrium frequencies as a function of true branch length.
+For comparison, the p-distance between the two simulated sequences is also shown.
+The top axis shows branch length in units of years assuming a substitution rate of 0.002/year.
+Error bars denote one standard deviation.
+      Panel B shows the distribution of rates across sites for both models.
+They differ slightly in the overall scale since rates have been rescaled such that the average substitution rate in both models is identical.}
+	\label{fig:HIV_branch_length}
+\end{figure*}
+
 \section*{Discussion}
@@
 
-We further explored how model choice affects the maximum likelihood estimates of branch length.
+We further explored how model choice affects the maximum likelihood estimates of branch lengths.
 As has been reported before, branch length are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
@@
 This mirrors observations by \citet{hilton_modeling_2018}, who found that preferences measured for influenza HA proteins of type H1 or H3 affect branch length of in the vicinity of the focal sequence, but not globally on the tree.
+Using site-specific models inferred from HIV alignments, we quantified the error made when ignoring site-specific preferences.
+While the true models are unknown in this case, this analysis nevertheless suggests that errors are substantial as soon as branch length exceed $t=0.2$ and sequence divergence saturates at levels around 0.3.
+This result is consistent with the observation of frequent reversion of HIV to a preferred sequence state \citep{zanini2015population,carlson_selection_2014,leslie_hiv_2004}.
+Average nucleotide distances between \gene{pol} sequences between HIV-1 subtypes are on the order of 0.1, HIV-1 and SIVcpz are on the order of 0.25 and between sequences in the HIV-1/HIV-2/SIV compendium alignment about 0.35 (all distances ignore sites with $>20$\% gaps).
 
-The potentially large error in branch length estimating can seriously affect deep divergence time estimates.
+This potentially large error in branch length estimates can seriously affect deep divergence time estimates.
 Typically, short branches close to the tips of the tree are used to calibrate molecular clock models.
 The deep branches, however, tend to be much longer and rate estimation and variation of site specific preferences will result in saturation effects not accounted for by the model \citep{hilton_modeling_2018}.
-This effective variation in rate is related to the effect of transient deleterious mutations that inflate the rate on short time scales \citep{ho_time_2005}: the time it takes purifying selection to prune deleterious mutations is related to the relaxation time scales of GTR models with site specific preferences.
+This effective variation in rate is related to the effect of transient deleterious mutations that inflate the rate on short time scales \citep{ho_time_2005,wertheim_purifying_2011}: the time it takes purifying selection to prune deleterious mutations is related to the relaxation time scales of GTR models with site specific preferences.
 Instead of the $q-1$ degenerate eigenvalues of a Jukes-Cantor model, each site has a spectrum of eigenvalues and different eigenmodes relax at different speeds, generating apparently time dependent rates.
@@
 
-\begin{itemize}
-	\item underestimation of long branches, dubious deep time trees
-	\item relation to frequent reversions
-	\item fitness inference, functional constraints
-	\item caveat about interactions
-\end{itemize}
+Estimating site-specific models from sequence alignments faces one fundamental problem:
+Reliable estimates require observation of many changes at each site which requires many sufficiently diverged sequences.
+At the same time, preferences at individual sites are expected to change due to epistatic interactions with other sites in the sequence, as for example witnessed by the gradual divergence of experimentally determined preferences \citep{doud_site-specific_2015,haddox_mapping_2018}.
+Hence the approach described in this work is largely restricted to cases like HIV where many moderately diverged sequences ($\sim$10\%) are available.
+Outside of this limit there either isn't enough data to reliably estimate the large number of coefficients, or epistatic interactions need to be taken into account -- probably at the expense of ignoring phylogenetic signals \citep{morcos_direct-coupling_2011}.
 
@@
 \end{equation}
-where $C$ is a pseudocount analogous to a Dirichlet prior that in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $C$ over the total tree length.
+where $\pc$ is a pseudocount analogous to a Dirichlet prior that in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
 To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.

### 2019-07-28 src/HIV_branch_length.py

@@ -1,0 +1,95 @@
+import os, gzip, glob, pickle
+from matplotlib import pyplot as plt
+
+from treetime import TreeAnc, GTR_site_specific, GTR
+from treetime.seq_utils import seq2prof, profile_maps, alphabets, prof2seq
+
+from estimation import *
+from generate_toy_data import save_model, load_model
+
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    hiv_out = 'HIV_inferences/'
+    parser.add_argument("--prefix", type=str, help="data set")
+    parser.add_argument("--gene", type=str, default='pol')
+    parser.add_argument("--subtype", type=str, default='B')
+    parser.add_argument("--aa", action='store_true', help="use amino acid alphabet")
+    parser.add_argument("--pc", type=float, default=1.0, help="pseudocount for reconstruction")
+    args=parser.parse_args()
+
+    alphabet='aa' if args.aa else 'nuc_nogap'
+    pc = args.pc
+    fs = 16
+    dset_name=os.path.basename(args.prefix)
+    model_name = hiv_out + dset_name+'_aligned.%1.2f_inferred_model%s.npz'%(args.pc, '_aa' if args.aa else '')
+
+    gtr=None
+    aln_freq=None
+    try:
+        gtr = load_model(model_name)
+    except:
+        pass
+
+    #gtr.Pi = gtr.Pi.mean(axis=1)
+    S = -np.sum(np.log(gtr.Pi)*gtr.Pi, axis=0)
+    print('Entropy:', S.mean(), S.std())
+    gtr.mu /= gtr.average_rate().mean()
+    gtr_flat = GTR_site_specific.custom(alphabet=gtr.alphabet,  mu=gtr.mu, W=gtr.W, pi=gtr.Pi.mean(axis=1))
+    gtr_flat.mu /= gtr_flat.average_rate().mean()
+
+    t=0.1
+    s1 = prof2seq(gtr.Pi.T, gtr, sample_from_prof=True)
+    s1_prof = seq2prof(s1[0], profile_map=gtr.profile_map)
+
+    branch_length = []
+    for r in range(20):
+        tmp_branch_length = []
+        for t in np.linspace(0,1,31):
[…]

### 2019-07-28 src/HIV_ref_alignment.py

@@ -1,0 +1,56 @@
+import numpy as np
+from Bio import AlignIO
+from matplotlib import pyplot as plt
+
+def get_distance(aln1, aln2):
+    diversity = np.ones(aln1.shape[1])
+    for nuc in ['A', 'C', 'G', 'T', '-']:
+        diversity -= np.mean(aln1==nuc, axis=0)*np.mean(aln2==nuc, axis=0)
+    return diversity
+
+aln = AlignIO.read('HIV_data/HIV1_REF_2010_pol_DNA.fasta', 'fasta')
+seq_names = [s.id for s in aln]
+
+aln_array = np.array(aln)
+
+ungapped = np.mean(aln_array=='-', axis=0)<0.2
+
+aln_array_gap_stripped = aln_array[:,ungapped]
+overall_div = get_distance(aln_array_gap_stripped, aln_array_gap_stripped)
+print("overall diversity:", overall_div.mean())
+
+
+HIV1_subtypes = ['B', 'C', 'H', 'D']
+CPZ_sequences = ['CPZ']
+
+for st1 in HIV1_subtypes:
+    ind1 = np.array([s.split('.')[1]==st1 for s in seq_names])
+    for st2 in HIV1_subtypes:
+        ind2 = np.array([s.split('.')[1]==st2 for s in seq_names])
+        dist = get_distance(aln_array_gap_stripped[ind1], aln_array_gap_stripped[ind2])
+        print(f"{st1}-{st2}:{dist.mean()}")
+
+
+for st1 in HIV1_subtypes+CPZ_sequences:
+    ind1 = np.array([s.split('.')[1]==st1 for s in seq_names])
+    for st2 in CPZ_sequences:
+        ind2 = np.array([s.split('.')[1]==st2 for s in seq_names])
+        dist = get_distance(aln_array_gap_stripped[ind1], aln_array_gap_stripped[ind2])
+        print(f"{st1}-{st2}:{dist.mean()}")
+
+
+
+
+
+aln_SIV = AlignIO.read('HIV_data/SIV_COM_2017_pol_DNA.fasta', 'fasta')
+seq_names_SIV = [s.id for s in aln]
+
+aln_array_SIV = np.array(aln_SIV)
+ungapped_SIV = np.mean(aln_array_SIV=='-', axis=0)<0.2
+aln_array_gap_stripped_SIV = aln_array_SIV[:,ungapped_SIV]
+
[…]

### 2019-07-28 submit_model_deviation.py

@@ -13,15 +13,15 @@ parser.add_argument("--date", type=str,  help="date prefix")
 parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
 
 args=parser.parse_args()
-aa = False
+aa = True
 JC = False
-alpha=1.0
+alpha=0.2
 
 date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
 prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(date_prefix, L, args.rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
 for n in args.nvals:
-    submit_script = "submit.sh" if n>500 else "submit_small.sh"
+    submit_script = "submit.sh"
     for mu in args.muvals:
         cmd = f"sbatch {submit_script} src/model_deviation.py -L {L} -n {n} -m {mu} --prefix {prefix}"
         print(cmd)

### 2019-12-28 manuscript/ms.tex
update figures, small edits in text

@@
 
-\title{Site-specific substitution model inference with iterative tree likelihood maximization.}
+\title{Potential and limits of site-specific substitution model inference.}
 \author{Vadim Puller$^{1,2,3}$ Pavel Sagulenko$^{1}$, Richard Neher$^{1,2,3}$}
@@
 
-Inferring phylogenies is a computationally challenging problem since the number of phylogenies grows super-exponentially with the number of taxa and because the calculation of the likelihood is compulationally costly (though linear in the number of taxa).
+Inferring phylogenies is a computationally challenging problem since the number of phylogenies grows super-exponentially with the number of taxa and because the calculation of the likelihood is computationally costly (though linear in the number of taxa).
 Due to this computational complexity, the most commonly used substitution models are simple caricatures of biological complexity.
 The simplest substitution models assume that all sites and sequence states are equivalent and evolve at the same rate, i.e., they assume an unconstrained non-functional sequence that mutates at random between the different sequence states.
-Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result, among other issues, in biased estimates of divergence times \citep{yang1996among}.
+Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result in biased estimates of divergence times or otherwise erroneous results \citep{yang1996among}.
 Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
@@
 In addition to rate variation, different sites in a protein differ in amino acid they allow.
-Recent deep mutational scanning experimentes have shown that site specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
-Using such experimentally inferred site specific models in phylogenetic inference greatly improves the likelihood of the data \cite{bloom2014experimentally}.
+Recent deep mutational scanning experiments have shown that site specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
+Using such experimentally inferred site specific models in phylogenetic inference greatly increases the likelihood of the data \cite{bloom2014experimentally}.
 More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
 \citet{hilton_modeling_2018} recently showed that experimentally measured preference not only improve the phylogenetic fit, but also results in longer branch length estimates.
+Models with site-specific preferences are also known as mutation-selection balance models \citep{bruno1996modeling,yang2008mutation} reflecting the intuition that equilibrium frequencies are determined by competition of diversifying processes (like mutation) and selection for an optimal function.
 
-On the otherhand, estimating such models, also known as mutation-selection balance models \citep{bruno1996modeling,yang2008mutation}, from data exacerbates the over-fitting problem.
+While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
 In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
 	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
 	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences per tree.
 	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
-	(B) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
 Regularization via pseudo-counts reduces over-fitting at low divergence.
@@
 Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
-$\pc$ is a pseudocount analogous to a Dirichlet prior that  in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
+$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
 To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
@@
 
-To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicity recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites.
+To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites.
 We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha3.0}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha1.5}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha1.5}.pdf}
 	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
@@
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
+This scaling becomes more accurate with less rate variation.
 At the largest substitution rates, branch length are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
-Nevertheless, the acccuracy of the $\hat{\eqp}_i^a$ continues to increase.
+Nevertheless, the accuracy of the $\hat{\eqp}_i^a$ continues to increase.
 By contrast, using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry.
 
-The accuracy of substitution rates, however, is affected by linearization and we consistently underestimate $\mu^b$.
+The accuracy of substitution rates is affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
 This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
-The relative subsitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.
+The relative substitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.
 
 %%% effect of reconstruction
-In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized)
-To test the influence of tree and ancestral state reconstruction, we built phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{Price} using the default 20 category model for simulated nucleotide and amino-acid sequences, respectively.
+In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized).
+To test the influence of tree and ancestral state reconstruction, we built phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for simulated nucleotide and amino-acid sequences, respectively.
 Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
-The simplest approach is to take the inferred tree as given, reconstruct the ancestral states a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
-This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate.
-Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvemet in accurary.
-More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site specific model.
+The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
+This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
+Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}.
+More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site specific model (``Iterative model estimation'', brown line).
 The accuracy now continues to improve up to root-to-tip distances of about one.
-At this level of divergence, tree reconstruction starts becoming problematic and branch length deviate substantially from their true values.
-Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence.
+At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
+Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence (yellow line).
 Similar improvements are achieved by optimizing tree branch lengths along with the model.
+
+We found little difference between 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.
 
@@
 Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
-In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate estimates \citet{wertheim_purifying_2011}.
+In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate \citet{wertheim_purifying_2011}.
 
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_length}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_depth}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/nuc_length_n300}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n300}.pdf}
 	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
@@
 While branch length are accurately estimated when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
-Underestimation is particularly severe for the root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
-When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distance greater than 1 (Fig.~\ref{fig:tree_length}B).
-This effect is entirely due to skewed equilbrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$.
+Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
+When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
+This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
 %% dependence on nucleotide and amino acid skew
-The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed.
+The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed, see Fig.~\ref{fig:tree_length_supp}.
 
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_2019-06-01_simulatedData_L1000_ratealpha1.5_alpha1.0_nuc_n100_m0.35}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation}.pdf}
 	\caption{{\bf Moderate model inaccuracies result in large substantial branch length underestimation}.
@@
 Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
-The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation.
-If, however, preferences are estimated poorly for example through excessive regularization ($\pc=10$ in Fig.~\ref{fig:tree_length}), using the correct rates hardly improves the estimate.
-The latter is consistent with the observation that standard ML tree inference using IQ-tree with a GTR+R10 model severely underestimates branch lengths.
+The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.
+%If, however, preferences are estimated poorly for example through excessive regularization ($\pc=10$ in Fig.~\ref{fig:tree_length}), using the correct rates hardly improves the estimate.
+%The latter is consistent with the observation that standard ML tree inference using IQ-tree with a GTR+R10 model severely underestimates branch lengths.
 
@@
 In the previous section, we have observed that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
-To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$.
+To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows.
+For a mixing fraction $\alpha$, we constructed a model with rates
+\begin{equation}
+	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
+\end{equation}
+where $\langle \mu^a\rangle$ is the expectation value of the rate.
+Mixtures of site-specific frequencies are constructed analogously.
 
 Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
-Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order 0.3.
-If the same preference are used for every site, however, deviations are substantial.
+Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
+If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
 Deviations in rate and preferences are approximately additive.
@@
 
-The fact that this relationship deviates from the `cross-sectional$\sim$ with-in host' expectation and instead is approximately `cross-sectional $\sim \log$ with-in host' points towards different process that drive cross-sectional and with-in host diversification.
+The fact that this relationship deviates from the expectation
+\begin{equation}
+	\nonumber
+	\textrm{cross-sectional} \sim \textrm{with-in host}
+\end{equation}
+and instead is approximately
+\begin{equation}
+	\nonumber
+	\textrm{cross-sectional} \sim \log\left(\textrm{with-in host}\right)
+\end{equation}
+points towards different process that drive cross-sectional and with-in host diversification.
+While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectionally.
 Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.001}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.001_aa}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.010}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_fitness_pc_0.010_aa}.pdf}
 	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
@@
 While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern1998evolutionary}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
-However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data sets are increasing rapidly such that more complex models can be estimated.
+However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data set sizes are increasing rapidly such that more complex models can be estimated.
 Large data sets, however, require efficient and scalable methods.
 
-Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-stype approach inspired by \citet{bruno1996modeling}.
+Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
 We show that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
@@
 The computational complexity of the inference is dominated by ancestral reconstruction and branch length optimization, which is quadratic in the alphabet size and linear in sample size and sequence length \citep{felsenstein2004inferring}.
-We have implemented the algorithm for site-specific model inference and branch lengths optimization in TreeTime \citep{sagulenko2017treetime}.
+We have implemented the algorithm for site-specific model inference and branch length optimization in TreeTime \citep{sagulenko2017treetime}.
 
 We further explored how model choice affects the maximum likelihood estimates of branch lengths.
-As has been reported before, branch length are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
+As has been reported before, branch lengths are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
 While such underestimation often has a moderate effect on the total length of a tree, it can result in substantial underestimation of root-to-tip distance that are dominated by a few long branches close to the root.
@@
 
-While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint infererence of model and branch lengths remains difficult.
+While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint inference of model and branch lengths remains difficult.
 This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one primary reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
@@
 \begin{equation}
-\label{eq:update}
+\label{eq:update_app}
 	\begin{split}
@@
 \end{equation}
-where $\pc$ is a pseudocount analogous to a Dirichlet prior that in absense of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
+where $\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
 To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha1.5}.pdf}
 	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha3.0}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha1.5}.pdf}
 	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree, for amino acid alphabets.
@@
 	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
-	(C) The relative subsitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
 Regularization via pseudo-counts reduces over-fitting at low divergence.
 }
-	\label{fig:dressed}
+	\label{fig:dressed_supp}
 \end{figure*}
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.001}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.001}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/aa_length_n300}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/aa_depth_n300}.pdf}
+	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates -- amino acid alphabet}.
+	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
+Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
+Not that inferred site specific models only partially ameliorate underestimation, see main text.
+}
+	\label{fig:tree_length_supp}
+\end{figure*}
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.010}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.010}.pdf}
 	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
@@
 	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
-	\label{fig:HIV_fitness}
+	\label{fig:HIV_fitness_supp}
 \end{figure*}
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.001}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.001_aa}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010_aa}.pdf}
 	\caption{{\bf Diversity and GTR rate estimates.}
 	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs the diversity in columns of the corresponding alignment.}
-	\label{fig:HIV_diversity}
+	\label{fig:HIV_diversity_supp}
 \end{figure*}

### 2019-12-28 src/HIV_branch_length.py

@@ -19,39 +19,40 @@ if __name__ == '__main__':
     parser.add_argument("--pc", type=float, default=1.0, help="pseudocount for reconstruction")
     args=parser.parse_args()
 
+    # load model computed by script "analyze_HIV_tree.py"
     alphabet='aa' if args.aa else 'nuc_nogap'
     pc = args.pc
     fs = 16
     dset_name=os.path.basename(args.prefix)
-    model_name = hiv_out + dset_name+'_aligned.%1.2f_inferred_model%s.npz'%(args.pc, '_aa' if args.aa else '')
+    model_name = hiv_out + dset_name+'_%saligned.%1.2f_inferred_model%s.npz'%('aa-' if args.aa else '', args.pc, '_aa' if args.aa else '')
+    gtr = load_model(model_name)
 
-    gtr=None
-    aln_freq=None
-    try:
-        gtr = load_model(model_name)
-    except:
-        pass
-
-    #gtr.Pi = gtr.Pi.mean(axis=1)
     S = -np.sum(np.log(gtr.Pi)*gtr.Pi, axis=0)
     print('Entropy:', S.mean(), S.std())
+
+    # make a replica of the model with frequencies averaged over the sequence
     gtr.mu /= gtr.average_rate().mean()
     gtr_flat = GTR_site_specific.custom(alphabet=gtr.alphabet,  mu=gtr.mu, W=gtr.W, pi=gtr.Pi.mean(axis=1))
     gtr_flat.mu /= gtr_flat.average_rate().mean()
 
-    t=0.1
-    s1 = prof2seq(gtr.Pi.T, gtr, sample_from_prof=True)
-    s1_prof = seq2prof(s1[0], profile_map=gtr.profile_map)
 
     branch_length = []
-    for r in range(20):
+    for r in range(20): # repeat 20 times
+        # sample a random sequence from the true profile
+        s1 = prof2seq(gtr.Pi.T, gtr, sample_from_prof=True)
+        s1_prof = seq2prof(s1[0], profile_map=gtr.profile_map)
         tmp_branch_length = []
         for t in np.linspace(0,1,31):
+            # evolve the ancestral sequence for time t
             s2_prof = gtr.evolve(s1_prof, t)
             s2 = prof2seq(s2_prof, gtr, sample_from_prof=True)
             s2_prof_q = seq2prof(s2[0], profile_map=gtr.profile_map)
[…]

### 2019-12-28 src/analyze_HIV_tree.py

@@ -43,7 +43,7 @@ def calculate_in_out_ratio(gtr, major_allele):
     '''
     major_index = np.zeros_like(major_allele, dtype=int)
     for ni, n in enumerate(gtr.alphabet):
-        major_index[major_allele==n.decode()]=ni
+        major_index[major_allele==n]=ni
     #major_index = np.argmax(gtr.Pi, axis=0)
 
     r_ind = np.arange(gtr.Pi.shape[-1])

### 2019-12-28 src/collect_tree_lengths.py

@@ -19,13 +19,18 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser(description = "", usage="collect statistics of re-optimized trees")
     parser.add_argument("--reoptimizedTrees", type=str, help="directory with reoptimized trees")
     parser.add_argument("--trueTrees", type=str, help="directory with true trees")
+    parser.add_argument("--nval", type=int, help="tree size to glob")
     parser.add_argument("--true-rates", action='store_true', help="include estimates with true rates")
     parser.add_argument("--output", type=str, help="file to save data")
     args=parser.parse_args()
 
     res = defaultdict(dict)
-    reoptimizedTrees = glob.glob(args.reoptimizedTrees+'/*reoptimized.nwk')
-    recTrees = glob.glob(args.trueTrees+'/*reconstructed.nwk')
+    if args.nval:
+        reoptimizedTrees = glob.glob(args.reoptimizedTrees+f'/*n{args.nval}_*reoptimized.nwk')
+        recTrees = glob.glob(args.trueTrees+f'/*n{args.nval}_*reconstructed.nwk')
+    else:
+        reoptimizedTrees = glob.glob(args.reoptimizedTrees+'/*reoptimized.nwk')
+        recTrees = glob.glob(args.trueTrees+'/*reconstructed.nwk')
 
     for treename in reoptimizedTrees:
         dset = parse_alignment_name(treename)

### 2019-12-28 src/compare_tree_length.py

@@ -6,65 +6,65 @@ if __name__ == '__main__':
     import argparse
     parser = argparse.ArgumentParser(description = "", usage="graph inferred tree lengths")
     parser.add_argument("--treelengths", type=str, help="simulated mutation rate")
+    parser.add_argument("--nval", type=int, help="data set size")
+    parser.add_argument("--aa", action='store_true', help="inference from amino acid sequences")
     args=parser.parse_args()
 
     df = pd.read_csv(args.treelengths, sep='\t')
 
     # make scatter plots comparing true to inferred tree length
-    n = 100
+    n = args.nval
     n_branch = 2*n-2
 
-    fig = plt.figure(1)
+    fs = 16
+    fig = plt.figure()
     df_slice = df.loc[df['n']==n,:]
     true_val = df_slice['trueTree_treelength']/n_branch
-    plt.scatter(true_val, df_slice['reconstructed_treelength']/n_branch)
-    plt.scatter(true_val, df_slice['inferred_c0.1_treelength']/n_branch)
-    plt.scatter(true_val, df_slice['inferred_c1.0_treelength']/n_branch)
-    plt.scatter(true_val, df_slice['trueRates_c0.1_treelength']/n_branch)
-    plt.scatter(true_val, df_slice['trueModel_treelength']/n_branch)
+    plt.scatter(true_val, df_slice['trueModel_treelength']/n_branch, label='true model')
+    plt.scatter(true_val, df_slice['reconstructed_treelength']/n_branch, label="FastTree" if args.aa else "IQ-tree GTR+10")
+    plt.scatter(true_val, df_slice['inferred_c0.1_treelength']/n_branch, label='c=0.1')
+    plt.scatter(true_val, df_slice['inferred_c1.0_treelength']/n_branch, label='c=1.0')
+    plt.scatter(true_val, df_slice['trueRates_c0.1_treelength']/n_branch, label='c=0.1 (true rates)')
+#    plt.scatter(true_val, df_slice['trueRates_c1.0_treelength']/n_branch, label='c=1.0 (true rates)')
     plt.plot([0,.3], [0,.3])
+    plt.ylabel('Inferred avg branch length', fontsize=fs)
+    plt.xlabel('True avg branch length', fontsize=fs)
+    plt.tick_params(labelsize=0.8*fs)
[…]

### 2019-12-28 src/generate_toy_data.py

@@ -72,7 +72,7 @@ def simplex(params, out_prefix = None, yule=True, n_model = 5, n_seqgen=5,
     n_seqgen : int, optional
         number of times sequences are evolved for each tree/model combination
     JC : bool, optional
-        Use a Jukes Cantor model
+        Use a Jukes Cantor model for the preference but include rate variation
     alphabet : str, optional
         alphabet of the GTR model
     alpha : float, optional
@@ -203,7 +203,7 @@ if __name__ == '__main__':
     parser.add_argument("-m", type=float, help="simulated mutation rate")
     parser.add_argument("-n", type=int, help="number of taxa")
     parser.add_argument("-L", type=int, default=300, help="length of sequence")
-    parser.add_argument("--JC", action='store_true', help="simulate JC model")
+    parser.add_argument("--JC", action='store_true', help="simulate JC model with rate variation")
     parser.add_argument("--aa", action='store_true', help="use amino acid alphabet")
     parser.add_argument("--alpha", default=1.0, type=float,  help="parameter of the dirichlet distribution for preferences")
     parser.add_argument("--rate-alpha", default=1.5, type=float,  help="parameter of the gamma distribution for rates")

### 2019-12-28 src/model_deviation.py

@@ -1,3 +1,8 @@
+'''
+script that reads in data obtained from mixing Jukes-Cantor type models with the true
+model to explore how sensitive reconstruction of branch length is to model
+mis-specification.
+'''
 import glob
 from matplotlib import pyplot as plt
 from Bio import Phylo, AlignIO
@@ -29,7 +34,7 @@ def make_means(data, conditions):
 if __name__ == '__main__':
     import argparse
     parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
-    parser.add_argument("--files", nargs="+", type=str, help="alignments to analyze")
+    parser.add_argument("--files", nargs="+", type=str, help="tables to ")
     parser.add_argument("--plot-mu", type=float, default=0.2, help="mutation rate used for the plot vs epsilong")
     parser.add_argument("--output", type=str, help="file to save figure to")
     args=parser.parse_args()
@@ -37,7 +42,7 @@ if __name__ == '__main__':
     df = load_data(args.files)
     eps_vals = list(np.linspace(0.0,1,11))
     mu_vals = list(np.unique(df['m']))
-    
+
     one_mu = make_subset(df, {"m":args.plot_mu})
     eps_graph = {x:[list() for e in eps_vals] for x in ['pi', 'mu', 'all']}
     mu_graph = {x:[list() for mu in mu_vals] for x in ['pi', 'mu', 'all']}
@@ -52,9 +57,9 @@ if __name__ == '__main__':
         for x in ['pi', 'mu', 'all']:
             mu_graph[x][mi].append(row[f'eps_{x}_1.00']/row['true_value'])
 
-                
+
     # make figure
-    fig, axs = plt.subplots(1,2, sharey=True, figsize=(13,7))
+    fig, axs = plt.subplots(1,2, sharey=True, figsize=(11,5))
 
     fs=16
     for k, label_str in [('pi', 'frequencies'), ('mu', 'rates'), ('all', 'both')]:
@@ -62,7 +67,7 @@ if __name__ == '__main__':
         std = np.std(eps_graph[k],axis=1)
         axs[0].errorbar(eps_vals, m, std, label=label_str, lw=3)
 
-    axs[0].set_xlabel('mixing ratio true/flat', fontsize=fs)
+    axs[0].set_xlabel(r'mixing ratio $\alpha$ true/flat', fontsize=fs)
[…]

### 2019-12-28 src/model_deviation_scan_mu.py

@@ -1,0 +1,157 @@
+import glob
+from matplotlib import pyplot as plt
+from Bio import Phylo, AlignIO
+from treetime import TreeAnc
+from treetime.gtr_site_specific import GTR_site_specific
+
+
+from generate_toy_data import *
+from filenames import *
+from estimation import *
+from reconstruct_toy_data import *
+
+
+def run_tree(fname, out_prefix, alphabet, model, n_iter=20):
+    params = parse_alignment_name(fname)
+    m = params['m']
+    tree = Phylo.read(tree_name(prefix, params), 'newick')
+    tree.root.branch_length = 0.001
+    tree.ladderize()
+    for n in tree.find_clades():
+        n.branch_length *= m*(0.6+0.4*np.random.random())
+
+    with gzip.open(alignment_name(prefix, params), 'rt') as fh:
+        aln = AlignIO.read(fh, 'fasta')
+
+    tt = TreeAnc(tree=tree, aln=aln, gtr = model, compress=False,
+                 alphabet=alphabet, verbose=3)
+
+    tt.optimize_tree(branch_length_mode='marginal', max_iter=n_iter,
+                     infer_gtr=False)
+
+    return tt
+
+if __name__ == '__main__':
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument("--files", nargs="+", type=str, help="alignments to analyze")
+    parser.add_argument("--output", type=str, help="folder to save data")
+    parser.add_argument("--aa", action='store_true', help="assume amino acid alphabet")
+    args=parser.parse_args()
+
+    try:
+        if not os.path.isdir(args.output):
+            os.mkdir(args.output)
+    except:
+        pass
+
+
+    alphabet = 'aa_nogap' if args.aa else 'nuc_nogap'
+    res = {}
+    eps_vals = list(np.linspace(0.0,1,11))
+
+    for fname in args.files:
+        print(fname)
+        params = parse_alignment_name(fname)
+        res[fname] = params
+
+        true_model = load_model(model_name(prefix, params))
+        n = true_model.n_states
+
+        tree = Phylo.read(tree_name(prefix, params), 'newick')
+        true_ttl = tree.total_branch_length()*params['m']
+        res[fname]['true_value'] = true_ttl
+
+        for eps in eps_vals:
[…]

### 2019-12-28 src/plot_toy_data.py

@@ -128,7 +128,7 @@ def plot_pentropy_vs_rtt(data, subsets, fname=None):
         rtt = d['mean'].index*np.log(n)
         plt.plot(rtt, d['mean']['model_entropy']-d['mean']['true_entropy'],
                             c=colors[tmp_p['method']], ls=line_styles[subsets['pc'].index(pc)],
-                            label=f"{labels[label]}, pc={pc:1.2f}" if n==subsets['n'][0] and  label!='naive' else '')
+                            label=f"{labels[label]}, c={pc:1.2f}" if n==subsets['n'][0] and  label!='naive' else '')
 
     plt.xscale('log')
     plt.legend(fontsize=fs)
@@ -166,13 +166,14 @@ def plot_avg_rate(data, subsets, fname=None):
 
 def plot_rate_correlation(data, subsets, fname=None):
     mean_vals = make_means(data, subsets)
-    ls={1.5:'-', 3.0:'--'}
+    ls=['-', '--']
     plt.figure()
     for params, d in mean_vals.items():
         tmp_p = {k:v for k,v in params}
         pc, label, n = tmp_p['pc'], tmp_p['method'], tmp_p['n']
-        plt.errorbar(d['mean'].index, d['mean']['r_mu'],d['std']['r_mu'],
-                     c="C%d"%subsets['pc'].index(pc), label=f'pc={pc:1.2f}' if n==subsets['n'][0] else '')
+        plt.errorbar(d['mean'].index*n, d['mean']['r_mu'],d['std']['r_mu'],
+                     c="C%d"%subsets['pc'].index(pc), label=f'c={pc:1.2f}, n={n}',
+                     ls=ls[subsets['n'].index(n)])
 
     plt.xscale('log')
     plt.ylim(0,1.1)
@@ -180,6 +181,7 @@ def plot_rate_correlation(data, subsets, fname=None):
     plt.tick_params(labelsize=0.8*fs)
     plt.xlabel('average number of substitution per site', fontsize=fs)
     plt.ylabel('inferred/true rate correlation', fontsize=fs)
+    plt.plot([10,max(subsets['n'])/3], [1,1], lw=2, c='#CCCCCC')
     plt.tight_layout()
 
     if fname:
@@ -239,7 +241,7 @@ if __name__ == '__main__':
     plot_avg_rate(data, subsets = {'pc':[pc_general], 'method':['naive', 'dressed'], 'n':n_vals_to_plot},
[…]

### 2019-12-28 submit_branch_length.py

@@ -16,8 +16,7 @@ parser.add_argument("--date", type=str,  help="date prefix")
 parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
 args=parser.parse_args()
 
-dsets = [(1.0, False, False)] #, (0.0, True, False)]
-#dsets = [(0.2, False, True), (0.5, False, True)]
+dsets = [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]
 
 date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
@@ -35,10 +34,11 @@ for pc in [0.01, 0.1, 0.5, 1.0]:
                 if len(files):
                     chunks = make_chunks(files, n=ceil(n/100))
                     for ci,chunk in enumerate(chunks):
-                        cmd = f'sbatch {submit_script} src/calculate_branch_length.py --files {" ".join(chunk)} --output {output} {"--aa" if aa else ""} --pc {pc}'
-                        print(cmd)
-                        if args.submit:
-                            os.system(cmd)
-                            os.system(cmd + ' --true-model')
-                            os.system(cmd + ' --true-rates')
+                        if len(chunk):
+                            cmd = f'sbatch {submit_script} src/calculate_branch_length.py --files {" ".join(chunk)} --output {output} {"--aa" if aa else ""} --pc {pc}'
+                            print(cmd)
+                            if args.submit:
+                                os.system(cmd)
+                                os.system(cmd + ' --true-model')
+                                os.system(cmd + ' --true-rates')
 

### 2019-12-28 submit_collect_tree_lengths.py

@@ -1,0 +1,22 @@
+import os, glob
+from datetime import date
+from itertools import product
+import argparse
+parser = argparse.ArgumentParser(description = "", usage="generate_toy_data")
+parser.add_argument("--nval", type=int,  help="number of taxa")
+parser.add_argument("--submit", action='store_true', help="submit")
+parser.add_argument("--date", type=str,  help="date prefix")
+parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
+args=parser.parse_args()
+L=1000
+dsets = [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]
+
+date_prefix = args.date or date.today().strftime('%Y-%m-%d')
+
+for alpha, JC, aa in dsets:
+    dirname = f'{date_prefix}_simulatedData_L{L}_ratealpha{args.rate_alpha:1.1f}_freqalpha{alpha:1.1f}_{"aa" if aa else "nuc"}' + ("_JC" if JC else "")
+    output = dirname + f"_results/collected_tree_lengths_n{args.nval}.tsv"
+    cmd = f"sbatch submit_small.sh src/collect_tree_lengths.py --output {output} --trueTrees {dirname} --reoptimizedTrees {dirname+'_results'}  --nval {args.nval}"
+    print(cmd)
+    if args.submit:
+        os.system(cmd)

### 2019-12-28 submit_model_deviation.py

@@ -1,30 +1,33 @@
-import os
+import os, glob
 from datetime import date
-from itertools import product
 
-L=1000
+if __name__ == '__main__':
 
-import argparse
-parser = argparse.ArgumentParser(description = "", usage="generate_toy_data")
-parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
-parser.add_argument("--muvals", nargs='+', type=float,  help="mutation rate for panelB")
-parser.add_argument("--submit", action='store_true', help="submit")
-parser.add_argument("--date", type=str,  help="date prefix")
-parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
+    L=1000
+    muvals =  [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]
 
-args=parser.parse_args()
-aa = True
-JC = False
-alpha=0.2
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="generate_toy_data")
+    parser.add_argument("--nvals", nargs='+', type=int,  help="number of taxa")
+    parser.add_argument("--muvals", nargs='+', default=muvals, type=float,  help="number of taxa")
+    parser.add_argument("--submit", action='store_true', help="submit")
+    parser.add_argument("--date", type=str,  help="date prefix")
+    parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
+    args=parser.parse_args()
 
-date_prefix = args.date or date.today().strftime('%Y-%m-%d')
+    date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
-prefix = '%s_simulatedData_L%d_ratealpha%1.1f_alpha%1.1f_%s'%(date_prefix, L, args.rate_alpha, alpha, "aa" if aa else "nuc") + ("_JC" if JC else "")
-for n in args.nvals:
-    submit_script = "submit.sh"
-    for mu in args.muvals:
-        cmd = f"sbatch {submit_script} src/model_deviation.py -L {L} -n {n} -m {mu} --prefix {prefix}"
-        print(cmd)
-        if args.submit:
-            os.system(cmd)
+    for alpha, JC, aa in [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]:
[…]

### 2019-12-29 manuscript/ms.tex
edits to manuscript, added additional references

@@
 With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
-Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on model deviations.
+Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site specific model parameters are known.
 
@@
 	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences per tree.
-	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
+	The estimates are more accurate than naive estimates from the state frequencies in alignment columns.
+The advantage over naive estimates increases with data set size as seem from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
 	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
 Regularization via pseudo-counts reduces over-fitting at low divergence.
+Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.
 }
@@
 \section*{Efficient inference of site specific substitution models}
-Following previous authors \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
+Following previous \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
 the overall rates and the equilibrium frequencies depend on the site.
@@
 To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
-This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
+This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
+These update rules are an example of non-negative factorization algorithms \citep{lee_algorithms_2001}.
 
@@
 \section*{Accuracy of inferences}
-The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and the accuracy of the tree reconstruction.
-In addition, the most likely model parameters will deviate from the true parameters due to over-fitting.
+The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and (iii) the accuracy of the tree reconstruction.
+In addition, the Maximum Likelihood estimates might deviate from the true parameters due to over-fitting.
 
-To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites.
+To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
 We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
@@
 	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
-	The left and right panels show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
+	Panels A and B show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
 	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
 	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
-	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=3$.}
+	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.}
 	\label{fig:reconstructed}
@@
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-This scaling becomes more accurate with less rate variation.
-At the largest substitution rates, branch length are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
+Deviations of the scaling $\chi_2\sim 1/\textrm{tree length}$ are due to rate variation.
+The data shown in Fig.~\ref{fig:reconstructed}, assumes that rates are Gamma distributed with $\alpha=1.5$ and variation in inference accuracy due to rate variation increases the average error.
+Increasing $\alpha$ to 3 improves this scaling.
+A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).
+
+At the largest substitution rates, branch lengths are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
 Nevertheless, the accuracy of the $\hat{\eqp}_i^a$ continues to increase.
-By contrast, using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry.
-
-The accuracy of substitution rates is affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
+While equilibrium frequencies are insensitive, the substitution rate estimates are affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
 This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
@@
 
-\subsection*{Ignoring site specific frequencies results in dramatic branch length underestimates}
+\subsection*{Ignoring site specific frequencies results in branch length underestimates}
 As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
@@
 %% dependence on pseudo count
-Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted tree (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
-While branch length are accurately estimated when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
-Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
+Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
+While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR=R10 model without site specificity.
+Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are prone to underestimation.
 When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation}.pdf}
-	\caption{{\bf Moderate model inaccuracies result in large substantial branch length underestimation}.
+	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_n300}.pdf}
+	\caption{{\bf Sensitivity of branch length estimates on model misspecification}.
+	Panels A \& B show the relative error in total tree length when using a mixture model as defined in Eq.~\ref{eq:mixture_model} for branch length inference.
+Panel A shows this error as a function of the mixing fraction $\alpha$ for $\langle \mu \rangle = 0.2$.
+Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\alpha=1$.
+	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
 	}
@@
 Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
-The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.
-%If, however, preferences are estimated poorly for example through excessive regularization ($\pc=10$ in Fig.~\ref{fig:tree_length}), using the correct rates hardly improves the estimate.
-%The latter is consistent with the observation that standard ML tree inference using IQ-tree with a GTR+R10 model severely underestimates branch lengths.
+The principle contributor to this deviation are inaccuracies in the rate estimates.
+Combining the true rates with inferred preferences reduces the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.
 
 \subsection*{Minor model deviations can have substantial effects on divergence estimates}
-In the previous section, we have observed that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
-To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows.
+In the previous section, we have observed that model deviations, for example the error made during model inference, can result in substantial errors in branch length estimates.
+To investigate this effect more systematically, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows:
 For a mixing fraction $\alpha$, we constructed a model with rates
 \begin{equation}
+\label{eq:mixture_model}
 	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
@@
 Deviations in rate and preferences are approximately additive.
-These observations are consistent with the finding above using inferred preferences and true substitution rates result in much more accurate branch length estimates than inferring both rates and preferences.
+These observations are consistent with the finding above using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.
 
@@
 For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
-This large sample of moderately diverged sequences is a unique opportunity to apply our iterative inference frame work.
-We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level.
+This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
+We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
 
 One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
-Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as
+Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
 \begin{equation}
@@
 The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
-We compare this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
-However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observe a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
+We compared this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
+However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observed a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
 see Fig.~\ref{fig:HIV_fitness})
-This relationship explains about half the variance of nucleotide effects and slightly less of amino acid effects.
+This relationship explains about half the variance of nucleotide effects and about one third of amino acid effects.
 
@@
 points towards different process that drive cross-sectional and with-in host diversification.
-While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectionally.
+While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectional variation.
 Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
-The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_fitness}).
+The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
 
@@
 
-Next, we quantified the effect that ignoring site-specific preferences has on branch length estimates.
-We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with the average preferences and the same per site rate variation.
+Next, we quantified the effect of ignoring site-specific preferences on branch length estimates.
+We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with homogeneous equilibrium frequencies set to the average frequencies across sites while maintaining site-specific rate variation.
 As soon as the length of the simulated branch approaches 0.2, the length inferred by a model without site-specific preferences deviates substantially from the true value and saturates around 0.5 for larger and larger $t$.
 As expected, these deviations are even more severe when estimating branch length as simple sequence divergence (p-distance), while using the generating model reproduces the correct branch length.
+
+Assuming a typical evolutionary rate of HIV of 0.002 changes per site and year, this analysis suggests that length estimates of branches longer than 100 years start to become inaccurate.
+Furthermore, this analysis suggests very little signal to estimate the length of branches that are longer than 300y.
 
@@
 For comparison, the p-distance between the two simulated sequences is also shown.
-The top axis shows branch length in units of years assuming a substitution rate of 0.002/year.
+The top axis shows branch length in units of years assuming a substitution rate of 0.002/year and site.
 Error bars denote one standard deviation.
       Panel B shows the distribution of rates across sites for both models.
-They differ slightly in the overall scale since rates have been rescaled such that the average substitution rate in both models is identical.}
+      About 20\% of sites are essentially invariable, while the rates of the remainder vary by at least 10fold.
+      The distributions differ slightly in the overall scale since rates have been rescaled such that the average substitution rate in both models is identical.}
 	\label{fig:HIV_branch_length}
@@
 
-Here, we introduce and implement a method to estimate site specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
-We show that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
-For short branch length a parsimonious ancestral reconstruction is sufficient, while for more diverged samples and iterative inference of the model, the ancestral states, and the branch lengths is necessary.
+Here, we presented method to estimate site specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
+We showed that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
+For short branch lengths a parsimonious ancestral reconstruction is sufficient, while for more diverged samples iterative inference of the model, the ancestral states, and the branch lengths is necessary.
 The computational complexity of the inference is dominated by ancestral reconstruction and branch length optimization, which is quadratic in the alphabet size and linear in sample size and sequence length \citep{felsenstein2004inferring}.
@@
 While using the correct model results in correct branch length estimates, joint inference of site-specific models and branch lengths is typically unable to recover the true branch lengths and substantial underestimation remains.
+This points to parameter identifiability problems rooted in similar effects of skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
+
+Branch length estimates using models constructed by mixing the true model and flat Jukes-Cantor type models showed that misspecified equilibrium frequencies result in substantial errors as soon as the model deviates from the true model by 30\% or more.
 This mirrors observations by \citet{hilton_modeling_2018}, who found that preferences measured for influenza HA proteins of type H1 or H3 affect branch length of in the vicinity of the focal sequence, but not globally on the tree.
 Using site-specific models inferred from HIV alignments, we quantified the error made when ignoring site-specific preferences.
-While the true models are unknown in this case, this analysis nevertheless suggests that errors are substantial as soon as branch length exceed $t=0.2$ and sequence divergence saturates at levels around 0.3.
-This result is consistent with the observation of frequent reversion of HIV to a preferred sequence state \citep{zanini2015population,carlson_selection_2014,leslie_hiv_2004}.
-Average nucleotide distances between \gene{pol} sequences between HIV-1 subtypes are on the order of 0.1, HIV-1 and SIVcpz are on the order of 0.25 and between sequences in the HIV-1/HIV-2/SIV compendium alignment about 0.35 (all distances ignore sites with $>20$\% gaps).
+While the true models are unknown in this case, this analysis nevertheless suggests that errors are substantial as soon as branch length exceed $t=0.2$ ($\sim 100$ years) and sequence divergence saturates at levels around 0.3.
+This result is consistent with the discrepancy between molecular clock based or biogeography based estimates of the divergence times of different SIV lineages \citep{worobey_island_2010}:
+Beyond a few hundred years, there is very little signal to estimate branch length -- in particular when the underlying site-specific model parameters need to be estimated themselves.
+This lack of information is further underscored by the average nucleotide distances between \gene{pol} sequences of different HIV and SIV strains:  HIV-1 subtypes differ at about 10\% of sites, distances between pol and HIV-1 and SIVcpz are on the order of 25\% and distances between sequences in the HIV-1/HIV-2/SIV compendium alignment are about 35\% (all distances ignore sites with $>20$\% gaps).
+These observations are compatible with evidence for frequent reversion of HIV to a preferred sequence state follwing immune escape \citep{zanini2015population,carlson_selection_2014,leslie_hiv_2004}.
 
@@
 While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint inference of model and branch lengths remains difficult.
-This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one primary reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
+This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one  reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
 
@@
 
+\subsection*{Implementation}
+We extended our package TreeTime \citep{sagulenko2017treetime} to handle site specific GTR models by adding an additional class \texttt{GTR\_site\_specific} that generalizes the \texttt{GTR} class.
+Since these classes have almost identical inferfaces, they can be used interchangeably in other analysis run by TreeTime.
+Using the new class, TreeTime can generate sequences with site specific evolutionary models and infer these models from sequence data using the algorithms above.
+
+\subsection*{Generation of simulated data}
+To generate models and sequence ensembles for which the ground truth is known, we sampled binary trees with $n=100, 300, 1000, 3000$ leaves and Yule tree branch length statistics using \href{https://github.com/neherlab/betatree}{betatree} \citep{neher2013coalescence} for the range of values of the average substitution rate $\langle \mu\rangle = [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]$.
+
+For each tree, we sampled two site-specific models for sequences of length $L=1000$ from following distribution:
+Site-specific rates $\mu^a$ were sampled iid from a Gamma distribution with parameter $\alpha=1.5$ or $3.0$.
+Site-specific preferences $\eqp_i^a$ (or equilibrium probabilities) were sampled from Dirichlet distributions with parameters $\alpha=1$ for alphabets with $q=4$ states (nucleotides) or $\alpha=0.2$ and $\alpha=0.5$ for alphabets with $q=20$ (amino acids).
+The entries of the transition matrix $W_{ij}$ were sampled from a Dirichlet distribution with $\alpha=2$.
+The average substitution rate of these models was fixed to required $\langle \mu\rangle$.
+In addition, we generated one set of models ($q=4$) with rate variation but uniform $\eqp_i^a$ and $W_{ij}$.
+For each combination of tree and model, two sets of sequences were evolved using the sequence generation function of TreeTime.
+
+In total, this amounts to 80 alignments for each of the data set sizes $n=100,300,1000,3000$ and four different ensembles.
+For each alignment, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for nucleotide and amino-acid sequences, respectively.
+These trees were used to infer site specific models from data, see below.
+The exact workflow is documented in the script \texttt{src/generate\_toy\_data.py} in the associated git repository at \href{https://github.org/neherlab/SiteSpecificGTR}{github.org/neherlab/SiteSpecificGTR}.
+
+\subsection*{Model inference from simulated data}
+Simulated data and trees (reconstructed or true) were read in by TreeTime and models reconstructed using functions of TreeTime to infer models with the different approximations discussed in the text.
+The exact workflow is documented in the script \texttt{src/reconstruct\_toy\_data.py} in the associated git repository at \href{https://github.org/neherlab/SiteSpecificGTR}{github.org/neherlab/SiteSpecificGTR}.
+
 \subsection*{HIV sequence analysis}
@@
 Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
-Not that inferred site specific models only partially ameliorate underestimation, see main text.
+Note that inferred site specific models only partially ameliorate underestimation, see main text.
 }

### 2019-12-29 src/HIV_branch_length.py
added panel labels to figures

@@ -7,6 +7,8 @@ from treetime.seq_utils import seq2prof, profile_maps, alphabets, prof2seq
 from estimation import *
 from generate_toy_data import save_model, load_model
 
+from plot_toy_data import add_panel_label
+
 
 if __name__ == '__main__':
     import argparse
@@ -76,6 +78,7 @@ if __name__ == '__main__':
     axs[0].set_ylabel('inferred branch length', fontsize=fs)
     axs[0].tick_params(labelsize=0.8*fs)
     axs[0].legend(fontsize=0.8*fs)
+    add_panel_label(axs[0],'A',  x_offset=-0.15,y_offset=0.95, fs=fs)
 
     # add a second axis that coresponds roughly to years for HIV
     sec_axx = axs[0].twinx()
@@ -94,6 +97,7 @@ if __name__ == '__main__':
     axs[1].set_ylabel('fraction below rate x', fontsize=fs)
     axs[1].set_xscale('log')
     axs[1].tick_params(labelsize=0.8*fs)
+    add_panel_label(axs[1],'B',  x_offset=-0.15,y_offset=0.95, fs=fs)
     plt.tight_layout()
 
     for fmt in ['.png', '.pdf']:

### 2019-12-29 src/analyze_HIV_tree.py

@@ -131,7 +131,7 @@ if __name__ == '__main__':
     plt.legend(fontsize=fs)
     plt.xlim([3e-5, 3e0])
     plt.ylim([3e-2, 3e3])
-    plt.title('amino acids' if args.aa else 'nucleotides',fontsize=fs)
+    plt.title('B: amino acids' if args.aa else 'A: nucleotides',fontsize=fs*1.2)
     plt.xlabel('intra-host fitness cost estimates', fontsize=fs)
     plt.ylabel(r'GTR based fitness estimates $\Gamma_{in}/\Gamma_{out}$', fontsize=fs)
     plt.xscale('log')

### 2019-12-29 src/compare_tree_length.py

@@ -1,6 +1,7 @@
 import glob
 import pandas as pd
 from matplotlib import pyplot as plt
+from plot_toy_data import add_panel_label
 
 if __name__ == '__main__':
     import argparse
@@ -27,10 +28,14 @@ if __name__ == '__main__':
     plt.scatter(true_val, df_slice['trueRates_c0.1_treelength']/n_branch, label='c=0.1 (true rates)')
 #    plt.scatter(true_val, df_slice['trueRates_c1.0_treelength']/n_branch, label='c=1.0 (true rates)')
     plt.plot([0,.3], [0,.3])
+    plt.xlim([0,.3])
+    plt.ylim([0,.3])
     plt.ylabel('Inferred avg branch length', fontsize=fs)
     plt.xlabel('True avg branch length', fontsize=fs)
     plt.tick_params(labelsize=0.8*fs)
+    add_panel_label(plt.gca(),'A',  x_offset=-0.16, fs=fs)
     plt.legend(fontsize=0.8*fs)
+    plt.tight_layout()
     plt.savefig(f"figures/{'aa' if args.aa else 'nuc'}_length_n{args.nval}.pdf")
 
 
@@ -46,8 +51,12 @@ if __name__ == '__main__':
     plt.ylabel('Inferred avg root-to-tip', fontsize=fs)
     plt.xlabel('True avg root-to-tip distance', fontsize=fs)
     plt.tick_params(labelsize=0.8*fs)
-    plt.plot([0,3], [0,3])
+    plt.plot([0,3.4], [0,3.4])
+    plt.xlim([0,3.4])
+    plt.ylim([0,3.4])
+    add_panel_label(plt.gca(),'B',  x_offset=-0.12, fs=fs)
     plt.legend(fontsize=0.8*fs)
+    plt.tight_layout()
     plt.savefig(f"figures/{'aa' if args.aa else 'nuc'}_depth_n{args.nval}.pdf")
 
 
@@ -63,7 +72,11 @@ if __name__ == '__main__':
     plt.xlabel('True avg terminal length', fontsize=fs)
     plt.tick_params(labelsize=0.8*fs)
     plt.plot([0,.3], [0,.3])
+    plt.xlim([0,.3])
+    plt.ylim([0,.3])
+    add_panel_label(plt.gca(),'C',  x_offset=-0.16, fs=fs)
     plt.legend(fontsize=0.8*fs)
+    plt.tight_layout()
[…]

### 2019-12-29 src/model_deviation.py

@@ -10,7 +10,7 @@ import pandas as pd
 import numpy as np
 from treetime import TreeAnc
 from treetime.gtr_site_specific import GTR_site_specific
-from plot_toy_data import make_subset
+from plot_toy_data import make_subset, add_panel_label
 
 
 def load_data(files):
@@ -67,16 +67,19 @@ if __name__ == '__main__':
         std = np.std(eps_graph[k],axis=1)
         axs[0].errorbar(eps_vals, m, std, label=label_str, lw=3)
 
-    axs[0].set_xlabel(r'mixing ratio $\alpha$ true/flat', fontsize=fs)
-    axs[0].set_ylabel('rel. total branch length error', fontsize=fs)
+    axs[0].set_xlabel(r'mixing ratio $\alpha$ true/flat, $\langle \mu \rangle=$' +str(args.plot_mu), fontsize=fs)
+    axs[0].set_ylabel('rel. tree length error', fontsize=fs)
     axs[0].tick_params(labelsize=fs*0.8)
+    axs[0].set_ylim([0.4,1.1])
+    add_panel_label(axs[0],'A',  x_offset=+0.02,y_offset=0.9, fs=fs)
 
     for k, label_str in [('pi', 'frequencies'), ('mu', 'rates'), ('all', 'both')]:
         m = np.mean(mu_graph[k],axis=1)
         std = np.std(mu_graph[k],axis=1)
         axs[1].errorbar(mu_vals, m, std, label=label_str, lw=3)
-    axs[1].set_xlabel('evolutionary rate', fontsize=fs)
+    axs[1].set_xlabel(r'evolutionary rate $\langle \mu \rangle$ ($\alpha=1$)', fontsize=fs)
     axs[1].legend(fontsize=fs)
     axs[1].tick_params(labelsize=fs*0.8)
+    add_panel_label(axs[1],'B',  x_offset=+0.02, y_offset=0.9, fs=fs)
     plt.tight_layout()
     plt.savefig(args.output)

### 2019-12-29 src/plot_toy_data.py

@@ -4,6 +4,12 @@ import pandas as pd
 from collections import defaultdict
 from matplotlib import pyplot as plt
 
+def add_panel_label(ax, label, x_offset=-0.1, y_offset=0.95, fs=12):
+    '''Add a label letter to a panel'''
+    ax.text(x_offset, y_offset, label,
+            transform=ax.transAxes,
+            fontsize=fs*1.5)
+
 fmts = ['.png', '.pdf']
 fs = 12
 
@@ -79,6 +85,7 @@ def plot_pdist_vs_tree_length(data, subsets, fname=None):
     plt.tick_params(labelsize=0.8*fs)
     plt.xlabel('average number of substitutions per site', fontsize=fs)
     plt.ylabel('squared deviation of $p_i^a$', fontsize=fs)
+    add_panel_label(plt.gca(),'A',  x_offset=-0.12, fs=fs*1.3)
     plt.tight_layout()
 
     if fname:
@@ -103,10 +110,12 @@ def plot_pdist_vs_rtt(data, subsets, fname=None):
 
     plt.yscale('log')
     plt.xscale('log')
-    plt.legend(fontsize=fs*0.8)
+    plt.legend(fontsize=fs*0.8, loc=3)
     plt.tick_params(labelsize=0.8*fs)
     plt.xlabel('root-to-tip distance', fontsize=fs)
     plt.ylabel('squared deviation', fontsize=fs)
+    plt.title('amino acids' if 'aa' in fname else 'nucleotides', fontsize=fs*1.3)
+    add_panel_label(plt.gca(), 'B' if 'aa' in fname else 'A', x_offset=-0.12, fs=fs*1.3)
     plt.tight_layout()
 
     if fname:
@@ -182,6 +191,7 @@ def plot_rate_correlation(data, subsets, fname=None):
     plt.xlabel('average number of substitution per site', fontsize=fs)
     plt.ylabel('inferred/true rate correlation', fontsize=fs)
     plt.plot([10,max(subsets['n'])/3], [1,1], lw=2, c='#CCCCCC')
+    add_panel_label(plt.gca(), 'B', x_offset=-0.12, fs=fs*1.3)
     plt.tight_layout()
 
     if fname:

### 2019-12-30 manuscript/ms.tex
added abstract

@@
 
+\begin{abstract}
+Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
+Some sites evolve close to neutral, while others are highly conserved, allow only specific states, or only change in concert with other sites.
+Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
+Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
+We find that the interative approximate maximum likelihood schemes uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
+We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
+However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems is typically unable to recover the correct branch lengths.
+Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
+\end{abstract}
+
 \maketitle
@@
 	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
+Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, grey lines with $\alpha=3$.
 	The estimates are more accurate than naive estimates from the state frequencies in alignment columns.
@@
 Regularization via pseudo-counts reduces over-fitting at low divergence.
-Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.
 }
@@
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-Deviations of the scaling $\chi_2\sim 1/\textrm{tree length}$ are due to rate variation.
-The data shown in Fig.~\ref{fig:reconstructed}, assumes that rates are Gamma distributed with $\alpha=1.5$ and variation in inference accuracy due to rate variation increases the average error.
-Increasing $\alpha$ to 3 improves this scaling.
+The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distributed with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
+The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is much more evident for $\alpha=3$ than for $\alpha=1.5$ as the average accuracy is sensitive to a small number of sites with low rates at which frequencies are estimated poorly.
 A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/nuc_length_n300}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n300}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/nuc_length_n1000}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n1000}.pdf}
 	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
@@
 Not that inferred site specific models only partially ameliorate underestimation, see main text.
+Parameters: $n=1000$, $\alpha=1.5$.
 }
@@
 
-\begin{figure}[tb]
+\begin{figure*}[tb]
 	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_n300}.pdf}
+	\includegraphics[width=0.7\textwidth]{{../figures/model_deviation_n300}.pdf}
 	\caption{{\bf Sensitivity of branch length estimates on model misspecification}.
@@
 	\label{fig:model_deviation}
-\end{figure}
+\end{figure*}
 
@@
 see Fig.~\ref{fig:HIV_fitness})
+Interestingly, this correlation between the intra-host estimates and cross-sectional estimates decreased as soon as regularization increased above 0.05 and we there for used a weak regularization of $\pc=0.01$.
 This relationship explains about half the variance of nucleotide effects and about one third of amino acid effects.
 
-The fact that this relationship deviates from the expectation
+The fact that the relationship deviates from the expectation
 \begin{equation}
@@
 Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site specific preference is the basis of all common homology search tools.
-While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern1998evolutionary}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
+While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern_evolutionary_1998}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
 However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data set sizes are increasing rapidly such that more complex models can be estimated.
@@
 We further explored how model choice affects the maximum likelihood estimates of branch lengths.
-As has been reported before, branch lengths are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern1998evolutionary,hilton_modeling_2018}.
+As has been reported before, branch lengths are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern_evolutionary_1998,hilton_modeling_2018}.
 While such underestimation often has a moderate effect on the total length of a tree, it can result in substantial underestimation of root-to-tip distance that are dominated by a few long branches close to the root.
@@
 Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
-\label{eq:Qij}
+\label{eq:Qij_supp}
 \end{eqnarray}

### 2019-12-30 src/compare_tree_length.py
added second rate alpha to Fig1

@@ -22,7 +22,7 @@ if __name__ == '__main__':
     df_slice = df.loc[df['n']==n,:]
     true_val = df_slice['trueTree_treelength']/n_branch
     plt.scatter(true_val, df_slice['trueModel_treelength']/n_branch, label='true model')
-    plt.scatter(true_val, df_slice['reconstructed_treelength']/n_branch, label="FastTree" if args.aa else "IQ-tree GTR+10")
+    plt.scatter(true_val, df_slice['reconstructed_treelength']/n_branch, label="FastTree" if args.aa else "IQ-tree GTR+R10")
     plt.scatter(true_val, df_slice['inferred_c0.1_treelength']/n_branch, label='c=0.1')
     plt.scatter(true_val, df_slice['inferred_c1.0_treelength']/n_branch, label='c=1.0')
     plt.scatter(true_val, df_slice['trueRates_c0.1_treelength']/n_branch, label='c=0.1 (true rates)')
@@ -43,7 +43,7 @@ if __name__ == '__main__':
     df_slice = df.loc[df['n']==n,:]
     true_val = df_slice['trueTree_depth']
     plt.scatter(true_val, df_slice['trueModel_depth'], label='true model')
-    plt.scatter(true_val, df_slice['reconstructed_depth'], label="FastTree" if args.aa else "IQ-tree GTR+10")
+    plt.scatter(true_val, df_slice['reconstructed_depth'], label="FastTree" if args.aa else "IQ-tree GTR+R10")
     plt.scatter(true_val, df_slice['inferred_c0.1_depth'], label='c=0.1')
     plt.scatter(true_val, df_slice['inferred_c1.0_depth'], label='c=1.0')
     plt.scatter(true_val, df_slice['trueRates_c0.1_depth'], label='c=0.1 (true rates)')
@@ -64,7 +64,7 @@ if __name__ == '__main__':
     df_slice = df.loc[df['n']==n,:]
     true_val = df_slice['trueTree_terminals']
     plt.scatter(true_val, df_slice['trueModel_terminals'], label='true model')
-    plt.scatter(true_val, df_slice['reconstructed_terminals'], label="FastTree" if args.aa else "IQ-tree GTR+10")
+    plt.scatter(true_val, df_slice['reconstructed_terminals'], label="FastTree" if args.aa else "IQ-tree GTR+R10")
     plt.scatter(true_val, df_slice['inferred_c0.1_terminals'], label='c=0.1')
[…]

### 2019-12-30 src/plot_toy_data.py

@@ -28,7 +28,7 @@ colors = {k:'C%d'%((i+1)%10) for i,k in enumerate(labels.keys())}
 
 
 def load_toy_data_results(path):
-    tsv_files = glob.glob(os.path.join(path, "*tsv"))
+    tsv_files = glob.glob(os.path.join(path, "L*tsv"))
     df = pd.concat([pd.read_csv(fname, sep='\t') for fname in tsv_files])
     return df
 
@@ -56,7 +56,7 @@ def make_means(data, conditions):
     return res
 
 
-def plot_pdist_vs_tree_length(data, subsets, fname=None):
+def plot_pdist_vs_tree_length(data, data_secondary, subsets, fname=None):
     # plot the squared distance of the inferred equilibrium frequencies
     # from the true frequencies as a function of total tree length. we expect this to be 1/n
     # We assume a Yule tree here, that is the total tree length can be calculated as:
@@ -76,6 +76,15 @@ def plot_pdist_vs_tree_length(data, subsets, fname=None):
         if tmp_p['method']=='naive':
             plt.text(subs_per_site[0], d['mean']['chisq_p'].iloc[0]*1.2,
                     f"n={tmp_p['n']}", fontsize=fs*0.8)
+
+    if not (data_secondary is None):
+        mean_vals_secondary = make_means(data_secondary, subsets)
+        for params, d in mean_vals_secondary.items():
+            tmp_p = {k:v for k,v in params}
+            subs_per_site = d['mean'].index*tmp_p['n']
+            plt.errorbar(subs_per_site, d['mean']['chisq_p'], d['std']['chisq_p'],
+                         label='', c="#AAAAAA", ls=ls[tmp_p['n']])
+
 
     plt.plot([10,1000], [0.1,0.001], label=r'$\sim x^{-1}$', c='k')
     plt.ylim([0.001, 1.6])
@@ -228,21 +237,23 @@ if __name__ == '__main__':
     parser.add_argument("--prefix", type=str, help="prefix of data set")
     parser.add_argument("--nvals", nargs='+', type=int, default=[1000], help="n values to plot")
     parser.add_argument("--pc", type=float, default=0.1, help="pc value to use in plots")
-    parser.add_argument("--rate-alpha", type=float, default=1.5, help="rate variation set to use")
     args=parser.parse_args()
 
     aa = 'aa' if '_aa' in args.prefix else 'nuc'
[…]

### 2019-12-30 submit_gendata.py

@@ -17,7 +17,7 @@ today = args.date or date.today().strftime('%Y-%m-%d')
 datasets = [(1.0, 1.5, False, False),
             (0.0, 1.5, True, False),
             (0.2, 1.5, False, True),
-            (0.5, 1.5, False, True)]
+            (0.5, 1.5, False, True),(1.0, 3.0, False, False)]
 
 for alpha, rate_alpha, JC, aa in datasets:
     prefix = f'{today}_simulatedData_L{L}_ratealpha{rate_alpha:1.1f}_freqalpha{alpha:1.1f}_{"aa" if aa else "nuc"}' + ("_JC" if JC else "")

### 2019-12-30 submit_reconstruction.py

@@ -21,14 +21,13 @@ if __name__ == '__main__':
     parser.add_argument("--muvals", nargs='+', default=muvals, type=float,  help="number of taxa")
     parser.add_argument("--submit", action='store_true', help="submit")
     parser.add_argument("--date", type=str,  help="date prefix")
-    parser.add_argument("--rate-alpha", type=float, default=1.5, help="parameter of the rate distribution")
     args=parser.parse_args()
 
     date_prefix = args.date or date.today().strftime('%Y-%m-%d')
 
     for pc in [0.01, 0.1, 0.5, 1.0]:
-        for alpha, JC, aa in [(1.0, False, False), (0.0, True, False), (0.2, False, True), (0.5, False, True)]:
-            dirname = f'{date_prefix}_simulatedData_L{L}_ratealpha{args.rate_alpha:1.1f}_freqalpha{alpha:1.1f}_{"aa" if aa else "nuc"}' + ("_JC" if JC else "")
+        for alpha, rate_alpha, JC, aa in [(1.0, 3.0, False, False)]: #[(1.0, 1.5, False, False), (0.0, 1.5, True, False), (0.2, 1.5, False, True), (0.5, 1.5, False, True)]:
+            dirname = f'{date_prefix}_simulatedData_L{L}_ratealpha{rate_alpha:1.1f}_freqalpha{alpha:1.1f}_{"aa" if aa else "nuc"}' + ("_JC" if JC else "")
 
             for n in args.nvals:
                 submit_script = "submit.sh" if n>200 else "submit_small.sh"

### 2019-12-30 manuscript/ms.tex
edited affiliations

@@
 \title{Potential and limits of site-specific substitution model inference.}
-\author{Vadim Puller$^{1,2,3}$ Pavel Sagulenko$^{1}$, Richard Neher$^{1,2,3}$}
-\affiliation{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany\\
-$^{2}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
-$^{3}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland}
+\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard Neher$^{1,2}$}
+\affiliation{
+$^{1}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
+$^{2}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland\\
+$^{3}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
 

### 2019-12-30 manuscript/ms.tex
corrected github link

@@
 These trees were used to infer site specific models from data, see below.
-The exact workflow is documented in the script \texttt{src/generate\_toy\_data.py} in the associated git repository at \href{https://github.org/neherlab/SiteSpecificGTR}{github.org/neherlab/SiteSpecificGTR}.
+The exact workflow is documented in the script \texttt{src/generate\_toy\_data.py} in the associated git repository at \href{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
 
@@
 Simulated data and trees (reconstructed or true) were read in by TreeTime and models reconstructed using functions of TreeTime to infer models with the different approximations discussed in the text.
-The exact workflow is documented in the script \texttt{src/reconstruct\_toy\_data.py} in the associated git repository at \href{https://github.org/neherlab/SiteSpecificGTR}{github.org/neherlab/SiteSpecificGTR}.
+The exact workflow is documented in the script \texttt{src/reconstruct\_toy\_data.py} in the associated git repository at \href{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
 
@@
 The sequences and trees were then used for site specific GTR inference as implemented in TreeTime \citep{sagulenko2017treetime}.
-The scripts detailing this analysis and producing the figures are available on GitHub in repository \href{https://github.com/neherlab/SiteSpecificGTR}{neherlab/SiteSpecificGTR}.
+The scripts detailing this analysis and producing the figures are available on GitHub in repository \url{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
 

### 2020-01-13 manuscript/ms.tex
fixed typos

@@
 Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
-We find that the interative approximate maximum likelihood schemes uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
+We find that the interative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
 We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
-However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems is typically unable to recover the correct branch lengths.
+However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
 Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
@@
 
-These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the appendix for the other (more laborious) update rules.
+These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the Materials and Methods for the other (more laborious) update rules.
 \begin{equation}

### 2020-01-17 manuscript/ms.tex
added supplement

@@
 \bq
-P(\vec{s}\leftarrow \vec{r},t,\mat{Q}) = \prod_{a=1}^L \left(e^{\mat{Q}^{a} t}\right)_{s_{a},r_{a}}
+P(\vec{s}\leftarrow \vec{r},t,\mat{Q}) = \prod_{a=1}^L \left(e^{\mat{Q}^{a} t}\right)_{s^{a},r^{a}}
 \label{eq:P_prob}
 \eq
-where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $i$, and $s_{a}$ and $r_{a}$ are the sequence states at position $i$.
+where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $i$, and $s^{a}$ and $r^{a}$ are the sequence states at position $i$.
 The product runs over all $L$ sites $a$ and amounts to assuming that different sites of the sequence evolve independently.
 
-In absence of recombination, homologous sequences are related by a tree and the likelihood of observing an alignment $\mat{A}=\{\vec{s^k}, k=1..n\}$ conditional on the tree $T$ and the substitution model $\mat{Q}^{a}$ can be written in terms of propagators defined in Eq.~\ref{eq:P_prob}.
+In absence of recombination, homologous sequences are related by a tree and the likelihood of observing an alignment $\mat{A}=\{\vec{s_k}, k=1..n\}$ conditional on the tree $T$ and the substitution model $\mat{Q}^{a}$ can be written in terms of propagators defined in Eq.~\ref{eq:P_prob}.
 It is helpful to express this likelihood as product of sequence propagators defined in Eq.~\ref{eq:P_prob} between sequences at the ends of each branch in the tree (implicitly assuming that on different branches is independent and follows the same time reversible model).
@@
 \begin{equation}
-	\LH(\mat{A}|T,\mat{Q}) = \sum_{\{\vec{s}'\}}\prod_{a=1}^L \eqp^{a}_{x^{0}_a} \prod_{k\in T}P(\vec{s}_{k_c}\leftarrow \vec{s}_{k_p},t,\mat{Q}) = \sum_{\{\vec{s}\}} e^{\lh(\{\vec{s}\} |T, \mat{Q})}  \ ,
+	\LH(\mat{A}|T,\mat{Q}) = \sum_{\{\vec{s}'\}}\prod_{a=1}^L \eqp^{a}_{s^{a}_0} \prod_{k\in T}P(\vec{s}_c\leftarrow \vec{s}_p,t,\mat{Q}) = \sum_{\{\vec{s}\}} e^{\lh(\{\vec{s}\} |T, \mat{Q})}  \ ,
 	\label{eq:LH}
 \end{equation}
-where $\vec{s}_{k_c}$ and $\vec{s}_{k_p}$ are the child and parent sequences of branch $k$, respectively, and the factor $\prod \eqp^{a}_{x^{0}_a}$ is the product of the probabilities of the root sequence $s^{0}_a$ over all positions $i$.
+where $\vec{s}_c$ and $\vec{s}_p$ are the child and parent sequences of branch $k$, respectively, and the factor $\prod_a \eqp^{a}_{s^{a}_0}$ is the product of the probabilities of the root sequence $s^{a}_0$ over all positions $a$.
 The probabilities $\vec{\eqp}^{a}$ are the equiblirium probabilities of the substitution model at position $a$.
@@
 \begin{equation}
-	e^{\lh(\mat{A}, \{\vec{s}'\} |T, \mat{Q})} = \sum_a \left[\log(\eqp^a_{s^{0}_a}) + \sum_{k\in T} \log\left( e^{Q^at_k} \right)_{k^a_c, k^a_p}\right]
+	\label{eq:logLH}
+	\lh(\mat{A}, \{\vec{s}'\} |T, \mat{Q}) = \sum_a \left[\log(\eqp^a_{s^{a}_0}) + \sum_{k\in T} \log\left( e^{Q^at_k} \right)_{s^a_c, s^a_p}\right]
 \end{equation}
-where $k^a_c$ and $k^a_p$ are indices corresponding to the child and parent sequence of branch $k$.
+where $s^a_c$ and $s^a_p$ are indices corresponding to the child and parent sequence of branch $k$.
 
@@
 Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
-\label{eq:Qij_supp}
+\label{eq:Qij_app}
 \end{eqnarray}
@@
 
-These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the Materials and Methods for the other (more laborious) update rules.
+These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the supplement for the other (more laborious) update rules.
+The derivative of the log-likelihood for a specific sequence assignment is given by
 \begin{equation}
-	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{k\in T} \frac{t_k}{\mu^b}\frac{\sum_i Q^b_{k^b_c,i}\left( e^{Q^bt_k} \right)_{i, k^b_p}}{\left( e^{Q^bt_k} \right)_{k^b_c, k^b_p}}
+	\frac{\partial \lh(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{\beta\in T} \frac{t_\beta}{\mu^b}\frac{\sum_i Q^b_{s^b_c,i}\left( e^{Q^bt_\beta} \right)_{i, s^b_p}}{\left( e^{Q^bt_\beta} \right)_{s^b_c, s^b_p}} \ ,
 \end{equation}
-The individual terms in this sum behave very differently for cases where $k^b_c=k^b_p$ (no change at site $b$ on branch $k$) and $k^b_c\neq k^b_p$ (at least one mutation).
-In the limit of short branches $\mu^b t_k\ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain more insightful expressions and we will separate branches with $k^b_c=k^b_p$  and $k^b_c\neq k^b_p$ right away
+where $s^b_p$ and $s^b_c$ are the states at site $b$ at the parent and child end of branch $\beta$.
+The individual terms in this sum behave very differently for cases where $s^b_c=s^b_p$ (no change at site $b$ on branch $\beta$) and $s^b_c\neq k^b_p$ (at least one mutation).
+In the limit of short branches $\mu^b t_\beta \ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain approximate but solvable conditions for maximum likelihood parameter estimates, see supplement.
+We will separate the sum over branches into those with $s^b_c=s^b_p$  and $s^b_c\neq s^b_p$.
+Suppressing the index $b$ for the position in the sequence, we find
 \begin{equation}
 \begin{split}
-	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} &\approx \sum_{k\in T, k^b_c=k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b,i}(\delta_{i,k^b} + t_k Q^b_{i,k^b}\cdots)}{1 - t_k\mu^b\sum_{j}\eqp_j W_{k^bj}+\ldots} +
-	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b_c,i}(\delta_{i,k^b_p} + t_k Q^b_{i,k^b_p}\cdots)}{t_k\mu^b\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
-		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} +
-	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{p_{k^b_c}W_{k^b_c, l^b_p} + \cdots}{t_k\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
-		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} + 	\sum_{k\in T, k^b_c\neq k^b_p} \frac{1}{\mu^b} \\
-		&= -\sum_{ij}\eqp_{i}W_{i,j}\tau_j^b + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
+	\frac{d}{d \mu} \lh (\{\vec{s}\} |T, \mat{Q}) &
+	 \approx  -\sum_{\beta\in T,s_c=s_p} \frac{t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c}}{1 - t_\beta \mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
+	 + \sum_{\beta\in T,s_c\neq s_p} \frac{t_\beta \eqp_{s_c} W_{s_cs_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
+	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau^b_j + \sum_{i\neq j} n^b_{ij}/\mu
 	\end{split}
@@
 where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
-This condition (and the corresponding ones in the supplement suggest solution at fixed $\tau^b_j$ and $n^b_{ij}$ using the following iterative update:
-\begin{equation}
-\label{eq:update_app}
-	\begin{split}
-		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
-		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
-		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
-	\end{split}
-\end{equation}
-where $\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
-To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
+Additional terms necessary for regularization and normalization are discussed in the supplement.
+Setting this expression to zero (and the corresponding ones in the supplement) suggests solving for $\mu^b$ at fixed $\tau^b_j$ and $n^b_{ij}$ using the iterative update rules given in Eqs.~\ref{eq:update}.
+
+In addition, we introduced a pseudocount $\pc$ analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length, see supplement.
+To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator in Eqs.~\ref{eq:update} with the parameter that is being updated.
 This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.

### 2020-01-17 manuscript/supplement.tex
added supplement

\documentclass[aps,rmp,onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Vadim}[1]{{\color{blue}Vadim: #1}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\pc}{c}
\newcommand{\LH}{\mathcal{L}}
\newcommand{\lh}{\ell}
\pdfinfo{%
  /Title    ()
  /Author   ()
  /Creator  ()
  /Producer ()
  /Subject  ()
  /Keywords ()
}

\begin{document}

\title{Supplement: Potential and limits of site-specific substitution model inference.}
\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard A.~Neher$^{1,2}$}
\affiliation{
$^{1}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
$^{2}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland\\
$^{3}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}

\date{\today}
\maketitle

For a given set of sequences $\{\vec{s_k}\}$ consisting of the alignment at the tips and the ancestral sequences, the log-likelihood is given by
\begin{equation}
	\label{eq:logLH}
	\lh(\{\vec{s}\} |T, \mat{Q}) = \sum_a \left[\log(\eqp^a_{s^{a}_0}) + \sum_{\beta\in T} \log\left( e^{Q^at_\beta} \right)_{s^a_c, s^a_p}\right]
\end{equation}
where $\mat{Q}$ is the site specific substitution model consisting of rates, frequencies, and a symmetric transition matrix
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki} \ .
\label{eq:Qij_supp}
\end{eqnarray}

At the maximum, the derivatives of $\lh$ with respect to each of the model parameters should vanish when averaged over all internal site assignments.
The derivative with respect to $\mu^b$ was is presented in Eq.~12 of the main text.
Here, we present the corresponding derivations for $\eqp^b_n$ and $W_{nm}$.

To facilitate the algebra below, we summarize the short branch length assumption and the associated linear approximations here:
\begin{equation}
	\left( e^{Q t}\right)_{ij} = \delta_{i,j} + \sum_{m=1}^\infty \frac{t^m}{m!}\left( Q^m\right)_{ij} \approx
	\begin{cases}
	1 - t\mu^b \sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
	t\mu^b \eqp^b_i W_{ik} & i\neq j
	\end{cases}
\end{equation}
The derivatives of $\left( e^{Q t}\right)_{ij}$ with respect to the different model parameters are therefore
\begin{eqnarray}
	\frac{d}{d\mu^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
	t \eqp^b_i W_{ik} & i\neq j
	\end{cases} \\
	\frac{d}{d\eqp_n^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\mu^b (1-\delta_{i,n})W_{ni} & i=j \\
	 t\mu^b \delta_{i,n}W_{nk} & i\neq j
	\end{cases}\\
	\frac{d}{dW_{mn}} \left( e^{Q^a t}\right)_{ij} &\approx &
	\begin{cases}
	-\delta_{n,j}(1-\delta_{m,i})t \mu^a \eqp^a_m  & i=j \\
	 \delta_{m,i}\delta_{n,j} t\mu^a \eqp^a_m & i\neq j
	\end{cases}
\end{eqnarray}
Since the matrix $W_{ij}$ is symmetric, the derivative with respect to $W_{ij}$ and $W_{ji}$ are the same.
We will account for this symmetry below explicitly and refrain from carrying the symmetric term through lengthy derivations.
Note that the derivatives with respect to $W_{ii}$ vanish as required.

\section*{Update rule for rates $\mu^b$}
Only site $b$ will contribute to the derivative of $\lh$ with respect to $\mu^b$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
As discussed in the main text, we will separate the sum over branches $\beta$ into those where sequence is the same of parent and child ($s_c=s_p$) and those where the sequence differs ($s_c\neq s_p$).
\begin{equation}
\begin{split}
	\frac{d}{d \mu} \lh (\{\vec{s}\} |T, \mat{Q}) &
	 \approx  -\sum_{\beta\in T,s_c=s_p} \frac{t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c}}{1 - t_\beta \mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \frac{t_\beta \eqp_{s_c} W_{s_cs_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau_j + \sum_{i\neq j} n_{ij}/\mu
\end{split}
\end{equation}
Here, $\tau_j$ is the sum of all branch lengths on which $s_c=j$ and $n_{ij}$ is the number of times the sequence changes from $j$ to $i$ somewhere along the tree.
Setting this expression to 0 and re-instantiating the site index $b$, we get
\begin{equation}
	\mu^b = \frac{\sum_{i\neq j} n^b_{ij}}{\sum_{j, k\neq j} \eqp^b_k W_{kj} \tau^b_j}
\end{equation}

\section*{Update rule for frequencies $\eqp_i^b$}
As above, only site $b$ will contribute to the derivative of $\lh$ with respect to $\eqp^b_n$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
\begin{equation}
\begin{split}
	\frac{d}{d \eqp_n} \lh (\{\vec{s}\} |T, \mat{Q}) & \approx \frac{\delta_{s_0,n}}{\eqp_n}
	-\sum_{\beta\in T,s_c=s_p} \frac{t_\beta\mu (1-\delta_{n,s_c})W_{n s_c}}{1 - t_\beta\mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \delta_{n,s_c}\frac{t_\beta\mu W_{ns_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{\beta\in T,s_c=s_p} t_\beta\mu (1-\delta_{s_c, n})W_{s_c n}
	 + \sum_{\beta\in T,s_c=n, s_c\neq s_p} \frac{1}{\eqp_{n}}\\
	 & = 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{j\neq n} \tau_j \mu W_{j n}
	 + \sum_{j\neq n}\frac{n_{nj}}{\eqp_{n}}
\end{split}
\end{equation}
Reinstantiating the position index $b$, we find
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{\sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}

\section*{Update rule for the transition matrix $W_{ij}$}
Since $W_{ij}$ does not depend on the position, the derivative of $\lh$ with respect to $W_{ij}$ depends on all sites in the alignment.
\begin{equation}
\begin{split}
	\frac{d}{d W_{mn}} \lh (\{\vec{s}\} |T, \mat{Q}) & \approx
	-\sum_a\sum_{\beta\in T,s^a_c=s^a_p} \frac{\delta_{n,s^a_p}(1-\delta_{m,s^a_c})t_\beta \mu^a \eqp^a_m}{1 - t_\beta\mu \sum_{k\neq s^a_c} \eqp_k W_{ks^a_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \delta_{m,s_c}\delta_{n,s_p} \frac{t_\beta \mu^a \eqp^a_m}{t_\beta \mu^a \eqp^a_{m} W_{m n}}\\
	 &\approx -\sum_a\sum_{\beta\in T,s^a_c=s^a_p}  \delta_{n,s^a_p}(1-\delta_{m,s^a_c})t_\beta \mu^a \eqp^a_m + \frac{1}{W_{mn}}\sum_a n^a_{mn}\\
	 & = \sum_a \tau^{a}_{n} \mu^a \eqp^a_m  + \frac{1}{W_{mn}}\sum_a n^a_{mn}
\end{split}
\end{equation}
Identifying $W_{nm}$ and $W_{mn}$ then results in the following symmetric expression
\begin{equation}
	\frac{d}{d W_{mn}} \lh (\{\vec{s}\} |T, \mat{Q}) \approx \sum_a \mu^a (\tau^{a}_{n} \eqp^a_m + \tau^{a}_{m} \eqp^a_n)  + \frac{1}{W_{mn}}\sum_a (n^a_{mn}+ n^a_{nm})
\end{equation}
Setting this expression to 0 and solving for $W_{ij}$ we find
\begin{equation}
	W_{mn} = \frac{\sum_a n^a_{mn}+ n^a_{nm}}{\sum_a \mu^a(\tau^{a}_{n} \eqp^a_m + \tau^{a}_{m} \eqp^a_n)}
\end{equation}

\section*{Implementation of the constraint $\sum_i \eqp_i^b=1$}
Equilibrium frequencies need to be normalized to one and constraints of this nature are typically accounted for by Lagrange parameters
\begin{equation}
	\frac{d}{d \eqp_n} \left[\lh (\{\vec{s}\} |T, \mat{Q}) + \lambda^b \sum_i \eqp^b_i \right] = 0
\end{equation}
which results in a modified extremal condition
\begin{equation}
	0 = \delta_{s^b_0,n}-\eqp_{n}\sum_{j\neq n} \tau^b_j \mu^b W_{j n} + \sum_{j\neq n} n^b_{nj} + \eqp^b_{n}\lambda^b
\end{equation}
Summing this condition over $n$, we find
\begin{equation}
	0 = 1 - \sum_{i,j; i \neq n}  \mu^b \tau^b_j \eqp^b_{i}W_{j i} + \sum_{i,j; i\neq j} n^b_{ij} + \lambda^b
\end{equation}
At the extremum of $\mu^v$, we have $\sum_{i,j; i \neq n} \mu^b \tau^b_j \eqp^b_{i}W_{j i} = \sum_{i,j; i\neq j} n_{ij}$ and therefore $\lambda^b=-1$.
With the constraint $\sum_i \eqp_i^b=1$, the update rule therefore becomes
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{1 + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}
In practice it is often helpful to explicitly enforce normalization after each iteration.

\section*{Regularization}
In absence of data, i.e., when there are very few mutations at a particular site we need to regularize the problem.
To this end, we assume a Dirichlet prior of the form
\begin{equation}
\prod_i \left(\eqp^b_i\right)^c
\end{equation}
The derivative of the logarithm of this is $c/\eqp^b_i$ and adding this term to the update rule results in
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0} + c +\sum_j n^b_{nj}}{1 + qc + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}
where $q$ is the size of the alphabet.
In an analogous fashion, we add a weak regularization to the update rules of $\mu^b$ and $W_{ij}$.

\section*{Summation over unobserved states}
So far, we have derived conditions that maximize the likelihood as a function of parameters of the substitution model for a fixed assignment of sequences at internal nodes.
In practice, the internal nodes are unknown and can either be fixed to the most-likely state or summed over.
Since the approximate maximum likelihood conditions are linear, this average over unknown ancestral states simply amounts to replacing $\tau^b_i$ and $n^b_{ij}$ by their averages which can be efficiently computed by standard recursive algorithms.

\end{document}

### 2020-01-18 manuscript/ms.tex
typos and streamlining. Move supp figures to supplement

@@
 \usepackage{color}
+\usepackage{xr}
+\usepackage{hyperref}
+\externaldocument{supplement}
 %\usepackage{cite}
@@
 Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
-Some sites evolve close to neutral, while others are highly conserved, allow only specific states, or only change in concert with other sites.
+Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
 Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
 Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
-We find that the interative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
-We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
+We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site-specific preferences from large data sets with moderately diverged sequences.
+Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
 However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
-Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
+Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
+Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
+Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
 \end{abstract}
@@
 \section*{Introduction}
-Over time, genome sequences change through mutations and are reshuffeled by recombination.
+Over time, genome sequences change through mutations and are reshuffled by recombination.
 Modifications to the genomes are filtered by selection for survival such that beneficial variants spread preferentially and those that impair function are purged.
@@
 Some positions in a protein, for example, might only allow for hydrophobic amino acids, while others require acidic side chains.
-Patterns of conservation and variation, possibly involving more than one site, are therefore shaped by functional contraints which in turn allows inference of biological function from genetic variation.
+Patterns of conservation and variation, possibly involving more than one site, are therefore shaped by functional constraints which in turn allows inference of biological function from genetic variation.
 
@@
 Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
-To avoid over-fitting, these methods typically don't estimate rates for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple uni-model distributions \citep{mayrose2005gamma}, or a small number of fixed rates).
+To avoid over-fitting, these methods typically don't estimate a rate for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple unimodal distributions \citep{mayrose2005gamma}, or a small number of fixed rates).
 
 In addition to rate variation, different sites in a protein differ in amino acid they allow.
-Recent deep mutational scanning experiments have shown that site specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
-Using such experimentally inferred site specific models in phylogenetic inference greatly increases the likelihood of the data \cite{bloom2014experimentally}.
+Recent deep mutational scanning experiments have shown that site-specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
+Using such experimentally inferred site-specific models in phylogenetic inference greatly increases the likelihood of the data \cite{bloom2014experimentally}.
 More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
@@
 In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
-With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
-Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site specific model parameters are known.
+With sufficient data, however, site-specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
+Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site-specific model parameters are known.
+We apply this algorithm to large HIV-1 alignments and explore the consequences for phylogenetic inference.
 
@@
 	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
-Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, grey lines with $\alpha=3$.
+Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, gray lines with $\alpha=3$.
 	The estimates are more accurate than naive estimates from the state frequencies in alignment columns.
@@
 Regularization via pseudo-counts reduces over-fitting at low divergence.
+	Analogous results for alphabets of size $q=20$ are shown in Fig.~\ref{fig:dressed_supp}.
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Efficient inference of site specific substitution models}
-Following previous \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
-the overall rates and the equilibrium frequencies depend on the site.
-Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
+\section*{Efficient inference of site-specific substitution models}
+Following work by \citet{halpern_evolutionary_1998}, we parameterize a site-specific general time-reversible (GTR) substitution model at site $a$ from state $j$ to $i$ as:
 \begin{eqnarray}
 Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
-Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
+Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki} \ .
 \label{eq:Qij}
 \end{eqnarray}
-where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
-In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.
+Here, $\mu^a$ is the substitution rate at site $a$, $\eqp^a_i$ is the equilibrium frequency of state $i$ at site $a$, and $W_{ij}$ is a symmetric substitution matrix that we assume to be the same for all sites.
+The second equation ensures conservation of probability.
+In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to normalize the frequencies and fix the scale of $W_{ij}$.
 
-Extending the iterative expectation maximization approach by \citet{bruno1996modeling}, we show in the appendix that the iterative update rules
+Extending the approach by \citet{bruno1996modeling}, we show in Materials and Methods and the supplement that the iterative update rules
 \begin{equation}
@@
 approximately maximize the likelihood of the data.
-Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
+Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree, $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$,
 $\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
@@
 
-To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
-We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
+To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
+For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data set per parameter combination.
+We quantified the accuracy of the inferences as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities, respectively.
+We compared inferences using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
 
@@
 	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
-	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.}
+	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.
+	}
 	\label{fig:reconstructed}
@@
 
-Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site for estimates of $\eqp_i^a$ from the alignment and using the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
-The plot is shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope 1 implies that the accuracy is inversely proportional to the expected number of substitutions.
+Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
+The graph contains estimates of $\eqp_i^a$ from the alignment and via the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
+The data are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the accuracy is inversely proportional to the expected number of substitutions.
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distributed with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
-The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is much more evident for $\alpha=3$ than for $\alpha=1.5$ as the average accuracy is sensitive to a small number of sites with low rates at which frequencies are estimated poorly.
+The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
+The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
+In case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
 A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).
@@
 In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized).
-To test the influence of tree and ancestral state reconstruction, we built phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for simulated nucleotide and amino-acid sequences, respectively.
+To test the influence of tree and ancestral state reconstruction, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model (simulated nucleotide sequences) or FastTree \citep{FastTree} using the default 20 category (simulated amino-acid sequences).
 Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
@@
 This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
-Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}.
-More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site specific model (``Iterative model estimation'', brown line).
+Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
+More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
 The accuracy now continues to improve up to root-to-tip distances of about one.
@@
 Similar improvements are achieved by optimizing tree branch lengths along with the model.
-
 We found little difference between 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.
 
-\subsection*{Ignoring site specific frequencies results in branch length underestimates}
+\subsection*{Ignoring site-specific frequencies results in branch length underestimates}
 As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
 This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
-Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
-In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate \citet{wertheim_purifying_2011}.
+Models that don't account for site-specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
+In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent evolutionary rates \citet{wertheim_purifying_2011}.
 
@@
 Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
-Not that inferred site specific models only partially ameliorate underestimation, see main text.
+Not that inferred site-specific models only partially ameliorate underestimation, see main text.
 Parameters: $n=1000$, $\alpha=1.5$.
@@
 Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
-While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR=R10 model without site specificity.
-Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are prone to underestimation.
+While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR+R10 model without site-specificity.
+Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies which are prone to underestimation.
 When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
@@
 %% dependence on nucleotide and amino acid skew
-The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed, see Fig.~\ref{fig:tree_length_supp}.
+The underestimation of branch lengths is less severe for larger alphabets of if frequencies are not heavily skewed, see Fig.~\ref{fig:tree_length_supp}.
 
@@
 
-Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
+Surprisingly, using the inferred site-specific models only partially rectified the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
 The principle contributor to this deviation are inaccuracies in the rate estimates.
@@
 Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
-Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
+Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
 If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
 Deviations in rate and preferences are approximately additive.
-These observations are consistent with the finding above using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.
+These observations are consistent with the finding above that using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.
 
@@
 This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
-We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
+We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site-specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
 
 One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
-Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
+Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ should depend on the fitness difference $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
 \begin{equation}
@@
 \end{equation}
-On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site specific GTR model $\mu^a Q^a_{ij}$.
-The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations.
+On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site-specific GTR model $Q^a_{ij}$.
+The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations \citep{neher_genetic_2013}.
 Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
@@
 While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectional variation.
-Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
+Within host fitness effects are masked and damped at the population level, possibly due to fluctuating selection by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
+Such fluctuations in genetic background (epistasis) and environment (immune system) can result in non-linear averages at the population level.
 The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
@@
 	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
-	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
+	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.
+	Analogous results for the genes \gene{gag} and \gene{nef} are shown in Fig.~\ref{fig:HIV_fitness_supp}.}
 	\label{fig:HIV_fitness}
@@
 Next, we quantified the effect of ignoring site-specific preferences on branch length estimates.
-We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with homogeneous equilibrium frequencies set to the average frequencies across sites while maintaining site-specific rate variation.
+We generated sequence pairs that evolved for a specific time $t$ under the site-specific model inferred from the HIV-1 pol alignment and then estimated a maximum likelihood branch length between these two sequences.
+This estimate was done using a model with homogeneous equilibrium frequencies set to the average frequencies across sites while maintaining site-specific rate variation.
 As soon as the length of the simulated branch approaches 0.2, the length inferred by a model without site-specific preferences deviates substantially from the true value and saturates around 0.5 for larger and larger $t$.
@@
 A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
-Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site specific preference is the basis of all common homology search tools.
+Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site-specific preference is the basis of all common homology search tools.
 While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern_evolutionary_1998}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
@@
 
-Here, we presented method to estimate site specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
-We showed that site specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
+Here, we presented method to estimate site-specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
+We showed that site-specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
 For short branch lengths a parsimonious ancestral reconstruction is sufficient, while for more diverged samples iterative inference of the model, the ancestral states, and the branch lengths is necessary.
@@
 While using the correct model results in correct branch length estimates, joint inference of site-specific models and branch lengths is typically unable to recover the true branch lengths and substantial underestimation remains.
-This points to parameter identifiability problems rooted in similar effects of skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
+This points to parameter identifiability problems rooted in the similar effects that skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
 
@@
 This lack of information is further underscored by the average nucleotide distances between \gene{pol} sequences of different HIV and SIV strains:  HIV-1 subtypes differ at about 10\% of sites, distances between pol and HIV-1 and SIVcpz are on the order of 25\% and distances between sequences in the HIV-1/HIV-2/SIV compendium alignment are about 35\% (all distances ignore sites with $>20$\% gaps).
-These observations are compatible with evidence for frequent reversion of HIV to a preferred sequence state follwing immune escape \citep{zanini2015population,carlson_selection_2014,leslie_hiv_2004}.
+These observations are compatible with evidence for frequent reversion of HIV to a preferred sequence state following immune escape \citep{zanini2015population,carlson_selection_2014,leslie_hiv_2004}.
 
@@
 Typically, short branches close to the tips of the tree are used to calibrate molecular clock models.
-The deep branches, however, tend to be much longer and rate estimation and variation of site specific preferences will result in saturation effects not accounted for by the model \citep{hilton_modeling_2018}.
-This effective variation in rate is related to the effect of transient deleterious mutations that inflate the rate on short time scales \citep{ho_time_2005,wertheim_purifying_2011}: the time it takes purifying selection to prune deleterious mutations is related to the relaxation time scales of GTR models with site specific preferences.
+The deep branches, however, tend to be much longer and rate estimation and variation of site-specific preferences will result in saturation effects not accounted for by the model \citep{hilton_modeling_2018}.
+This effective variation in rate is related to the effect of transient deleterious mutations that inflate the rate on short time scales \citep{ho_time_2005,wertheim_purifying_2011}: the time it takes purifying selection to prune deleterious mutations is related to the relaxation time scales of GTR models with site-specific preferences.
 Instead of the $q-1$ degenerate eigenvalues of a Jukes-Cantor model, each site has a spectrum of eigenvalues and different eigenmodes relax at different speeds, generating apparently time dependent rates.
-
-While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint inference of model and branch lengths remains difficult.
-This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one  reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
 
@@
 Outside of this limit there either isn't enough data to reliably estimate the large number of coefficients, or epistatic interactions need to be taken into account -- probably at the expense of ignoring phylogenetic signals \citep{morcos_direct-coupling_2011}.
+
+\section*{Acknowledgments}
+We are grateful to Sarah Hilton and Pierre Barrat for stimulating discussions and insightful comments on the manuscript.
+This work was supported by core funding by the University of Basel, the Max Planck society, and ERC Stg 260686.
 
@@
 Most models of sequence evolution express the probability that sequence $\vec{s}$ evolved from sequence $\vec{r}$ in time $t$ as
-\bq
+\begin{equation}
 P(\vec{s}\leftarrow \vec{r},t,\mat{Q}) = \prod_{a=1}^L \left(e^{\mat{Q}^{a} t}\right)_{s^{a},r^{a}}
 \label{eq:P_prob}
-\eq
-where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $i$, and $s^{a}$ and $r^{a}$ are the sequence states at position $i$.
+\end{equation}
+where $\mat{Q}^{a}$ is the substitution matrix governing evolution at site $a$, and $s^{a}$ and $r^{a}$ are the sequence states at position $a$.
 The product runs over all $L$ sites $a$ and amounts to assuming that different sites of the sequence evolve independently.
@@
 In absence of recombination, homologous sequences are related by a tree and the likelihood of observing an alignment $\mat{A}=\{\vec{s_k}, k=1..n\}$ conditional on the tree $T$ and the substitution model $\mat{Q}^{a}$ can be written in terms of propagators defined in Eq.~\ref{eq:P_prob}.
-It is helpful to express this likelihood as product of sequence propagators defined in Eq.~\ref{eq:P_prob} between sequences at the ends of each branch in the tree (implicitly assuming that on different branches is independent and follows the same time reversible model).
+It is helpful to express this likelihood as product of sequence propagators defined in Eq.~\ref{eq:P_prob} between sequences at the ends of each branch in the tree (implicitly assuming that evolution on different branches is independent and follows the same time reversible model).
 Unknown sequences of internal nodes $\{\vec{s}'\}$ need to be summed over and the likelihood can be expressed as
@@
 where $\vec{s}_c$ and $\vec{s}_p$ are the child and parent sequences of branch $k$, respectively, and the factor $\prod_a \eqp^{a}_{s^{a}_0}$ is the product of the probabilities of the root sequence $s^{a}_0$ over all positions $a$.
-The probabilities $\vec{\eqp}^{a}$ are the equiblirium probabilities of the substitution model at position $a$.
+The probabilities $\vec{\eqp}^{a}$ are the equilibrium probabilities of the substitution model at position $a$.
 The latter ensures that the likelihood is insensitive to a particular choice of the tree root.
@@
 
-Instead of inferring completely independent models for every site in the genome, we follow \citet{halpern_evolutionary_1998} and only allow for site specific rates and equilibrium frequencies while using the same transition matrix for every site.
+Instead of inferring completely independent models for every site in the genome, we follow \citet{halpern_evolutionary_1998} and only allow for site-specific rates and equilibrium frequencies while using the same transition matrix for every site.
 Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
@@
 \subsection*{Iterative optimization algorithm}
-The derivatives of $\LH$ wrt to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
+The derivatives of $\LH$ with respect to $\mu^a$, $\eqp_i^a$, and $W_{ij}$ need to vanish at the values that maximize $\LH$.
 \begin{equation}
@@
 
-These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the supplement for the other (more laborious) update rules.
+These conditions can be solved iteratively.
+Here, we derive the update for $\mu^b$ and refer to the supplement for the other update rules.
 The derivative of the log-likelihood for a specific sequence assignment is given by
@@
 \end{equation}
-where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
+where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree (we have re-instantiated the position index $b$ in the last line).
 Additional terms necessary for regularization and normalization are discussed in the supplement.
 Setting this expression to zero (and the corresponding ones in the supplement) suggests solving for $\mu^b$ at fixed $\tau^b_j$ and $n^b_{ij}$ using the iterative update rules given in Eqs.~\ref{eq:update}.
-
-In addition, we introduced a pseudocount $\pc$ analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length, see supplement.
-To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator in Eqs.~\ref{eq:update} with the parameter that is being updated.
-This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
+The quantities $n^b_{ij}$ and $\tau^b_i$ can be averaged over unknown ancestral states.
 
 \subsection*{Implementation}
-We extended our package TreeTime \citep{sagulenko2017treetime} to handle site specific GTR models by adding an additional class \texttt{GTR\_site\_specific} that generalizes the \texttt{GTR} class.
-Since these classes have almost identical inferfaces, they can be used interchangeably in other analysis run by TreeTime.
-Using the new class, TreeTime can generate sequences with site specific evolutionary models and infer these models from sequence data using the algorithms above.
+We extended our package TreeTime \citep{sagulenko2017treetime} to handle site-specific GTR models by adding an additional class \texttt{GTR\_site\_specific} that generalizes the \texttt{GTR} class.
+Since these classes have an almost identical interface and can be used interchangeably in other analysis run by TreeTime.
+Using the new class, TreeTime can generate sequences with site-specific evolutionary models and infer these models from sequence data using the algorithms above.
 
 \subsection*{Generation of simulated data}
-To generate models and sequence ensembles for which the ground truth is known, we sampled binary trees with $n=100, 300, 1000, 3000$ leaves and Yule tree branch length statistics using \href{https://github.com/neherlab/betatree}{betatree} \citep{neher2013coalescence} for the range of values of the average substitution rate $\langle \mu\rangle = [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]$.
+To generate models and sequence ensembles for which the ground truth is known, we sampled binary trees with $n=100, 300, 1000, 3000$ leaves and Yule tree branch length statistics using \href{https://github.com/neherlab/betatree}{betatree} \citep{neher2013coalescence} for values of the average substitution rate given by $\langle \mu\rangle = [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]$.
 
@@
 The entries of the transition matrix $W_{ij}$ were sampled from a Dirichlet distribution with $\alpha=2$.
-The average substitution rate of these models was fixed to required $\langle \mu\rangle$.
+The average substitution rate of these models was fixed to the required $\langle \mu\rangle$.
 In addition, we generated one set of models ($q=4$) with rate variation but uniform $\eqp_i^a$ and $W_{ij}$.
@@
 For each alignment, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for nucleotide and amino-acid sequences, respectively.
-These trees were used to infer site specific models from data, see below.
+These trees were used to infer site-specific models from data, see below.
 The exact workflow is documented in the script \texttt{src/generate\_toy\_data.py} in the associated git repository at \href{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
@@
 
-The sequences and trees were then used for site specific GTR inference as implemented in TreeTime \citep{sagulenko2017treetime}.
-The scripts detailing this analysis and producing the figures are available on GitHub in repository \url{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
-
-\clearpage
-\appendix
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-
-\section*{Supplementary figures}
-\begin{figure*}[htb]
-	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha1.5}.pdf}
-	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha1.5}.pdf}
-	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree, for amino acid alphabets.
-(A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
-	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
-	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
-	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
-Regularization via pseudo-counts reduces over-fitting at low divergence.
-}
-	\label{fig:dressed_supp}
-\end{figure*}
-
-\begin{figure*}[tb]
-	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/aa_length_n300}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/aa_depth_n300}.pdf}
-	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates -- amino acid alphabet}.
-	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
-Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
-Note that inferred site specific models only partially ameliorate underestimation, see main text.
-}
-	\label{fig:tree_length_supp}
-\end{figure*}
-
-\begin{figure*}[tb]
-	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.010}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.010}.pdf}
-	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
-	In analogy to Fig.~\ref{fig:HIV_fitness}, the panels A\&B show the ratio of in/out rates for consensus nucleotides acids along the \gene{gag} and \gene{nef} gene of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
-	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
-	\label{fig:HIV_fitness_supp}
-\end{figure*}
-
-\begin{figure*}[htb]
-	\centering
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010}.pdf}
-	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010_aa}.pdf}
-	\caption{{\bf Diversity and GTR rate estimates.}
-	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs the diversity in columns of the corresponding alignment.}
-	\label{fig:HIV_diversity_supp}
-\end{figure*}
+The sequences and trees were then used for site-specific GTR inference as implemented in TreeTime \citep{sagulenko2017treetime}.
+The scripts detailing this analysis and producing the figures are available on GitHub in repository \href{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.
 

### 2020-01-18 manuscript/supplement.tex
typos and streamlining. Move supp figures to supplement

@@
 \usepackage{color}
+\usepackage{hyperref}
+\usepackage{xr}
+\externaldocument{ms}
 %\usepackage{cite}
@@
 \begin{document}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
 
@@
 
+\clearpage
+
+\section*{Supplementary figures}
+\begin{figure*}[htb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha1.5}.pdf}
+	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha1.5}.pdf}
+	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree, for amino acid alphabets.
+(A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
+	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
+	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
+	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
+Regularization via pseudo-counts reduces over-fitting at low divergence.
+}
+	\label{fig:dressed_supp}
+\end{figure*}
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/aa_length_n300}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/aa_depth_n300}.pdf}
+	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates -- amino acid alphabet}.
+	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
+Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
+Note that inferred site-specific models only partially ameliorate underestimation, see main text.
+}
+	\label{fig:tree_length_supp}
+\end{figure*}
+
+\begin{figure*}[tb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.010}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.010}.pdf}
+	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
+	In analogy to Fig.~\ref{fig:HIV_fitness}, the panels A\&B show the ratio of in/out rates for consensus nucleotides acids along the \gene{gag} and \gene{nef} gene of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
+	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
+	\label{fig:HIV_fitness_supp}
+\end{figure*}
+
+\begin{figure*}[htb]
+	\centering
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010}.pdf}
+	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010_aa}.pdf}
+	\caption{{\bf Diversity and GTR rate estimates.}
+	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs the diversity in columns of the corresponding alignment.}
+	\label{fig:HIV_diversity_supp}
+\end{figure*}
+
 \end{document}

### 2020-01-18 manuscript/ms.tex
typos, changed position label from a to b

@@
 
-\title{Potential and limits of site-specific substitution model inference.}
-\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard Neher$^{1,2}$}
+\title{Efficient inference, potential, and limitations of site-specific substitution models}
+\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard A.~Neher$^{1,2}$}
 \affiliation{
@@
 	The estimates are more accurate than naive estimates from the state frequencies in alignment columns.
-The advantage over naive estimates increases with data set size as seem from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
+The advantage over naive estimates increases with data set size as seen from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
 	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
@@
 \end{eqnarray}
-Here, $\mu^a$ is the substitution rate at site $a$, $\eqp^a_i$ is the equilibrium frequency of state $i$ at site $a$, and $W_{ij}$ is a symmetric substitution matrix that we assume to be the same for all sites.
+Here, $\mu^a$ is the substitution rate at site $a$, $\eqp^a_i$ is the equilibrium frequency of state $i$ at site $a$, and $W_{ij}$ is a symmetric substitution matrix that we assume to be the same for all sites (in what follows, super script $a$ will always refer to the position in the sequence, while subscript $i,j,n,m$ refers to the state).
 The second equation ensures conservation of probability.
@@
 	\begin{split}
-		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
-		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
-		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
+		\mu^a & \leftarrow \mu^a\frac{\pc+\sum_{ij} n^a_{ij}}{\sum_{ij}\mu^a \eqp_{i}W_{ij}\tau_j^a} \\
+		\eqp^a_i & \leftarrow \eqp^a_i\frac{\pc+\delta_{is^{a}} + \sum_{j\neq i} n^a_{ij}}{\eqp^a_i(q\pc+1) + \mu^a\eqp^a_i \sum_{j}W_{ij}\tau_j^a} \\
+		W_{ij} & \leftarrow W_{ij}\frac{\sum_a (n^a_{ij}+n^a_{ji})}{\sum_a \mu^a W_{ij}(\eqp_i^a\tau_j^a + \eqp_j^a \tau_i^a)} \\
 	\end{split}
@@
 approximately maximize the likelihood of the data.
-Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree, $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$,
-$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
+Here $\tau^a_j$ is the time site $a$ spends in state $j$ across the tree, $n^a_{ij}$ are the number of transitions from state $j$ to state $i$ at site $a$,
+$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^a$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
 To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
-This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
+This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^a_{ij}$ and the {\it expected} number $\eqp^a_i W_{ij}\tau_j^a$ -- each appropriately summed over sites or states.
 These update rules are an example of non-negative factorization algorithms \citep{lee_algorithms_2001}.
@@
 \section*{Accuracy of inferences}
-The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and (iii) the accuracy of the tree reconstruction.
+The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^a$ and $\tau_{j}^a$ can be estimated, and (iii) the accuracy of the tree reconstruction.
 In addition, the Maximum Likelihood estimates might deviate from the true parameters due to over-fitting.
 
-To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
+To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^a$ and the $\tau_j^a$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
 For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data set per parameter combination.
@@
 	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
-	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
+	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch lengths using the updated models (or using the true tree, yellow/pink).
 	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.
@@
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
-The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
+The simulation data was generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
+The scaling behavior $\chi_2 (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
 In case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
-A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).
+A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).
 
@@
 Nevertheless, the accuracy of the $\hat{\eqp}_i^a$ continues to increase.
-While equilibrium frequencies are insensitive, the substitution rate estimates are affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
+While equilibrium frequencies are insensitive, the substitution rate estimates are affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^a$.
 This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
@@
 Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
-The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
+The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^a_{ij}$ and $\tau_i^a$ from this reconstruction.
 This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
@@
 Similar improvements are achieved by optimizing tree branch lengths along with the model.
-We found little difference between 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.
+We found qualitatively similar patterns for 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.
+For larger alphabets, the breakdown at large root-to-tip distances is less dramatic and sets in at larger values.
+However, these difference depend on the skew of the equilibrium probabilities.
 
@@
 	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n1000}.pdf}
-	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
+	\caption{{\bf Skewed equilibrium concentration results in branch length underestimates}.
 	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
@@
 	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
+	The models assume an alphabet size $q=4$ (nucleotides).
 	}
@@
 Assuming a typical evolutionary rate of HIV of 0.002 changes per site and year, this analysis suggests that length estimates of branches longer than 100 years start to become inaccurate.
-Furthermore, this analysis suggests very little signal to estimate the length of branches that are longer than 300y.
+Furthermore, this analysis suggests very little signal to estimate the length of branches that are longer than 300 years.
 
@@
 \section*{Acknowledgments}
-We are grateful to Sarah Hilton and Pierre Barrat for stimulating discussions and insightful comments on the manuscript.
+We are grateful to Sarah Hilton and Pierre Barrat-Charlaix for stimulating discussions and insightful comments on the manuscript.
 This work was supported by core funding by the University of Basel, the Max Planck society, and ERC Stg 260686.
@@
 These conditions can be solved iteratively.
-Here, we derive the update for $\mu^b$ and refer to the supplement for the other update rules.
+Here, we derive the update for $\mu^a$ and refer to the supplement for the other update rules.
 The derivative of the log-likelihood for a specific sequence assignment is given by
 \begin{equation}
-	\frac{\partial \lh(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{\beta\in T} \frac{t_\beta}{\mu^b}\frac{\sum_i Q^b_{s^b_c,i}\left( e^{Q^bt_\beta} \right)_{i, s^b_p}}{\left( e^{Q^bt_\beta} \right)_{s^b_c, s^b_p}} \ ,
+	\frac{\partial \lh(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^a} = \sum_{\beta\in T} \frac{t_\beta}{\mu^a}\frac{\sum_i Q^a_{s^a_c,i}\left( e^{Q^at_\beta} \right)_{i, s^a_p}}{\left( e^{Q^at_\beta} \right)_{s^a_c, s^a_p}} \ ,
 \end{equation}
-where $s^b_p$ and $s^b_c$ are the states at site $b$ at the parent and child end of branch $\beta$.
-The individual terms in this sum behave very differently for cases where $s^b_c=s^b_p$ (no change at site $b$ on branch $\beta$) and $s^b_c\neq k^b_p$ (at least one mutation).
-In the limit of short branches $\mu^b t_\beta \ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain approximate but solvable conditions for maximum likelihood parameter estimates, see supplement.
-We will separate the sum over branches into those with $s^b_c=s^b_p$ and $s^b_c\neq s^b_p$.
-Suppressing the index $b$ for the position in the sequence, we find
+where $s^a_p$ and $s^a_c$ are the states at site $a$ at the parent and child end of branch $\beta$.
+The individual terms in this sum behave very differently for cases where $s^a_c=s^a_p$ (no change at site $a$ on branch $\beta$) and $s^a_c\neq k^a_p$ (at least one mutation).
+In the limit of short branches $\mu^a t_\beta \ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain approximate but solvable conditions for maximum likelihood parameter estimates, see supplement.
+We will separate the sum over branches into those with $s^a_c=s^a_p$ and $s^a_c\neq s^a_p$.
+Suppressing the index $a$ for the position in the sequence, we find
 \begin{equation}
@@
 	 + \sum_{\beta\in T,s_c\neq s_p} \frac{t_\beta \eqp_{s_c} W_{s_cs_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
-	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau^b_j + \sum_{i\neq j} n^b_{ij}/\mu
+	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau^a_j + \sum_{i\neq j} n^a_{ij}/\mu
 	\end{split}
 \end{equation}
-where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree (we have re-instantiated the position index $b$ in the last line).
+where $\tau^a_j$ is the sum of all branch length along which site $a$ is in state $j$ and $n^a_{ij}$ is the number of times the sequence at site $a$ changes from $j$ to $i$ along branches of the tree (we have re-instantiated the position index $a$ in the last line).
 Additional terms necessary for regularization and normalization are discussed in the supplement.
-Setting this expression to zero (and the corresponding ones in the supplement) suggests solving for $\mu^b$ at fixed $\tau^b_j$ and $n^b_{ij}$ using the iterative update rules given in Eqs.~\ref{eq:update}.
-The quantities $n^b_{ij}$ and $\tau^b_i$ can be averaged over unknown ancestral states.
+Setting this expression to zero (and the corresponding ones in the supplement) suggests solving for $\mu^a$ at fixed $\tau^a_j$ and $n^a_{ij}$ using the iterative update rules given in Eqs.~\ref{eq:update}.
+The quantities $n^a_{ij}$ and $\tau^a_i$ can be averaged over unknown ancestral states.
 

### 2020-01-18 manuscript/supplement.tex
typos, changed position label from a to b

@@
 
-\title{Supplement: Potential and limits of site-specific substitution model inference.}
+\title{Supplement: Efficient inference, potential, and limitations of site-specific substitution models}
 \author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard A.~Neher$^{1,2}$}
@@
 At the maximum, the derivatives of $\lh$ with respect to each of the model parameters should vanish when averaged over all internal site assignments.
-The derivative with respect to $\mu^b$ was is presented in Eq.~12 of the main text.
-Here, we present the corresponding derivations for $\eqp^b_n$ and $W_{nm}$.
-
 To facilitate the algebra below, we summarize the short branch length assumption and the associated linear approximations here:
@@
 	\begin{cases}
-	1 - t\mu^b \sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
-	t\mu^b \eqp^b_i W_{ik} & i\neq j
+	1 - t\mu^a \sum_{k\neq i} \eqp_k^a W_{ki} & i=j \\
+	t\mu^a \eqp^a_i W_{ik} & i\neq j
 	\end{cases}
@@
 \begin{eqnarray}
-	\frac{d}{d\mu^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
+	\frac{d}{d\mu^a} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
 	\begin{cases}
-	-t\sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
-	t \eqp^b_i W_{ik} & i\neq j
+	-t\sum_{k\neq i} \eqp_k^a W_{ki} & i=j \\
+	t \eqp^a_i W_{ik} & i\neq j
 	\end{cases} \\
-	\frac{d}{d\eqp_n^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
+	\frac{d}{d\eqp_n^a} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
 	\begin{cases}
-	-t\mu^b (1-\delta_{i,n})W_{ni} & i=j \\
-	 t\mu^b \delta_{i,n}W_{nk} & i\neq j
+	-t\mu^a (1-\delta_{i,n})W_{ni} & i=j \\
+	 t\mu^a \delta_{i,n}W_{nk} & i\neq j
 	\end{cases}\\
@@
 
-\section*{Update rule for rates $\mu^b$}
-Only site $b$ will contribute to the derivative of $\lh$ with respect to $\mu^b$.
+\section*{Update rule for rates $\mu^a$}
+Only site $a$ will contribute to the derivative of $\lh$ with respect to $\mu^a$.
 We will therefore drop all indices referring to the position and re-instantiate the index later.
@@
 Here, $\tau_j$ is the sum of all branch lengths on which $s_c=j$ and $n_{ij}$ is the number of times the sequence changes from $j$ to $i$ somewhere along the tree.
-Setting this expression to 0 and re-instantiating the site index $b$, we get
+Setting this expression to 0 and re-instantiating the site index $a$, we get
 \begin{equation}
-	\mu^b = \frac{\sum_{i\neq j} n^b_{ij}}{\sum_{j, k\neq j} \eqp^b_k W_{kj} \tau^b_j}
+	\mu^a = \frac{\sum_{i\neq j} n^a_{ij}}{\sum_{j, k\neq j} \eqp^a_k W_{kj} \tau^a_j}
 \end{equation}
 
-\section*{Update rule for frequencies $\eqp_i^b$}
-As above, only site $b$ will contribute to the derivative of $\lh$ with respect to $\eqp^b_n$.
+\section*{Update rule for frequencies $\eqp_i^a$}
+As above, only site $a$ will contribute to the derivative of $\lh$ with respect to $\eqp^a_n$.
 We will therefore drop all indices referring to the position and re-instantiate the index later.
@@
 \end{equation}
-Reinstantiating the position index $b$, we find
+Reinstantiating the position index $a$, we find
 \begin{equation}
-	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{\sum_{j\neq n} \mu W_{nj}\tau^b_j}
+	\eqp^a_n = \frac{\delta_{n,s^a_0}+\sum_j n^a_{nj}}{\sum_{j\neq n} \mu W_{nj}\tau^a_j}
 \end{equation}
@@
 
-\section*{Implementation of the constraint $\sum_i \eqp_i^b=1$}
-Equilibrium frequencies need to be normalized to one and constraints of this nature are typically accounted for by Lagrange parameters
+\section*{Implementation of the constraint $\sum_i \eqp_i^a=1$}
+Equilibrium frequencies need to be normalized to one and constraints of this nature can be accounted for by Lagrange parameters $\lambda^a$ coupling to the constraint:
 \begin{equation}
-	\frac{d}{d \eqp_n} \left[\lh (\{\vec{s}\} |T, \mat{Q}) + \lambda^b \sum_i \eqp^b_i \right] = 0
+	\frac{d}{d \eqp_n} \left[\lh (\{\vec{s}\} |T, \mat{Q}) + \lambda^a \sum_i \eqp^a_i \right] = 0
 \end{equation}
-which results in a modified extremal condition
+The constraint results in a modified extremal condition
 \begin{equation}
-	0 = \delta_{s^b_0,n}-\eqp_{n}\sum_{j\neq n} \tau^b_j \mu^b W_{j n} + \sum_{j\neq n} n^b_{nj} + \eqp^b_{n}\lambda^b
+	0 = \delta_{s^a_0,n}-\eqp_{n}\sum_{j\neq n} \tau^a_j \mu^a W_{j n} + \sum_{j\neq n} n^a_{nj} + \eqp^a_{n}\lambda^a
 \end{equation}
@@
 \begin{equation}
-	0 = 1 - \sum_{i,j; i \neq n}  \mu^b \tau^b_j \eqp^b_{i}W_{j i} + \sum_{i,j; i\neq j} n^b_{ij} + \lambda^b
+	0 = 1 - \sum_{i,j; i \neq n}  \mu^a \tau^a_j \eqp^a_{i}W_{j i} + \sum_{i,j; i\neq j} n^a_{ij} + \lambda^a
 \end{equation}
-At the extremum of $\mu^v$, we have $\sum_{i,j; i \neq n} \mu^b \tau^b_j \eqp^b_{i}W_{j i} = \sum_{i,j; i\neq j} n_{ij}$ and therefore $\lambda^b=-1$.
-With the constraint $\sum_i \eqp_i^b=1$, the update rule therefore becomes
+At the extremum of $\mu^a$, we have $\sum_{i,j; i \neq n} \mu^a \tau^a_j \eqp^a_{i}W_{j i} = \sum_{i,j; i\neq j} n_{ij}$ and therefore $\lambda^a=-1$.
+With the constraint $\sum_i \eqp_i^a=1$, the update rule therefore becomes
 \begin{equation}
-	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{1 + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
+	\eqp^a_n = \frac{\delta_{n,s^a_0}+\sum_j n^a_{nj}}{1 + \sum_{j\neq n} \mu W_{nj}\tau^a_j}
 \end{equation}
@@
 \section*{Regularization}
-In absence of data, i.e., when there are very few mutations at a particular site we need to regularize the problem.
+When there are very few mutations at a particular site we need to regularize the problem to avoid overfitting and numerical instabilities.
 To this end, we assume a Dirichlet prior of the form
 \begin{equation}
-\prod_i \left(\eqp^b_i\right)^c
+\prod_i \left(\eqp^a_i\right)^c
 \end{equation}
-The derivative of the logarithm of this is $c/\eqp^b_i$ and adding this term to the update rule results in
+which drives frequency estimates away from the boundary for $c>0$.
+The derivative of the logarithm of this prior is $c/\eqp^a_i$ and adding this term to the update rule results in
 \begin{equation}
-	\eqp^b_n = \frac{\delta_{n,s^b_0} + c +\sum_j n^b_{nj}}{1 + qc + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
+	\eqp^a_n = \frac{\delta_{n,s^a_0} + c +\sum_j n^a_{nj}}{1 + qc + \sum_{j\neq n} \mu W_{nj}\tau^a_j}
 \end{equation}
 where $q$ is the size of the alphabet.
-In an analogous fashion, we add a weak regularization to the update rules of $\mu^b$ and $W_{ij}$.
+In an analogous fashion, we add a weak regularization to the update rules of $\mu^a$ and $W_{ij}$.
 
@@
 In practice, the internal nodes are unknown and can either be fixed to the most-likely state or summed over.
-Since the approximate maximum likelihood conditions are linear, this average over unknown ancestral states simply amounts to replacing $\tau^b_i$ and $n^b_{ij}$ by their averages which can be efficiently computed by standard recursive algorithms.
+Since the approximate maximum likelihood conditions are linear, this average over unknown ancestral states simply amounts to replacing $\tau^a_i$ and $n^a_{ij}$ by their averages which can be efficiently computed by standard recursive algorithms.
 

### 2020-01-18 manuscript/ms.tex
small edits to abstract:

@@
 \begin{abstract}
-Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
-Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
+Abstract: Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
+Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
 Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
 Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
-We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site-specific preferences from large data sets with moderately diverged sequences.
+We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences.
 Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.

### 2020-06-22 manuscript/response_to_referees.tex
add stub response

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\Robert}[1]{{\color{celestialblue}Robert: #1}}
\newcommand{\Jan}[1]{{\color{deepsaffron}Jan: #1}}
\newcommand{\Emma}[1]{{\color{purple}Emma: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\color{black}\textbf{Response:} #1}}

\begin{document}
\section*{Response to reviewers}

\subsection*{Reviewer: 1}

\refa{Puller, Sagulenko, and Neher present a systematic look at the strengths and weaknesses of site-specific substitution models. They propose a new method for inferring site-specific parameters, investigate the limitations of the model, and applied it to within-in host and global HIV sequences. This work is clear, methodical, and will be of great interest to those who work on site-specific models. I recommend this work for publication. I have also listed some comments below but do not see any of them critical to acceptance.}

\response{Thank you very much for this assessment.}

\refa{- The authors mention that other methods to infer site-specific parameters do exist. How does their new algorithm compare in terms of speed or accuracy?}

\refa{- For the simulations, the author draw site-specific preferences from a Dirichlet distribution with $\alpha$ equal to 0.2 or 0.5. It would be nice to present a more interpretable value of the preferences as well, such as entropy or effective number.}

\refa{- Fig 5 and Fig S4 are both over-plotted and making the points transparent or adding histograms would help with interpretability. In Fig 1, the differences in linetype should be included in the legend. In Fig 2, it would be easier to keep the models straight if the lines were annotated on the graph directly instead of relying exclusively on color.}

\subsection*{Reviewer: 2}

\refb{Abstract, line 19: I would phrase this more carefully, as I personally feel that site-specific substitution models are in fact quite popular (with indeed their own site-specific evolutionary rate being assigned/assumed as well) and easy to set up in many inference packages.}

\refb{Abstract, line 21: It needs to be pointed out that a careful model selection strategy should already be part of a standard phylogenetic/phylodynamic data analysis, where indeed site-specific models are being included in the mix. That’s certainly how I do things, but then I don’t see what would be different with what the authors present here (more to be uncovered as I read through the manuscript obviously, but would be good to point out how the suggested approach will differ from such a model selection study).}

\refb{Abstract, line 22: why ‘approximate’? How does this differ from ‘regular’ ML?}

\refb{Abstract, lines 24-25: I assume it will be pointed out that the BEAST/BEAST2 packages do allow for site-specific models to be used (with for example a simple button that allows to select the SRD06 model, of which the paper should be cited)?}

\refb{Lines 120-122: This brings to mind a recent paper by Gascuel and Steel ‘A Darwinian Uncertainty Principle’ in which the issue of jointly estimating the ancestral states and the transitions of those states is investigated. Given the parameters the authors are interested in here and their assessment through simulations, it seems to me that the authors need to clarify the novelty of their investigations and how theirs sets them apart from those investigated by Gascuel and Steel (actually Figure 1 reminds me of Figure 2 in the paper by Gascuel and Steel).}

\refb{Line 141: True, and this deserves a lot more explanation in the manuscript, i.e. how far should one typically push the site-specific assumptions? The authors need to make it explicitly clear if they plan to estimate (and assess) one or more site-specific parameters for each site, or whether some of these parameters are shared across categories of sites as is most commonly done when dealing with data partitions (such as partitioning by codon position). If each site gets its own parameter(s), then overfitting seems to be a very real issue; but what about other cases such as pre-defined partitions? Or what about drawing certain parameters from a certain distribution or a hierarchical phylogenetic model, which could very much aid with overfitting issues by sharing data?}

\refb{Line 150: ‘data sets’}

\refb{Line 157: ‘tree’}

\refb{Lines 204-208: Again, this seems related to Gascuel and Steel.}

\refb{In my opinion, the ‘Accuracy of inferences’ section is hard to digest, mostly because of its technicality and – in my personal opinion – due to the fact that the figures (specifically figures 1 and 2) aren’t very intuitive so I would suggest the authors to rework both the use of the Y-axis (if the authors want to discuss accuracy in the text, then the graphs should be going up in the figures as well), as well as the legends within the figures, which also aren’t self-explanatory and require to go through the text in detail.}

\refb{Lines 239-240: I’m not particularly convinced that uniform frequencies are still the norm ...}

\refb{Line 308: I fear this example analysis is quite (or too) technical and am not sure this will appeal to typical readers of Virus Evolution. Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}

\refb{But what I’m missing the most from the results and discussion sections of the manuscript, is clear advice on how to perform analyses on real data, what to look out for and how to potentially remedy the problem. The manuscript is too technical in that regard, and while assessing accuracy problems is important, a step-by-step approach for empirical data sets- to potentially/hopefully mitigate the problems mention in this manuscript - is even more important. It would be most helpful if the authors comment on the typical workflow, i.e. running ModelTest or something similar (PartitionFinder, automatic/included model selection in IQ-TREE, ...) and then sticking with the model found to infer the phylogeny. How often would one get in trouble with such an approach? And if problems arise, how severe are they? Could the clustering potentially be impacted or does the problem stay limited to certain branches (of certain length)? Or is the tree as a whole deemed older/younger than it really is, i.e. does the underestimation propogate through the tree all the way to the root or not?}

\refb{In a way, I’m not too sure what the conclusions from Figure 3 are in practice? Everything except for the true model results in underestimation, even the site-specific model in TreeTime. From a practical point of view, what could be done to fix the underestimation problem? Would some form of post-hoc correction be advised, but how would one know which branches to correct and which to leave along? In that regard, could thorough knowledge of the data provide any guidance; but even then, and related to the HIV example (and other virus examples), inaccurate branch lengths of 100 years or more are most likely not a major concern.}

\refb{The manuscript deals with estimates (i.e. trees) that were obtained using IQ-TREE and FastTree (which produce unrooted trees and don’t employ the concept of a molecular clock), which were subsequently fed into TreeTime which employs the provided trees as is. This makes me wonder if the conclusions from the manuscript hold up in frameworks that i) employ a molecular clock directly in their estimations, and ii) do not employ a fixed topology but rather explore a distribution of trees as typically done in Bayesian phylogenetic inference. I’m not suggesting the authors do the same experiments in a Bayesian framework but their take on this could be interesting for the discussion section.}

\refb{Lines 651-653: I’m going to assume that most readers, like myself, will not be very familiar with the work of Halpern and Bruno (1998). Hence these sentences will need to provide more information regarding the following statement: ‘only allow for site-specific rates and equilibrium frequencies while using the same transition matrix for every site.’ Regarding the site-specific (evolutionary) rates, there may be confusion as to whether the process meant is actually among-site rate variation (typically modelled as a discretized gamma distribution according to various paper of Ziheng Yang), or relative rates between positions which can be superimposed on the former model. I’m going to assume the latter, but in that case it’s not clear whether for each site a unique rate will be estimated (if so, that’s a lot of additional parameters and this needs to be discussed) or whether classes of sites will be imposed (the three different codon positions for example) and that for only each of these classes a unique rate will be estimated? Essentially, I have the same question regarding the equilibrium frequencies, as it’s not clear from the methods section (but I can guess from the simulated data section; still, the authors should be more explicit here).}

\refb{Implementation: this type of modelling is not new, although it surely is a useful -- and needed -- extension to TreeTime. It’s unclear at this point whether the transition matrix will also be (potentially) site-specific or not, and I encourage the authors to mention this explicitly in the implementation paragraph.}

\refb{Line 681: This makes it sounds like either two data partitions will be simulated that will either be concatenated or the contents will be scrambled to create one alignment of length L=2000 that is then a mixture of the different (site-specific) processes. But I have the feeling that this is not what the authors are after; can this be more explicitly explained?}

\refb{Lines 690-691: this may be confusing for readers, i.e. why is TreeTime not mentioned here? It should be pointed out then that a tree topology is required to be provided to TreeTime.}

\refb{Line 708: ‘site-specific GTR inference’, will the site-specific processes be bound using a (Gamma) distribution as was explained/used in the simulation section?}

\subsection*{Associate Editor: Martin, Darren}

This seems like a good paper but in its current state virus evolution may not be a very good fit for it.  I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.

\end{document}

### 2020-06-22 manuscript/response_to_referees.tex
added comments or preliminary responses

@@
 \newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
-\newcommand{\Robert}[1]{{\color{celestialblue}Robert: #1}}
-\newcommand{\Jan}[1]{{\color{deepsaffron}Jan: #1}}
-\newcommand{\Emma}[1]{{\color{purple}Emma: #1}}
 \newcommand{\refa}[1]{\textbf{R1:} #1}
@@
 \section*{Response to reviewers}
+
+\begin{itemize}
+    \item general issue: site specific models are computationally expensive, but only useful for large data sets.
+    \item main objective here is to provide a method for estimating them from large data sets, but point out the limitations.
+\end{itemize}
 
@@
 
+\Richard{ask Sarah}
+
 \refa{- For the simulations, the author draw site-specific preferences from a Dirichlet distribution with $\alpha$ equal to 0.2 or 0.5.
 It would be nice to present a more interpretable value of the preferences as well, such as entropy or effective number.}
+
+\Richard{Probably easiest to sample these. }
 
@@
 
+\Richard{maybe use hex-grids for Fig 5. add additional legend to Fig 1, think about Fig 2 }
+
 \subsection*{Reviewer: 2}
@@
 \refb{Abstract, line 19: I would phrase this more carefully, as I personally feel that site-specific substitution models are in fact quite popular (with indeed their own site-specific evolutionary rate being assigned/assumed as well) and easy to set up in many inference packages.}
+
+\response{Thank you for this note.
+We changed this to refer to 'analyses' rather than 'software'.
+In our experience, most uses do not account for this, either because of computational complexity or lack of data.
+}
 
@@
 
+\response{The scheme is approximate, in that the parameters that maximize the likelihood are found through iterative solution of extremal conditions obtained through an approximation of the full likelihood.
+}
+
 \refb{Abstract, lines 24-25: I assume it will be pointed out that the BEAST/BEAST2 packages do allow for site-specific models to be used (with for example a simple button that allows to select the SRD06 model, of which the paper should be cited)?}
+
+\response{To our knowledge, these are are partition models for different positions in a codon rather than models that allow independent frequencies at each position.
+We discuss these more carefully in the main text.}
 
@@
 Given the parameters the authors are interested in here and their assessment through simulations, it seems to me that the authors need to clarify the novelty of their investigations and how theirs sets them apart from those investigated by Gascuel and Steel (actually Figure 1 reminds me of Figure 2 in the paper by Gascuel and Steel).}
+
+\response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do.
+Gascuel and Steel are mostly concerned with reconstruction of a single character where the entire rate matrix needs to be estimated, while we investigate sequence evolution where site-specific parameters are restricted to frequencies and rates.
+}
 
@@
 
-\refb{Line 150: ‘data sets’}
+\response{We indeed investigate models with multiple free parameters per site -- in contrast to category models.
+Much of the remainder of the manuscript is devoted to the issue of identifiablility and over-fitting.
+}
+\Richard{Come back with specific fixes.}
 
-\refb{Line 157: ‘tree’}
+\refb{Line 150: `data sets'}
+\response{fixed, thanks.}
+
+\refb{Line 157: 'tree'}
+\response{fixed, thanks.}
 
 \refb{Lines 204-208: Again, this seems related to Gascuel and Steel.}
+\response{Indeed, Gascuel and Steel show a similar graph for a single character reconstruction using different models. We now point out this analogy.}
+\Richard{TODO}
 
-\refb{In my opinion, the ‘Accuracy of inferences’ section is hard to digest, mostly because of its technicality and – in my personal opinion – due to the fact that the figures (specifically figures 1 and 2) aren’t very intuitive so I would suggest the authors to rework both the use of the Y-axis (if the authors want to discuss accuracy in the text, then the graphs should be going up in the figures as well), as well as the legends within the figures, which also aren’t self-explanatory and require to go through the text in detail.}
+\refb{In my opinion, the ‘Accuracy of inferences’ section is hard to digest, mostly because of its technicality and -- in my personal opinion -- due to the fact that the figures (specifically figures 1 and 2) aren’t very intuitive so I would suggest the authors to rework both the use of the Y-axis (if the authors want to discuss accuracy in the text, then the graphs should be going up in the figures as well), as well as the legends within the figures, which also aren’t self-explanatory and require to go through the text in detail.}
 
@@
 Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}
+
+\response{We have made an effort to simplify and clarify this example.
+But we feel that the application of to HIV data is one aspect of what makes this paper suitable for Virus Evolution: Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution.
+}
 
@@
 
+\Richard{See what model selectors do for simulated data}
+
 \refb{In a way, I’m not too sure what the conclusions from Figure 3 are in practice?
@@
 
+\Richard{This is one core part of the paper: this is a difficult problem to fix. }
+
 \refb{The manuscript deals with estimates (i.e.
@@
 I’m not suggesting the authors do the same experiments in a Bayesian framework but their take on this could be interesting for the discussion section.}
+
+\Richard{We are not dealing with time scaled trees here. We only use treetime to estimate branch lengths using specific models and estimate models from data.}
 
@@
 
+\Richard{This shows that this reviewer really didn't understand what we are doing. We don't do partition models. }
+
 \refb{Implementation: this type of modelling is not new, although it surely is a useful -- and needed -- extension to TreeTime.
 It’s unclear at this point whether the transition matrix will also be (potentially) site-specific or not, and I encourage the authors to mention this explicitly in the implementation paragraph.}
+
+\response{We currently implement the model as specified in Equation 1. We now mention this explicitly and point out that site specific $W_{ij}$ are not currently implemented.}
 
@@
 
+\response{The reviewer is correct, we don't have two data partitions of length 1000.
+Instead, we produce hierarchical data sets that allow us to explore the effect of different aspects of the problem (tree, model realization, realization of the sequence evolution given the model) on the inferences.
+We added sentence to make this clear. }
+
 \refb{Lines 690-691: this may be confusing for readers, i.e. why is TreeTime not mentioned here? It should be pointed out then that a tree topology is required to be provided to TreeTime.}
 
+\response{We added `using TreeTime' to the sentence.}
+
 \refb{Line 708: ‘site-specific GTR inference’, will the site-specific processes be bound using a (Gamma) distribution as was explained/used in the simulation section?}
+
+\response{The site specific models were estimated using iterative approach described in Eq. 2 which effectively regularizes the problem with a Dirichlet prior. }
 

### 2020-06-22 manuscript/ms.tex
added comments or preliminary responses

@@
-\documentclass[aps,rmp,twocolumn]{revtex4}
+\documentclass[aps,rmp,twocolumn,linenumbers]{revtex4-1}
 %\documentclass[a4paper,10pt]{scrartcl}
@@
 We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences.
-Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
+Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogenetic analyses -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
 However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
@@
 To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^a$ and the $\tau_j^a$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
-For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data set per parameter combination.
+For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data sets per parameter combination.
 We quantified the accuracy of the inferences as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities, respectively.
@@
 
-Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
+Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a tree with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
 The graph contains estimates of $\eqp_i^a$ from the alignment and via the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
@@
 \subsection*{Implementation}
-We extended our package TreeTime \citep{sagulenko2017treetime} to handle site-specific GTR models by adding an additional class \texttt{GTR\_site\_specific} that generalizes the \texttt{GTR} class.
+We extended our package TreeTime \citep{sagulenko2017treetime} to handle site-specific GTR models as defined in Eq.~\ref{eq:Qij} by adding an additional class \texttt{GTR\_site\_specific} that generalizes existing the \texttt{GTR} class.
 Since these classes have an almost identical interface and can be used interchangeably in other analysis run by TreeTime.
 Using the new class, TreeTime can generate sequences with site-specific evolutionary models and infer these models from sequence data using the algorithms above.
+In the future, this could extended to also allow site specific $W_{ij}$ but this is not implemented as of now.
 
@@
 
+Given these tree, we generated multiple sets of sequences of length $L=1000$ for each tip of the tree.
+Specifically, we explored the effect of model choice and realization of the evolutionary process by generating data sets as follows:
 For each tree, we sampled two site-specific models for sequences of length $L=1000$ from following distribution:
@@
 For each alignment, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for nucleotide and amino-acid sequences, respectively.
-These trees were used to infer site-specific models from data, see below.
+These trees were used to infer site-specific models from data using TreeTime, see below.
 The exact workflow is documented in the script \texttt{src/generate\_toy\_data.py} in the associated git repository at \href{https://github.com/neherlab/2019_Puller_SiteSpecificGTR}{github.org/neherlab/2019\_Puller\_SiteSpecificGTR}.

### 2020-07-17 src/summarize_preferences.py
script to summarize entropy

@@ -1,0 +1,30 @@
+import matplotlib.pyplot as plt
+from generate_toy_data import *
+
+def entropy(a):
+    eps = 1e-10
+    return np.sum(np.log(a+eps)*(a+eps), axis=0)
+
+
+if __name__ == '__main__':
+
+    import argparse
+    parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
+    parser.add_argument('--mask', type=str, nargs='+')
+    args = parser.parse_args()
+
+    fs = 16
+    plt.figure()
+    for mask in args.mask
+        files = glob.glob(mask)
+        allS = []
+        for file in files:
+            m = load_model(file)
+            S = entropy(m.Pi)
+            allS.extend(S)
+
+        plt.plot(sorted(allS), np.linspace(0,1,len(allS)))
+        print(f"{mask}, average entropy: {np.mean(allS)}")
+    plt.ylabel('fraction with less entropy', fontsize=fs)
+    plt.xlabel('fraction of sites', fontsize=fs)
+    plt.savefig('figures/entropy_stats.pdf')

### 2020-07-17 manuscript/response_to_referees.tex
consider hex plot

@@
 
-\Richard{Probably easiest to sample these. }
+\response{
+Thank you for this suggestion.
+For amino acid alphabets, $\alpha=0.2$ and $\alpha=0.5$ correspond to an average entropy of $1.8$ or $2.3$ (in natural log units), respectively.
+Calculating the effective number of states as $\left(\sum_i p_i^2\right)^{-1}$ yields 4.7 or 7.8 effective states.
+We added this to the manuscript.}
 
@@
 We changed this to refer to 'analyses' rather than 'software'.
-In our experience, most uses do not account for this, either because of computational complexity or lack of data.
+In our experience, most users do not account for this, either because of computational complexity or lack of data.
 }

### 2020-07-17 src/analyze_HIV_tree.py

@@ -107,12 +107,18 @@ if __name__ == '__main__':
     div_rec = diversity(gtr.Pi)
     cc_rec = ccoef(np.log(fabio_fitness['median'])[valid], np.log(1e-10+div_rec)[valid])
     plt.figure()
-    plt.scatter(fabio_fitness['median'], div_rec, label='reconstructed, r2=%1.2f'%cc_rec[0]**2)
-    plt.legend()
-    plt.xlim([3e-5, 3e0])
-    plt.ylim([3e-5, 1.5])
-    plt.yscale('log')
-    plt.xscale('log')
+    g = sns.jointplot(fabio_fitness['median'], div_rec, kind=hex, )
+    g.ax_joint.set_yscale('log')
+    g.ax_joint.set_xscale('log')
+    g.ax_joint.set_xlim([3e-5, 3e0])
+    g.ax_joint.set_ylim([3e-5, 1.5])
+
+    # plt.scatter(fabio_fitness['median'], div_rec, label='reconstructed, r2=%1.2f'%cc_rec[0]**2)
+    # plt.legend()
+    # plt.xlim([3e-5, 3e0])
+    # plt.ylim([3e-5, 1.5])
+    # plt.yscale('log')
+    # plt.xscale('log')
 
     div_aln = diversity(aln_freq)
     cc_aln = ccoef(np.log(fabio_fitness['median'][valid]), np.log(1e-10+div_aln)[valid])

### 2020-07-17 src/analyze_HIV_tree.py
try hexbins

@@ -4,6 +4,7 @@ import numpy as np
 from scipy.stats import pearsonr, spearmanr
 from collections import defaultdict
 import matplotlib.pyplot as plt
+import seaborn as sns
 
 from treetime.treeanc import TreeAnc
 from treetime.gtr_site_specific import GTR_site_specific
@@ -107,11 +108,11 @@ if __name__ == '__main__':
     div_rec = diversity(gtr.Pi)
     cc_rec = ccoef(np.log(fabio_fitness['median'])[valid], np.log(1e-10+div_rec)[valid])
     plt.figure()
-    g = sns.jointplot(fabio_fitness['median'], div_rec, kind=hex, )
-    g.ax_joint.set_yscale('log')
-    g.ax_joint.set_xscale('log')
-    g.ax_joint.set_xlim([3e-5, 3e0])
-    g.ax_joint.set_ylim([3e-5, 1.5])
+    g = plt.hexbin(fabio_fitness['median'], div_rec, xscale='log', yscale='log')
+#    g.ax_joint.set_yscale('log')
+#    g.ax_joint.set_xscale('log')
+#    g.ax_joint.set_xlim([3e-5, 3e0])
+#    g.ax_joint.set_ylim([3e-5, 1.5])
 
     # plt.scatter(fabio_fitness['median'], div_rec, label='reconstructed, r2=%1.2f'%cc_rec[0]**2)
     # plt.legend()
@@ -133,15 +134,17 @@ if __name__ == '__main__':
 
     plt.figure()
     cc_fit = ccoef(np.log(fabio_fitness['median'][valid]), np.log(1e-10+fitness_estimates_gtr)[valid])
-    plt.scatter(fabio_fitness['median'], fitness_estimates_gtr, label=r'$r^2=$'+'%1.2f'%cc_fit[0]**2)
-    plt.legend(fontsize=fs)
+    #plt.scatter(fabio_fitness['median'], fitness_estimates_gtr, label=r'$r^2=$'+'%1.2f'%cc_fit[0]**2)
+    #plt.legend(fontsize=fs)
+    plt.hexbin(fabio_fitness['median'], fitness_estimates_gtr, xscale='log', yscale='log', gridsize=16, mincnt=1, vmin=1, vmax=75)
     plt.xlim([3e-5, 3e0])
     plt.ylim([3e-2, 3e3])
[…]

### 2020-07-17 src/analyze_HIV_tree.py
plot mods

@@ -157,15 +157,17 @@ if __name__ == '__main__':
 
     plt.figure()
     cc_div = ccoef(np.log(1e-10+div_aln), np.log(1e-10+fitness_estimates_gtr))
-    plt.scatter(div_aln, fitness_estimates_gtr, label=r'$r^2=$'+'%1.2f'%cc_div[0]**2)
-    plt.legend(fontsize=fs)
+    # plt.scatter(div_aln, fitness_estimates_gtr, label=r'$r^2=$'+'%1.2f'%cc_div[0]**2)
+    #plt.legend(fontsize=fs)
+    plt.hexbin(div_aln, fitness_estimates_gtr, xscale='log', yscale='log', gridsize=16, mincnt=1, vmin=1, vmax=75)
+    plt.text(3e-4, 3e3, r'$r^2=$'+'%1.2f'%cc_div[0]**2, fontsize=fs)
     plt.title('amino acids' if args.aa else 'nucleotides',fontsize=fs)
     plt.xlim([2e-4, 3e0])
     plt.ylim([3e-2, 9e3])
     plt.xlabel('diversity in alignment', fontsize=fs)
     plt.ylabel(r'GTR based fitness estimates $\Gamma_{in}/\Gamma_{out}$', fontsize=fs)
-    plt.xscale('log')
-    plt.yscale('log')
+    # plt.xscale('log')
+    # plt.yscale('log')
     plt.tick_params(labelsize=0.8*fs)
     plt.tight_layout()
     plt.savefig('figures/'+args.prefix.split('/')[-1]+'_alignment_div_pc_%1.3f'%args.pc+ ('_aa' if args.aa else '')+'.pdf')

### 2020-07-17 src/plot_toy_data.py

@@ -66,16 +66,20 @@ def plot_pdist_vs_tree_length(data, data_secondary, subsets, fname=None):
     cols = {n:'C%d'%i for i,n in enumerate(subsets['n'])}
 
     mean_vals = make_means(data, subsets)
+    lines = []
+    dash_legend = []
     for params, d in mean_vals.items():
         tmp_p = {k:v for k,v in params}
         subs_per_site = d['mean'].index*tmp_p['n']
-        plt.errorbar(subs_per_site, d['mean']['chisq_p'], d['std']['chisq_p'],
-                     label=labels[tmp_p['method']] if tmp_p['n']==np.max(subsets['n']) else '',
+        l = plt.errorbar(subs_per_site, d['mean']['chisq_p'], d['std']['chisq_p'],
                      c=colors[tmp_p['method']], ls=ls[tmp_p['n']])
+        if tmp_p['n']==np.max(subsets['n']):
+            lines.append([l, labels[tmp_p['method']]])
 
         if tmp_p['method']=='naive':
-            plt.text(subs_per_site[0], d['mean']['chisq_p'].iloc[0]*1.2,
-                    f"n={tmp_p['n']}", fontsize=fs*0.8)
+            dash_legend.append(l, f"n={tmp_p['n']}")
+            # plt.text(subs_per_site[0], d['mean']['chisq_p'].iloc[0]*1.2,
+            #         f"n={tmp_p['n']}", fontsize=fs*0.8)
 
     if not (data_secondary is None):
         mean_vals_secondary = make_means(data_secondary, subsets)
@@ -86,11 +90,13 @@ def plot_pdist_vs_tree_length(data, data_secondary, subsets, fname=None):
                          label='', c="#AAAAAA", ls=ls[tmp_p['n']])
 
 
-    plt.plot([10,1000], [0.1,0.001], label=r'$\sim x^{-1}$', c='k')
+    l = plt.plot([10,1000], [0.1,0.001]) #, label=r'$\sim x^{-1}$', c='k')
+    lines.append([l,r'$\sim x^{-1}$', c='k'])
     plt.ylim([0.001, 1.6])
     plt.yscale('log')
     plt.xscale('log')
-    plt.legend(fontsize=fs)
+    plt.legend([x[0] for x in lines], [x[1] for x in lines], loc=3, fontsize=fs)
+    plt.legend([x[0] for x in dash_legend], [x[1] for x in dash_legend], loc='center left', fontsize=fs)
[…]

### 2020-07-23 manuscript/response_to_referees.tex
worked on response and adjusted manuscript

@@
 \Richard{ask Sarah}
+\response{we have attempted to compare our results with
+\texttt{swmutsel}. However, \texttt{swmutsel} seems completely focussed on codon models, something that isn't currenly implemented in treetime. In terms of run-time,
+swmutsel takes hours where the site specific inference by treetime takes minutes.
+But we are afraid the results are not immediately comparable. }
 
@@
 
-\Richard{maybe use hex-grids for Fig 5. add additional legend to Fig 1, think about Fig 2 }
+\response{
+We are grateful for these suggestions to improve the figures:
+\begin{itemize}
+    \item we changed Fig 5 and Fig S4 to two dimensional histograms
+    \item we added second legend to figure 1 that maps line style to samples size
+    \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
+\end{itemize}
+}
 
@@
 \response{Thank you for this note.
-We changed this to refer to 'analyses' rather than 'software'.
-In our experience, most users do not account for this, either because of computational complexity or lack of data.
+We realized that the intention and model class studied in our manuscript wasn't clear enough.
+We now emphasize that we study whether models with different state preferences at every site can be accurately estimated.
+
+We now point out in the abstract that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
  }
@@
 That’s certainly how I do things, but then I don’t see what would be different with what the authors present here (more to be uncovered as I read through the manuscript obviously, but would be good to point out how the suggested approach will differ from such a model selection study).}
+
+\response{
+We now point out our objective more clearly: Our primary goal is to infer site specific preferences to learn about the selection pressures operating on the genome, not phylogenetic inference.
+Specifically, we ask how much data is required to estimate state preferences at every site, how uncertainty in topology affects the estimates, and which aspects of the model are well or poorly constraint.
+As such, the problem is distinct (but related) to a model selection problem.
+}
 
@@
 
-\response{To our knowledge, these are are partition models for different positions in a codon rather than models that allow independent frequencies at each position.
-We discuss these more carefully in the main text.}
+\response{We discuss site-heterogenous partition models carefully in the main text.}
 
@@
 \response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do.
-Gascuel and Steel are mostly concerned with reconstruction of a single character where the entire rate matrix needs to be estimated, while we investigate sequence evolution where site-specific parameters are restricted to frequencies and rates.
+Gascuel and Steel investigate whether the state of the root if a character can be inferred jointly with the equilibrium probabilities or the transition matrix.
+Consistent with our results, they find that this is possible for Yule trees.
+Here, we are mmostly concerned with estimating the model, not the root state.
+We also consider sequence evolution rather than the evolution of a single character.
+This allows us to investigate the extent to which branch lengths can be estimated along with the model.
+
+We now discuss our results in the light of the results by Gascuel and Steel and thank the reviewer for pointing out this work to us.
 }
@@
 \response{We indeed investigate models with multiple free parameters per site -- in contrast to category models.
+In our model the symmetric substitution matrix is shared across sites, but each site has has its own rate and frequencies.
 Much of the remainder of the manuscript is devoted to the issue of identifiablility and over-fitting.
+As data sets grow in size, such site specific parameters become increasingly identifiable and this is focus of this study.
+The properties of category models have for example been studied systematically by Kainer and Lanfear.
+
+We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of the inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
 }
-\Richard{Come back with specific fixes.}
 
@@
 \refb{Lines 204-208: Again, this seems related to Gascuel and Steel.}
-\response{Indeed, Gascuel and Steel show a similar graph for a single character reconstruction using different models. We now point out this analogy.}
-\Richard{TODO}
+
+\response{Indeed, Gascuel and Steel show a similar graph for the joint estimation of ancestral state of a single discrete character and the model of its evolution. We now point out this analogy.
+
+The connection to Gascuel and Steel is particularly apt when discussion the inability to jointly infer the ancestral state and the model.
+We also draw attention to the fact that accurate model estimation is possible for know trees, consistent with the results by G\&S that estimation is possible for Yule trees.
+}
 
@@
 
+\response{We have rewritten a good part of this section.
+We now point out that we first investigate the performance of the inference scheme with known tree and ancestral states and then investigate the feasibility of estimating tree, ancestral states, and model together.
+
+Reviewer 1 also had suggestions to improve the figures and we have added more information to the legends and made it easier to differentiate between different line.
+We adjusted to text to more explicitly point out that the graphs show squared error, i.e., a quantity that decreases with increasing accuracy.
+}
+
 \refb{Lines 239-240: I’m not particularly convinced that uniform frequencies are still the norm ...}
+
+\response{We are not making a statement here on what is the norm and we are aware that complex models with multiple partitions are common.
+However, many large phylogenetic analysis of viruses use models that account for rate heterogeneity among sites, but use a common matrix of relative substitution rates for all site (PMID 25278604 for example).
+}
 
@@
 Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}
+
+\response{We have made an effort to make this section more accessible, but we think it is an important part of the manuscript and one of the reasons we think Virus Evolution is an appropriate journal.
+This section explores the ability to estimate site specific preferences from large and diverse alignments of RNA viruses.
+Similar analysis could be done for other RNA viruses.
+Furthermore, it points out implications for phylogenetic inferences.
+We don't have solutions for the latter, but feel that it is still important to point them out.
+}
 
@@
 
-\Richard{We are not dealing with time scaled trees here. We only use treetime to estimate branch lengths using specific models and estimate models from data.}
+\response{
+We are afraid there is a misunderstanding here.
+This manuscript deals with GTR models and the root doesn't matter.
+We only use root-to-tip distances in mid-point-rooted trees as a measure of divergence.
+The only place where time comes in are the HIV examples, but we only use time as a proxy for evolutionary rate and don't do any molecular clock analysis.
+All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}
 

### 2020-07-23 manuscript/ms.tex
consider hex plot
worked on response and adjusted manuscript

@@
 Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
-Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
-Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
-We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences.
+Modeling this complexity, however, requires an extensive number or parameters that can not be infered from small data sets and are costly to infer from large data sets.
+Most commonly used evolutionary models therefore reduce this complexity by partitioning sites into discrete classes with different rates and/or preferences where appropriate complexity is determined by model selection procedures.
+Here, we present an efficient algorithm to estimate more complex models that allow for different preferences at every site and explore the accuracy at which such models can be estimated from simulated data.
+We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
 Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogenetic analyses -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
@@
 
-In addition to rate variation, different sites in a protein differ in amino acid they allow.
+In addition to rate variation, different sites in a protein differ in the amino acids they allow and constraints on the DNA or RNA sequence impose additional levels of selection.
+Such variation can again be accounted for by modeling multiple categories with different preferences for amino acids or nucleotides to which sites can be assigned or which can be integrated out during phylogenetic inference \citep{Lartillot01062004,kainer_effects_2015,shapiro_choosing_2006}.
+Here, we ask study under what circumstances models that allow different state frequencies at every site in the alignment can be estimated and what implications such variation has for phylogenetic inference.
+
 Recent deep mutational scanning experiments have shown that site-specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
@@
 The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^a$ and $\tau_{j}^a$ can be estimated, and (iii) the accuracy of the tree reconstruction.
-In addition, the Maximum Likelihood estimates might deviate from the true parameters due to over-fitting.
+Furthermore, the estimation of this extensive number of parameters requires large data sets in which most positions mutate multiple times on the tree -- otherwise the model will overfit the data and return inaccurate estimates.
 
-To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^a$ and the $\tau_j^a$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
-For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data sets per parameter combination.
-We quantified the accuracy of the inferences as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities, respectively.
-We compared inferences using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.
+To assess these sources of error and effect of data set sizes independently, we simulated sequences evolving along trees and explicitly recorded the number of transitions between states at every site ($n_{ij}^a$) and the time spent in each state ($\tau_j^a$) for a range of sample sizes, levels of divergence, and models with different degrees of variation between sites, see Materials and Methods.
+% For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data sets per parameter combination.
+From these simulated data sets, we inferred the site specific models using different aspects of the simulated data: (i) full knowledge of all transitions and ancestral states, (ii) the tree and the tip sequences, or (iii) only the tip sequences from which a tree was reconstructed.
+We quantified the accuracy of the inferences as the squared estimation error $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities, respectively.
+Lower $\chi^2$ corresponds to more accurate estimates.
 
@@
 
-Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a tree with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
-The graph contains estimates of $\eqp_i^a$ from the alignment and via the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
-The data are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the accuracy is inversely proportional to the expected number of substitutions.
+In our first analysis summarized in Fig.~\ref{fig:dressed} we quantify the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
+As expected, the average squared error $\chi^2$ of the $\eqp_i^a$ decreases with the substitution rate and the size of the data set.
+The error is well predicted by the average number of substitutions per site, which increases both with data set size and with the substitution rate.
+The data in Fig.~\ref{fig:dressed}A are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the squared deviation is inversely proportional to the expected number of substitutions.
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-The simulation data was generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
+In contrast, using the frequencies of different sequence states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).
+
+The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
 The scaling behavior $\chi_2 (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
 In case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
-A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).
 
@@
 The relative substitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.
+Regularization by pseudo-counts $\pc$ reduces the overfitting problem when the number of substitutions per site is small.
 
 %%% effect of reconstruction
-In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized).
+Above, we investigated the accuracy of model inferences when the ancestral states and the tree are known.
+In practice, however, ancestral sequences are unknown and need to be inferred or summed over (marginalized) which will introduce additional uncertainty.
+This problem was recently described as the ``Darwinian uncertainty principle'' by \citet{gascuel_darwinian_2020} showing that there are fundamental limits to the accuracy at which  model and ancestral states can be estimated jointly.
+
 To test the influence of tree and ancestral state reconstruction, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model (simulated nucleotide sequences) or FastTree \citep{FastTree} using the default 20 category (simulated amino-acid sequences).
 Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
-The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^a_{ij}$ and $\tau_i^a$ from this reconstruction.
+The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a simple evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^a_{ij}$ and $\tau_i^a$ from this reconstruction.
 This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
-Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
+
+Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest reduction of the error (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
 More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
-The accuracy now continues to improve up to root-to-tip distances of about one.
+The error now continues to decrease up to root-to-tip distances of about one.
 At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
@@
 Similar improvements are achieved by optimizing tree branch lengths along with the model.
+These improvements are consistent with \citet{gascuel_darwinian_2020} in that joint estimation of ancestral states and model is possible for Yule trees in which tree length grows linearly with the number of tips while tree depth increases only logarithmically.
+
 We found qualitatively similar patterns for 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.
@@
 	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length.
-Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation.
-Not that inferred site-specific models only partially ameliorate underestimation, see main text.
+Panel B shows the results of the same optimization for the average root-to-tip distance which is dominated by deep long branches that are more strongly affected.
+While using the true model results in unbiased branch lengths, inferred site-specific models only partially ameliorate underestimation, in particular with high pseudo-counts $\pc$ (see main text).
 Parameters: $n=1000$, $\alpha=1.5$.
@@
 Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
-While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR+R10 model without site-specificity.
-Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies which are prone to underestimation.
+While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR+R10 model without site-specific preferences.
+Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies that are prone to underestimation.
 When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
@@
 %% dependence on nucleotide and amino acid skew
-The underestimation of branch lengths is less severe for larger alphabets of if frequencies are not heavily skewed, see Fig.~\ref{fig:tree_length_supp}.
+The underestimation of branch lengths is less severe for larger alphabets or if frequencies are not heavily skewed, see Fig.~\ref{fig:tree_length_supp}.
 
@@
 For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
-This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
-We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site-specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
+This large sample of moderately diverged sequences should be suited to estimate site-specific preference using the iterative inference frame work and quantify how selection shaped the evolution of HIV.
 
-One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
 Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ should depend on the fitness difference $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
@@
 \end{equation}
-On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site-specific GTR model $Q^a_{ij}$.
+On longer time scales, the fixation rates $f^a_{ij}$ correspond to transitions rates of the site-specific GTR model $Q^a_{ij}$.
 The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations \citep{neher_genetic_2013}.
@@
 The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
-We compared this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
-However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observed a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
-see Fig.~\ref{fig:HIV_fitness})
-Interestingly, this correlation between the intra-host estimates and cross-sectional estimates decreased as soon as regularization increased above 0.05 and we there for used a weak regularization of $\pc=0.01$.
+
+We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site-specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
+We compared $\frac{\Gamma_{in}}{\Gamma_{out}}$ to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
+However, instead of a linear relationship between $s_i^a$ and within host estimates of fitness costs, we observed a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (see Fig.~\ref{fig:HIV_fitness})
 This relationship explains about half the variance of nucleotide effects and about one third of amino acid effects.
+Interestingly, this correlation between the intra-host estimates and cross-sectional estimates decreased as soon as regularization increased above 0.05 and we therefore used a weak regularization of $\pc=0.01$.
 
-The fact that the relationship deviates from the expectation
+The fact that the relationship of fitness proxies deviates from the expectation
 \begin{equation}
@@
 Within host fitness effects are masked and damped at the population level, possibly due to fluctuating selection by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
-Such fluctuations in genetic background (epistasis) and environment (immune system) can result in non-linear averages at the population level.
-The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
+The ratio of in/out rates $e^{-2Ns_i^a}$ is a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
 
@@
 Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize.
-A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
+A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences.
 Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site-specific preference is the basis of all common homology search tools.
@@
 \subsection*{Generation of simulated data}
+\label{sec:MM_simulated}
 To generate models and sequence ensembles for which the ground truth is known, we sampled binary trees with $n=100, 300, 1000, 3000$ leaves and Yule tree branch length statistics using \href{https://github.com/neherlab/betatree}{betatree} \citep{neher2013coalescence} for values of the average substitution rate given by $\langle \mu\rangle = [0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.25, 0.35, 0.5]$.
@@
 Site-specific preferences $\eqp_i^a$ (or equilibrium probabilities) were sampled from Dirichlet distributions with parameters $\alpha=1$ for alphabets with $q=4$ states (nucleotides) or $\alpha=0.2$ and $\alpha=0.5$ for alphabets with $q=20$ (amino acids).
+These distributions correspond to an average number of effective states of $N_{eff} = 2.6$ (nucleotides) and $4.7$ or $7.8$ (amino acids), where the number of effective states is given as $\left(\sum_i (\eqp_i^a)^2\right)^{-1}$.
 The entries of the transition matrix $W_{ij}$ were sampled from a Dirichlet distribution with $\alpha=2$.

### 2020-07-25 manuscript/response_to_referees.tex
edits to discussion

@@
 Or is the tree as a whole deemed older/younger than it really is, i.e.
-does the underestimation propogate through the tree all the way to the root or not?}
+does the underestimation propagate through the tree all the way to the root or not?}
+
+\response{
+    Such inference from large viral data sets is increasingly feasible as data set size increases and complements deep mutations scanning approaches.
+    While we identify problems for phylogenetic inference associated with strong variation model variation along the sequence, we don't have a solution.
+}
 
@@
 
-\Richard{This is one core part of the paper: this is a difficult problem to fix. }
+\response{In essence, Fig 3 shows that branch length underestimation caused by extensive variation of rates and preferences is very hard to avoid.
+As outlined above, we don't really have a good fix for that, but we investigate the causes in Fig 4.
+That said, the main result of the paper is that site-specific frequencies can be accurately estimated.
+The inability is estimating the length of long branches in RNA viruses is a serious concern for dating zoonosis and exploring the history of viruses causing measles, dengue, lassa etc.
+This problem is clearly apparent when trying to date the common ancestor of SIVs.
+Molecular clock approaches suggest an ancestor a few hundred years in the past, while viruses in island populations show that the common ancestor is much more ancient.
+}
 
@@
 
-\Richard{This shows that this reviewer really didn't understand what we are doing. We don't do partition models. }
+\response{
+The points raised by this reviewer have made it clear to us that the description of our basic motivation and model set-up was not clear enough.
+In response to this, we have rewritten several parts of the manuscript and point out explicitly in multiple places that we do not consider partition models but investigate to what extend models in which every site has their own set of of frequencies and an overall rate can be estimated.
+}
 
@@
 \response{The reviewer is correct, we don't have two data partitions of length 1000.
+Every site has its own set of frequencies and an overall rate.
 Instead, we produce hierarchical data sets that allow us to explore the effect of different aspects of the problem (tree, model realization, realization of the sequence evolution given the model) on the inferences.

### 2020-07-25 manuscript/ms.tex
edits to discussion

@@
 \section*{Discussion}
-Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize.
-A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences.
-Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site-specific preference is the basis of all common homology search tools.
-While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern_evolutionary_1998}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
-However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data set sizes are increasing rapidly such that more complex models can be estimated.
-Large data sets, however, require efficient and scalable methods.
+Different positions in a genome sequence or a protein change at different rates and explore different subsets of the available states.
+These site-specific properties are evolutionary conserved and are the basis of all common homology search tools.
+Which states are permissible at which position can be measured with high-throughput using deep mutational scanning techniques \citep{fowler_deep_2014} or can be estimated from large alignments of homologous sequences.
+Both approaches can reveal biological function and organization of the entity coded for by the sequence.
+\citet{bruno1996modeling} pointed out that the frequencies of states in columns of an alignment are distorted by phylogenetic correlations and how these phylogenetic correlations can be accounted for.
+However, estimating these frequencies accurately requires large data sets that have only recently become available.
+
+While accurate estimates of site-specific frequencies require correction of phylogenetic correlations, accurate phylogenetic inference requires sufficiently realistic models of sequence evolution while being computationally tractable and easy to parameterize.
+A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences \citep{Lartillot01062004}.
 

### 2020-07-25 manuscript/ms.tex
more rewriting

@@
 Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
-Modeling this complexity, however, requires an extensive number or parameters that can not be infered from small data sets and are costly to infer from large data sets.
-Most commonly used evolutionary models therefore reduce this complexity by partitioning sites into discrete classes with different rates and/or preferences where appropriate complexity is determined by model selection procedures.
+One one hand, such constraints on sequence evolution are used to infer biological function, one the other hand they need to be accounted for in phylogenetic reconstruction.
+Most commonly used evolutionary models try to account for this complexity by partitioning sites into a small number of discrete classes with different rates and/or preferences where appropriate complexity is determined by model selection procedures.
 Here, we present an efficient algorithm to estimate more complex models that allow for different preferences at every site and explore the accuracy at which such models can be estimated from simulated data.
-We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
-Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogenetic analyses -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
+Our iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
+We further explore the importance of site-specific preferences during estimation of branch length of phylogenetic trees and show that ignoring site-specific preferences results in substantial underestimation comparable to the error incurred when ignoring rate variation.
 However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
@@
 As a result, some parts of genomes change rapidly, while other are strongly conserved.
-In addition to variation of the evolutionary rate, different sites in a genome only explore different subsets of the available states.
+In addition to variation of the evolutionary rate, different sites in a genome explore different subsets of the available states.
 Some positions in a protein, for example, might only allow for hydrophobic amino acids, while others require acidic side chains.
@@
 Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result in biased estimates of divergence times or otherwise erroneous results \citep{yang1996among}.
-Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
+Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more frequent substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
 To avoid over-fitting, these methods typically don't estimate a rate for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple unimodal distributions \citep{mayrose2005gamma}, or a small number of fixed rates).
 
-In addition to rate variation, different sites in a protein differ in the amino acids they allow and constraints on the DNA or RNA sequence impose additional levels of selection.
+In addition to rate variation, different sites in a protein differ in the amino acids they allow, and constraints on the DNA or RNA sequence impose additional levels of selection.
 Such variation can again be accounted for by modeling multiple categories with different preferences for amino acids or nucleotides to which sites can be assigned or which can be integrated out during phylogenetic inference \citep{Lartillot01062004,kainer_effects_2015,shapiro_choosing_2006}.
-Here, we ask study under what circumstances models that allow different state frequencies at every site in the alignment can be estimated and what implications such variation has for phylogenetic inference.
 
@@
 
+Instead of partitioning the sequence into a small number of categories, we ask under what circumstances it is possible to estimate models that allow different state frequencies at every site in the alignment and what implications such variation has for phylogenetic inference.
 While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
@@
 	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
-	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
-Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, gray lines with $\alpha=3$.
-	The estimates are more accurate than naive estimates from the state frequencies in alignment columns.
-The advantage over naive estimates increases with data set size as seen from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
+	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\hat{\eqp}_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
+    This scaling holds for different number of sequences per tree ($n\in [100,300,1000,3000]$) and different degrees of rate variation.
 	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Accuracy of inferences}
-The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^a$ and $\tau_{j}^a$ can be estimated, and (iii) the accuracy of the tree reconstruction.
-Furthermore, the estimation of this extensive number of parameters requires large data sets in which most positions mutate multiple times on the tree -- otherwise the model will overfit the data and return inaccurate estimates.
+\section*{Accuracy of model inferences}
+The accuracy of this iterative solution will depend (i) on the validity of the linear approximation, (ii) the accuracy to which $n_{ij}^a$ and $\tau_{j}^a$ can be estimated, and (iii) the accuracy of the tree reconstruction.
+Furthermore, the estimation of this extensive number of parameters requires large data sets in which most positions mutate multiple times on the tree -- otherwise the model will overfit the data and result in inaccurate estimates.
 
@@
 	Panels A and B show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
+    Alignment frequencies are poor estimates of the $\eqp_i^a$ even in very diverse samples, while the different phylogeny aware methods initially improve rapidly with root-to-tip distance.
 	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
@@
 In our first analysis summarized in Fig.~\ref{fig:dressed} we quantify the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
-As expected, the average squared error $\chi^2$ of the $\eqp_i^a$ decreases with the substitution rate and the size of the data set.
+As expected, the average squared error $\chi^2$ decreases with the substitution rate and the size of the data set.
 The error is well predicted by the average number of substitutions per site, which increases both with data set size and with the substitution rate.
@@
 This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
-In contrast, using the frequencies of different sequence states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).
 
 The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
-The scaling behavior $\chi_2 (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
-In case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
+The scaling behavior $\chi_2 \sim (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
+In the latter case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
 
@@
 
-To test the influence of tree and ancestral state reconstruction, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model (simulated nucleotide sequences) or FastTree \citep{FastTree} using the default 20 category (simulated amino-acid sequences).
+To test the influence of tree and ancestral state reconstruction, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model (simulated nucleotide sequences) or FastTree \citep{FastTree} using the default settings (simulated amino-acid sequences).
 Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
@@
 These improvements are consistent with \citet{gascuel_darwinian_2020} in that joint estimation of ancestral states and model is possible for Yule trees in which tree length grows linearly with the number of tips while tree depth increases only logarithmically.
+In contrast, using the frequencies of different sequence states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).
 
@@
 For larger alphabets, the breakdown at large root-to-tip distances is less dramatic and sets in at larger values.
-However, these difference depend on the skew of the equilibrium probabilities.
+Overall, we conclude that site specific frequencies can be estimated accurately when the overall tree length comfortably exceeds 10 such that most sites experienced multiple mutations.
 

### 2020-07-25 src/summarize_preferences.py
update plots

@@ -1,30 +1,50 @@
+import glob
 import matplotlib.pyplot as plt
 from generate_toy_data import *
 
 def entropy(a):
     eps = 1e-10
-    return np.sum(np.log(a+eps)*(a+eps), axis=0)
+    return -np.sum(np.log(a+eps)*(a+eps), axis=0)
 
+def partip(a):
+    return 1/np.sum(a**2, axis=0)
 
 if __name__ == '__main__':
 
     import argparse
     parser = argparse.ArgumentParser(description = "", usage="analyze simulated data for site specific GTR reconstruction project")
     parser.add_argument('--mask', type=str, nargs='+')
+    parser.add_argument('--labels', type=str, nargs='+')
     args = parser.parse_args()
 
     fs = 16
-    plt.figure()
-    for mask in args.mask
+    plt.figure('S')
+    plt.figure('N')
+    for label, mask in zip(args.labels, args.mask):
         files = glob.glob(mask)
         allS = []
+        allN = []
         for file in files:
             m = load_model(file)
             S = entropy(m.Pi)
+            N = partip(m.Pi)
             allS.extend(S)
+            allN.extend(N)
 
-        plt.plot(sorted(allS), np.linspace(0,1,len(allS)))
-        print(f"{mask}, average entropy: {np.mean(allS)}")
+        plt.figure('S')
+        plt.plot(sorted(allS), np.linspace(0,1,len(allS)), label=label)
+        plt.figure('N')
+        plt.plot(sorted(allN), np.linspace(0,1,len(allN)), label=label)
+        print(f"{mask}, average entropy: {np.mean(allS):1.3f}, average N: {np.mean(allN):1.3f}")
+
+    plt.figure('S')
     plt.ylabel('fraction with less entropy', fontsize=fs)
     plt.xlabel('fraction of sites', fontsize=fs)
+    plt.legend()
     plt.savefig('figures/entropy_stats.pdf')
+
+    plt.figure('N')
+    plt.ylabel('fraction with fewer effective states', fontsize=fs)
+    plt.xlabel('fraction of sites', fontsize=fs)
+    plt.legend()
+    plt.savefig('figures/partip_stats.pdf')

### 2020-07-25 manuscript/ms.tex

@@
 
-\subsection*{Ignoring site-specific frequencies results in branch length underestimates}
+\subsection*{Implications for branch length estimates in phylogenies}
 As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
-This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
+This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{(\eqp_i^a)}^2$, which is increasing sharply with more peaked preferences.
 Models that don't account for site-specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
@@
 %% dependence on pseudo count
-Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
+Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch lengths optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
 While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR+R10 model without site-specific preferences.
-Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies that are prone to underestimation.
-When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
+This problem is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies that are prone to underestimation.
+When ignoring site-specific preferences, the inferred root-to-tip distance is essentially flat for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
 This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
@@
 \end{equation}
-where $\langle \mu^a\rangle$ is the expectation value of the rate.
+where $\langle \mu^a\rangle$ is the average of the rate.
 Mixtures of site-specific frequencies are constructed analogously.
@@
 
-\subsection*{Applications to large HIV alignments}
+\section*{Applications to large HIV alignments}
 Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
@@
 We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site-specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
-We compared $\frac{\Gamma_{in}}{\Gamma_{out}}$ to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
+We compared $2N s_i^a = \log\left(\frac{\Gamma_{in}}{\Gamma_{out}}\right)$ to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
 However, instead of a linear relationship between $s_i^a$ and within host estimates of fitness costs, we observed a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (see Fig.~\ref{fig:HIV_fitness})
@@
 Within host fitness effects are masked and damped at the population level, possibly due to fluctuating selection by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
-The ratio of in/out rates $e^{-2Ns_i^a}$ is a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
+The ratio of in/out rates $e^{2Ns_i^a}$ is a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).
 
@@
 Different positions in a genome sequence or a protein change at different rates and explore different subsets of the available states.
-These site-specific properties are evolutionary conserved and are the basis of all common homology search tools.
+These site-specific properties are evolutionary conserved and are the basis of common homology search tools.
 Which states are permissible at which position can be measured with high-throughput using deep mutational scanning techniques \citep{fowler_deep_2014} or can be estimated from large alignments of homologous sequences.
@@
 While accurate estimates of site-specific frequencies require correction of phylogenetic correlations, accurate phylogenetic inference requires sufficiently realistic models of sequence evolution while being computationally tractable and easy to parameterize.
-A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences \citep{Lartillot01062004}.
+A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence.
+More complex models allow for a small number of data partitions with different preferences \citep{Lartillot01062004,kainer_effects_2015}.
 
-Here, we presented method to estimate site-specific GTR models using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
+Here, we presented method to estimate site-specific preferences using an iterative EM-type approach inspired by \citet{bruno1996modeling}.
 We showed that site-specific models can be inferred with an accuracy limited by the Poisson statistics of the substitution process using an efficient iterative scheme.
@@
 This points to parameter identifiability problems rooted in the similar effects that skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
+This problems are distinct from, but related to, the inability to jointly infer ancestral states and rate parameters of evolving discrete states \citep{gascuel_darwinian_2020} as we can marginalize over the ancestral states here.
 

### 2020-07-26 manuscript/response_to_referees.tex
worked on response

@@
 
-\begin{itemize}
-    \item general issue: site specific models are computationally expensive, but only useful for large data sets.
-    \item main objective here is to provide a method for estimating them from large data sets, but point out the limitations.
-\end{itemize}
+We have revised our manuscript extensively in response to the reviewer comments.
+In particular, we have put more emphasis on the parts of the manuscript that address estimation of site-specific preferences as means to learn about the biology of the organism as a opposed tto nuisance parameters in phylogenetic inference.
+We now also stress that we are not studying site-heterogeneity in the form of a small number of data partitions, but are interested in the ability to estimate preferences for every site in the alignment.
+Overall, we feel that the manuscript has much improved from the this refocus and has become simpler to follow.
+
+\subsection*{Associate Editor: Martin, Darren}
+
+\editor{This seems like a good paper but in its current state virus evolution may not be a very good fit for it.
+I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.}
+
+\response{
+Thank you very much for handling the paper and providing these two reviews.
+We realized that our original manuscript was misleading to people familar complex mixture models that pratition the seqeunce data into categories with similar properties.
+We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specifci properties of every columns of an alignment.
+Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available.
+
+}
 
@@
 
-\Richard{ask Sarah}
 \response{we have attempted to compare our results with
-\texttt{swmutsel}. However, \texttt{swmutsel} seems completely focussed on codon models, something that isn't currenly implemented in treetime. In terms of run-time,
-swmutsel takes hours where the site specific inference by treetime takes minutes.
+\texttt{swmutsel}.
+However, \texttt{swmutsel} seems completely focussed on codon models, something that isn't currenly implemented in treetime.
+In terms of run-time, swmutsel takes hours where the site specific inference by treetime takes minutes.
 But we are afraid the results are not immediately comparable. }
@@
     \item we changed Fig 5 and Fig S4 to two dimensional histograms
-    \item we added second legend to figure 1 that maps line style to samples size
+    \item we added second legend to figure 1 that maps line style to samples size.
+We also removed the curves corresponding to naive estimates by alignment frequencies.
+These are also include in Fig 2 and hence not needed here.
     \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
@@
 
-We now point out in the abstract that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
+We now point out in multiple places that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
 }
@@
 
-\response{We discuss site-heterogenous partition models carefully in the main text.}
+\response{We now discuss site-heterogenous partition models carefully in the main text.}
 
@@
 Consistent with our results, they find that this is possible for Yule trees.
-Here, we are mmostly concerned with estimating the model, not the root state.
-We also consider sequence evolution rather than the evolution of a single character.
-This allows us to investigate the extent to which branch lengths can be estimated along with the model.
+Here, we are mostly concerned with estimating the model, not the root state, and treat ancestral states as nuissance parameters that at least in principle can be integrated out.
+In constrast to Gascuel and Steel, we consider sequence evolution rather than the evolution of a single character which allows our model to share the symmetric transmission matrix across sites.
+It also allows us to investigate the extent to which branch lengths can be estimated along with the model.
 
@@
 
-We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of the inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
+We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of model inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
 }
@@
 The connection to Gascuel and Steel is particularly apt when discussion the inability to jointly infer the ancestral state and the model.
-We also draw attention to the fact that accurate model estimation is possible for know trees, consistent with the results by G\&S that estimation is possible for Yule trees.
+We also draw attention to the fact that accurate model estimation is possible for known trees, consistent with the results by G\&S that joint ancestral reconstruction and model inference is possible for Yule trees.
 }
@@
 This section explores the ability to estimate site specific preferences from large and diverse alignments of RNA viruses.
-Similar analysis could be done for other RNA viruses.
-Furthermore, it points out implications for phylogenetic inferences.
+Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution.
+Furthermore, our work has implications for dating of divergence times of RNA viruses.
 We don't have solutions for the latter, but feel that it is still important to point them out.
-}
-
-\response{We have made an effort to simplify and clarify this example.
-But we feel that the application of to HIV data is one aspect of what makes this paper suitable for Virus Evolution: Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution.
 }
@@
 \response{
+We hope that the revision makes now clear that phylogenetic inference is not the focus of the manuscript, but that we instead study how properties of sites can be learned from sequence diversity.
 Such inference from large viral data sets is increasingly feasible as data set size increases and complements deep mutations scanning approaches.
 While we identify problems for phylogenetic inference associated with strong variation model variation along the sequence, we don't have a solution.
+The site specific models we estimate could serve as input for ParitionFinder or the "-MF+MERGE" feature of IQ-tree.
 }
-
-\Richard{See what model selectors do for simulated data}
 
@@
 This manuscript deals with GTR models and the root doesn't matter.
-We only use root-to-tip distances in mid-point-rooted trees as a measure of divergence.
+We use root-to-tip distances in mid-point-rooted trees as a measure of divergence, but the root is no special place otherwise.
 The only place where time comes in are the HIV examples, but we only use time as a proxy for evolutionary rate and don't do any molecular clock analysis.
@@
 
-\subsection*{Associate Editor: Martin, Darren}
-
-This seems like a good paper but in its current state virus evolution may not be a very good fit for it.
-I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.
-
 \end{document}

### 2020-07-26 manuscript/response_to_referees.tex
more edits

@@
 
-We have revised our manuscript extensively in response to the reviewer comments.
-In particular, we have put more emphasis on the parts of the manuscript that address estimation of site-specific preferences as means to learn about the biology of the organism as a opposed tto nuisance parameters in phylogenetic inference.
-We now also stress that we are not studying site-heterogeneity in the form of a small number of data partitions, but are interested in the ability to estimate preferences for every site in the alignment.
-Overall, we feel that the manuscript has much improved from the this refocus and has become simpler to follow.
+\subsection*{Associate Editor: Martin, Darren}
 
-\subsection*{Associate Editor: Martin, Darren}
+Thank you very much for handling the paper and providing these two constructive reviews that have helped us to improve the paper.
+Please accept our apologies for the delayed resubmission.
+My two co-authors have moved one (more than a year ago) and I have been caught up in SARS-CoV-2 related work for much of the year.
+
+We realized that our original manuscript was misleading to people familiar complex mixture models that partition the sequence data into categories with similar properties.
+We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specific properties of every column of an alignment.
+Many of the comments by Reviewer 2 suggest that they were expecting an analysis of partition models while in fact, we studied the identifiability of parameters describing selection at every single site in the sequence.
+Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available.
+We, therefore, think that the paper is of interest to the readership of Virus Evolution.
 
@@
 \response{
-Thank you very much for handling the paper and providing these two reviews.
-We realized that our original manuscript was misleading to people familar complex mixture models that pratition the seqeunce data into categories with similar properties.
-We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specifci properties of every columns of an alignment.
-Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available.
-
+We have rewritten large parts of the manuscript to make our work more accessible and relevant to the readership of Virus Evolution.
+Since our primary objective is to learn about the biology of virus from sequence diversity, rather than develop methods for phylogenetic inference, we believe Virus Evolution and not Systematic Biology is an appropriate venue for this work.
+The referee reports were very useful in rewriting and refocussing our work.
 }
@@
 
-\response{we have attempted to compare our results with
-\texttt{swmutsel}.
+\response{we have attempted to compare our results with \texttt{swmutsel}.
 However, \texttt{swmutsel} seems completely focussed on codon models, something that isn't currenly implemented in treetime.
@@
 \subsection*{Reviewer: 2}
+
+We thank the reviewer for their detailed review of our manuscript and the many constructive comments.
+This review as revealed that we should have been much more explicit about the relation of our manuscript to partition models commonly used to model site-heterogeneity of rates or preferences.
+Instead of partitioning the sequence data into a small number of categories, we investigate to what extent we can learn parameters that describe specific selective pressures at every column in the alignment.
+The primary objective is to learn about the function of the sequence, not phylogenetic inference.
 

### 2020-07-27 manuscript/response_to_referees.tex
more edits, polish

@@
 \newcommand{\comment}[1]{{\color{red}#1}}
+\definecolor{response}{rgb}{0.1, 0.1, 0.1}
 \definecolor{drab}{rgb}{0.59, 0.44, 0.09}
@@
 %\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
-\newcommand{\response}[1]{{\color{black}\textbf{Response:} #1}}
+\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}
 
@@
 
-Thank you very much for handling the paper and providing these two constructive reviews that have helped us to improve the paper.
+\response{
+Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper.
 Please accept our apologies for the delayed resubmission.
@@
 
-We realized that our original manuscript was misleading to people familiar complex mixture models that partition the sequence data into categories with similar properties.
+We realized that our original manuscript was misleading to people familiar with complex mixture models that partition the sequence data into categories with similar properties.
 We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specific properties of every column of an alignment.
@@
 We, therefore, think that the paper is of interest to the readership of Virus Evolution.
+}
 
@@
 The referee reports were very useful in rewriting and refocussing our work.
+
+We provide a point-by-point response below.
 }
@@
 
-\response{we have attempted to compare our results with \texttt{swmutsel}.
-However, \texttt{swmutsel} seems completely focussed on codon models, something that isn't currenly implemented in treetime.
+\response{The most obvious software to compare our results to would be \texttt{swmutsel}.
+However, \texttt{swmutsel} specifically models sequences are amino acides and their codons, something that isn't currenly implemented in treetime.
 In terms of run-time, swmutsel takes hours where the site specific inference by treetime takes minutes.
-But we are afraid the results are not immediately comparable. }
+But we are afraid the results are not immediately comparable.
+We have hence refrained from attempting a comparision.
+}
 
@@
 Calculating the effective number of states as $\left(\sum_i p_i^2\right)^{-1}$ yields 4.7 or 7.8 effective states.
-We added this to the manuscript.}
+We added th the latter to the manuscript.}
 
@@
 \begin{itemize}
-    \item we changed Fig 5 and Fig S4 to two dimensional histograms
+    \item we reduced marker size and opacity in Fig 5 and Fig S4 to reduce overplotting.
+We also experimented with two dimensional histograms, but bins at the border of the domain (almost invariable sites) dominate to a degree that it is hard to see differences in the remainder of the histogram.
     \item we added second legend to figure 1 that maps line style to samples size.
@@
 \subsection*{Reviewer: 2}
-
+\response{
 We thank the reviewer for their detailed review of our manuscript and the many constructive comments.
@@
 The primary objective is to learn about the function of the sequence, not phylogenetic inference.
+}
 
@@
 We now emphasize that we study whether models with different state preferences at every site can be accurately estimated.
-
 We now point out in multiple places that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
@@
 \response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do.
-Gascuel and Steel investigate whether the state of the root if a character can be inferred jointly with the equilibrium probabilities or the transition matrix.
+Gascuel and Steel investigate whether the state of the root of a character can be inferred jointly with the equilibrium probabilities or the transition matrix.
 Consistent with our results, they find that this is possible for Yule trees.
@@
 We now discuss our results in the light of the results by Gascuel and Steel and thank the reviewer for pointing out this work to us.
+G\&S's work is mentioned in the section discussing the accuracy of inferences and the discussion.
 }
@@
 The properties of category models have for example been studied systematically by Kainer and Lanfear.
-
 We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of model inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
@@
 \response{Indeed, Gascuel and Steel show a similar graph for the joint estimation of ancestral state of a single discrete character and the model of its evolution. We now point out this analogy.
-
 The connection to Gascuel and Steel is particularly apt when discussion the inability to jointly infer the ancestral state and the model.
@@
 We now point out that we first investigate the performance of the inference scheme with known tree and ancestral states and then investigate the feasibility of estimating tree, ancestral states, and model together.
-
 Reviewer 1 also had suggestions to improve the figures and we have added more information to the legends and made it easier to differentiate between different line.
@@
 We are afraid there is a misunderstanding here.
-This manuscript deals with GTR models and the root doesn't matter.
+Our manuscript infers GTR models and the root does not matter.
 We use root-to-tip distances in mid-point-rooted trees as a measure of divergence, but the root is no special place otherwise.
-The only place where time comes in are the HIV examples, but we only use time as a proxy for evolutionary rate and don't do any molecular clock analysis.
+The only place where time comes in are the HIV examples, but we only use time as a proxy for divergence via the known evolutionary rate and don't do any molecular clock analysis.
 All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}

### 2020-07-27 manuscript/ms.tex
more edits
more edits, polish

@@
 Abstract: Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
-Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
-One one hand, such constraints on sequence evolution are used to infer biological function, one the other hand they need to be accounted for in phylogenetic reconstruction.
-Most commonly used evolutionary models try to account for this complexity by partitioning sites into a small number of discrete classes with different rates and/or preferences where appropriate complexity is determined by model selection procedures.
+Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
+One one hand, such constraints on sequence evolution can be to infer biological function, one the other hand they need to be accounted for in phylogenetic reconstruction.
+Phylogenetic models often account for this complexity by partitioning sites into a small number of discrete classes with different rates and/or state preferences.
+Appropriate model complexity is typically determined by model selection procedures.
 Here, we present an efficient algorithm to estimate more complex models that allow for different preferences at every site and explore the accuracy at which such models can be estimated from simulated data.
 Our iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
-We further explore the importance of site-specific preferences during estimation of branch length of phylogenetic trees and show that ignoring site-specific preferences results in substantial underestimation comparable to the error incurred when ignoring rate variation.
-However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
+However, the joint estimation of site-specific rates, and site-specific preferences, and phylogenetic branch length can suffer from identifiability problems, while ingoring variation in preferences across sites results in branch length underestimates.
 Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
-Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
+Analysis of these substitution models suggests near saturation of divergence after a few hundred years.
 Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
@@
 
-In addition to rate variation, different sites in a protein differ in the amino acids they allow, and constraints on the DNA or RNA sequence impose additional levels of selection.
+In addition to rate variation different sites in a protein differ in the amino acids they allow, different positions in codons experience different constraints, and secondary structure or protein binding to DNA or RNA imposes additional levels of selection.
 Such variation can again be accounted for by modeling multiple categories with different preferences for amino acids or nucleotides to which sites can be assigned or which can be integrated out during phylogenetic inference \citep{Lartillot01062004,kainer_effects_2015,shapiro_choosing_2006}.
@@
 	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
-	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\hat{\eqp}_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
+	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\hat{\eqp}_i^a$ scales inversely with the tree length (indicated by the black line), suggesting the accuracy is limited by the Poisson statistics of observable mutations.
     This scaling holds for different number of sequences per tree ($n\in [100,300,1000,3000]$) and different degrees of rate variation.
@@
 
-In our first analysis summarized in Fig.~\ref{fig:dressed} we quantify the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
+In our first analysis summarized in Fig.~\ref{fig:dressed} we quantified the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
 As expected, the average squared error $\chi^2$ decreases with the substitution rate and the size of the data set.
@@
 
-The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
+The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (blue lines) and $\alpha=3$ (orange lines).
 The scaling behavior $\chi_2 \sim (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
-In the latter case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
+In the latter case of strong rate variation, the average squared error $\chi^2$ is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
 
@@
 
-Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest reduction of the error (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
+Instead of reconstructing the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest reduction of the error (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
 More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
@@
 \subsection*{Implications for branch length estimates in phylogenies}
-As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
+As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if these skews are not modeled appropriately.
 This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{(\eqp_i^a)}^2$, which is increasing sharply with more peaked preferences.
@@
 	Panels A \& B show the relative error in total tree length when using a mixture model as defined in Eq.~\ref{eq:mixture_model} for branch length inference.
-Panel A shows this error as a function of the mixing fraction $\alpha$ for $\langle \mu \rangle = 0.2$.
-Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\alpha=1$.
+Panel A shows this error as a function of the mixing fraction $\gamma$ for $\langle \mu \rangle = 0.2$.
+Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\gamma=1$.
 	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
@@
 
-\subsection*{Minor model deviations can have substantial effects on divergence estimates}
+\subsection*{The effect of model misspecification on divergence estimates}
 In the previous section, we have observed that model deviations, for example the error made during model inference, can result in substantial errors in branch length estimates.
 To investigate this effect more systematically, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows:
-For a mixing fraction $\alpha$, we constructed a model with rates
+For a mixing fraction $\gamma$, we constructed a model with rates
 \begin{equation}
 \label{eq:mixture_model}
-	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
+	\mu^a = \gamma \langle \mu^a\rangle + (1-\gamma) \mu^a_{\mathrm{true}}
 \end{equation}
@@
 Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
-Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
-If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
+Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\gamma = 0.3$.
+If the same preferences are used for every site ($\gamma=1$), however, deviations are substantial.
 Deviations in rate and preferences are approximately additive.
@@
 \end{equation}
-The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
+The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the common multiplicative factor $2N$ is unknown.
 
@@
 This points to parameter identifiability problems rooted in the similar effects that skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
-This problems are distinct from, but related to, the inability to jointly infer ancestral states and rate parameters of evolving discrete states \citep{gascuel_darwinian_2020} as we can marginalize over the ancestral states here.
+This problems are related to the inability to jointly infer ancestral states and rate parameters of evolving discrete states \citep{gascuel_darwinian_2020}.
+Model and tree inference alone, however, does not require acurate reconstruction of ancestral states as these can be marginalized.
 

### 2020-07-27 src/analyze_HIV_tree.py
update plot scripts

@@ -105,28 +105,31 @@ if __name__ == '__main__':
 
     max_bin = int(50*len(fitness_estimates_gtr)/3000)
 
+
+    opa = 0.3
+    msize = 8
     ccoef=pearsonr
     #ccoef=spearmanr
     div_rec = diversity(gtr.Pi)
     cc_rec = ccoef(np.log(fabio_fitness['median'])[valid], np.log(1e-10+div_rec)[valid])
     plt.figure()
-    g = plt.hexbin(fabio_fitness['median'], div_rec, xscale='log', yscale='log')
+    # g = plt.hexbin(fabio_fitness['median'], div_rec, xscale='log', yscale='log')
 #    g.ax_joint.set_yscale('log')
 #    g.ax_joint.set_xscale('log')
 #    g.ax_joint.set_xlim([3e-5, 3e0])
 #    g.ax_joint.set_ylim([3e-5, 1.5])
 
-    # plt.scatter(fabio_fitness['median'], div_rec, label='reconstructed, r2=%1.2f'%cc_rec[0]**2)
-    # plt.legend()
-    # plt.xlim([3e-5, 3e0])
-    # plt.ylim([3e-5, 1.5])
-    # plt.yscale('log')
-    # plt.xscale('log')
+    plt.scatter(fabio_fitness['median'], div_rec, label='reconstructed, r2=%1.2f'%cc_rec[0]**2, alpha=opa, s=msize)
+    plt.legend()
+    plt.xlim([3e-5, 3e0])
+    plt.ylim([3e-5, 1.5])
+    plt.yscale('log')
+    plt.xscale('log')
 
     div_aln = diversity(aln_freq)
     cc_aln = ccoef(np.log(fabio_fitness['median'][valid]), np.log(1e-10+div_aln)[valid])
     plt.figure()
-    plt.scatter(fabio_fitness['median'], div_aln, label='alignment, r2=%1.2f'%cc_aln[0]**2)
+    plt.scatter(fabio_fitness['median'], div_aln, label='alignment, r2=%1.2f'%cc_aln[0]**2, alpha=opa, s=msize)
     plt.legend()
     plt.xlim([3e-5, 3e-0])
     plt.ylim([3e-5, 1.5])
@@ -136,17 +139,17 @@ if __name__ == '__main__':
 
     plt.figure()
[…]

### 2020-07-27 src/model_deviation.py

@@ -67,7 +67,7 @@ if __name__ == '__main__':
         std = np.std(eps_graph[k],axis=1)
         axs[0].errorbar(eps_vals, m, std, label=label_str, lw=3)
 
-    axs[0].set_xlabel(r'mixing ratio $\alpha$ true/flat, $\langle \mu \rangle=$' +str(args.plot_mu), fontsize=fs)
+    axs[0].set_xlabel(r'mixing ratio $\gamma$ true/flat, $\langle \mu \rangle=$' +str(args.plot_mu), fontsize=fs)
     axs[0].set_ylabel('rel. tree length error', fontsize=fs)
     axs[0].tick_params(labelsize=fs*0.8)
     axs[0].set_ylim([0.4,1.1])
@@ -77,7 +77,7 @@ if __name__ == '__main__':
         m = np.mean(mu_graph[k],axis=1)
         std = np.std(mu_graph[k],axis=1)
         axs[1].errorbar(mu_vals, m, std, label=label_str, lw=3)
-    axs[1].set_xlabel(r'evolutionary rate $\langle \mu \rangle$ ($\alpha=1$)', fontsize=fs)
+    axs[1].set_xlabel(r'evolutionary rate $\langle \mu \rangle$ ($\gamma=1$)', fontsize=fs)
     axs[1].legend(fontsize=fs)
     axs[1].tick_params(labelsize=fs*0.8)
     add_panel_label(axs[1],'B',  x_offset=+0.02, y_offset=0.9, fs=fs)

### 2020-08-10 manuscript/ms.tex
resubmission/acceptance

@@
 Our iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
-However, the joint estimation of site-specific rates, and site-specific preferences, and phylogenetic branch length can suffer from identifiability problems, while ingoring variation in preferences across sites results in branch length underestimates.
+However, the joint estimation of site-specific rates, and site-specific preferences, and phylogenetic branch length can suffer from identifiability problems, while ignoring variation in preferences across sites results in branch length underestimates.
 Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
